FKBP35 secures ribosome homeostasis in Plasmodium falciparum

Version 1

Abstract

Abstract

Plasmodium falciparum accounts for the majority of over 600’000 malaria-associated deaths annually. Parasites resistant to nearly all antimalarials have emerged and the need for drugs with alternative modes of action is thus undoubted. The FK506-binding protein PfFKBP35 has gained attention as a promising drug target due to its high affinity to the macrolide compound FK506 (tacrolimus). Whilst there is considerable interest in targeting PfFKBP35 with small molecules, a genetic validation of this factor as a drug target is missing and its function in parasite biology remains elusive. Here, we show that limiting PfFKBP35 levels are lethal to P. falciparum and result in a delayed-death phenotype that is characterized by defective ribosome homeostasis and stalled protein translation. We furthermore show that FK506, unlike the role of this drug in model organisms, exerts its anti-proliferative activity in a PfFKBP35-independent manner and, using cellular thermal shift assays, we identify FK506-targets beyond PfFKBP35. In addition to revealing first insights into the function of PfFKBP35, our results show that FKBP-binding drugs can adopt non-canonical modes of action – with major implications for the development of FK506-derived molecules active against Plasmodium parasites and other eukaryotic pathogens.

Introduction

Despite considerable progress in recent years, malaria remains one of the major global health threats (1). The apicomplexan parasite Plasmodium falciparum is responsible for the vast majority of the over 600’000 malaria deaths in 2020. Most of this burden is carried by infants and children under the age of five in sub-Saharan Africa. Malaria deaths increased by 12% compared to 2019, which can at least in part be explained by the COVID-19 related suspension of malaria control and treatment measures (1). In addition, the emergence of parasites resistant to most antimalarials, including artemisinin-based combination therapies, endangers current and future disease elimination efforts and underscores the need for developing drugs with alternative modes of action (2).

FK506 (tacrolimus), a well-characterized immunosuppressant, shows considerable activity against asexual blood stage parasites (3). P. falciparum encodes a single FK506-binding protein (FKBP) dubbed PfFKBP35 (PF3D7_1247400), which is considered to be a promising drug target (4, 5). FKBPs belong to the immunophilin family and are conserved throughout the eukaryotic kingdom (6, 7). In contrast to the malaria parasite, most other eukaryotes encode several FKBPs that, besides showing high affinity for the FK506 drug, exert peptidyl-prolyl isomerase (PPIase) activity. The PPIase moiety catalyzes cis-trans isomerization of proline residues – an event thought to be rate-limiting for the folding of many proteins (8). Independent of their PPIase activity, FKBPs act as chaperones to prevent protein aggregation under stress conditions (9, 10). Furthermore, some FKBP variants play a role in cellular signaling or gene regulation (11, 12). Human FKBP12 for instance is best-known for its ability to form a complex with the immunosuppressant drugs FK506 and rapamycin. These complexes inhibit the phosphatase activity of calcineurin and the kinase activity of mTOR (mechanistic target of rapamycin), respectively (13). Amongst many other effects, this inhibition results in decreased T-cell activation. Of note, the regulatory role of HsFKBP12 depends on the presence of rapamycin or FK506. In absence of drugs, this immunophilin neither interacts with mTOR nor with calcineurin and does hence not affect activity of these key components of eukaryotic cells. Due to the conserved nature of FKBPs, rapamycin and FK506 have also gained attention as potential antimicrobial drugs (3, 7).

Similar to its homologs in model organisms, PfFKBP35 harbors PPIase and chaperoning activity in vitro (5). Despite sharing key features with FKBPs of other eukaryotes, its role in P. falciparum remains elusive. PfFKBP35 is believed to interact with heat shock proteins (HSPs) and co-immunoprecipitates with cytoskeletal factors of P. falciparum (14). While PfFKBP35 is inhibiting the phosphatase activity of recombinant calcineurin in a FK506-independent manner (15–17), it does not co-localize with calcineurin in vivo (16). The protein is expressed throughout the 48-hour intra-erythrocytic development of P. falciparum and is found within the cytosol of ring stage parasites. In the older trophozoite and the replicating schizont stages, the protein was reported to also localize to the nucleus (16). While a transposon-based mutagenesis screen suggests that PfFKBP35 is essential for intra-erythrocytic replication (18), neither its cellular function nor its essentiality was confirmed experimentally in live parasites. The unknown role in parasite biology notwithstanding, PfFKBP35 is considered to be a viable drug target and, given the fact that P. vivax and P. knowlesi encode FKBP homologs that exhibit comparable PPIase activities (19, 20), future FKBP-targeted therapies may be effective across different Plasmodium species.

Crystal structures of PfFKBP35 in complex with FK506 and rapamycin revealed high affinity interactions with the FK506-binding domain of the protein (15, 21, 22). Consistent with these observations, in vitro enzyme activity assays showed that FK506 inhibits the PPIase activity of recombinant PfFKBP35 in a dose-dependent manner (5). Previous research aimed at exploiting structural differences between human and parasite-encoded FKBPs to circumvent the immunosuppressive activity of drug-bound HsFKBP12 (23–26). These efforts include investigation of non-covalent FKBP inhibitors such as adamantyl derivatives, macrocycles, the small molecule D44 (23-25, 27, 28) and the synthetic ligand for FKBP (SLF) as well as derivatives thereof designed to covalently bind PfFKBP35 (29). While D44 shows promising antimalarial activity, it fails to alter thermostability of the FK506-binding domains of PfFKBP35 and HsFKBP12 in vitro (29), suggesting that this compound does not directly interact with PfFKBP35. Up until now, PfFKBP35 has not been validated as a drug target using reverse genetics, and the link between PfFKBP35-interacting drugs and their antimalarial activity remains elusive (3).

Here, we generated inducible PfFKBP35 knock-out, knock-down as well as over-expression cell lines and show that PfFKBP35 is essential for asexual replication of blood forms. We demonstrate that PfFKBP35 is vital for ribosome homeostasis on the post-transcriptional level and likely acts as a chaperone during the biogenesis of ribosomes in the nucleus. As a result, PfFKBP35 knock-out parasites stall protein translation prior to dying at the early schizont stage, which underpins the potential of PfFKBP35 as a drug target. Using cellular thermal shift assays (CETSA), we corroborate the high affinity binding of FK506 to PfFKBP35 and reveal interactions of the PfFKBP35/FK506 pair with ribosome-associated proteins. Importantly, however, our data demonstrate that FK506 fails at inhibiting the essential function of PfFKBP35 and exerts its antimalarial activity in a PfFKBP35-independent manner. While this has major implications for the development of future FK506-based antimalarials and/or PfFKBP35-inhibiting drugs, we identify putative targets of FK506 beyond PfFKBP35 in P. falciparum.

Results

PfFKBP35 is essential for asexual replication

In a first step, using CRISPR/Cas9-mediated gene editing, we engineered a P. falciparum cell line allowing for a conditional knock-down of PfFKBP35 using the DD/Shield system (30, 31) (Figs. 1A and S1A-B). Despite the efficient depletion of PfFKBP35 under knock-down conditions, parasites did not show apparent growth defects (Figs. S1C-D). Immunofluorescence assays (IFA) showed that PfFKBP35 localizes to distinct foci within the parasite nucleus (Fig. 1B).

PfFKBP35 depletion causes delayed parasite death.

(A) Schematic of the N-terminally modified fkbp35 locus in NF54/iKD-FKBP35 parasites. dd, destabilization domain. gfp, green fluorescent protein. (B) Localization of PfFKBP35 in NF54/iKD-FKBP35 parasites at the ring, trophozoite and schizont stage assessed by immunofluorescence. DNA was stained with DAPI. Representative images are shown. Scale bar: 5 µm. DIC, differential interference contrast.(C) Schematic of the modified fkbp35 locus in NF54/iKO-FKBP35 parasites before and after loxP recombination. gfp, green fluorescent protein. (D) Flow cytometry plot showing GFP expression in G1 schizonts of FKBP35KO and FKBP35WT. DNA was stained with DRAQ5. (E) Time course showing the development of NF54/iKO-FKBP35 parasites. Upper panel: Schematic of the experimental setup. Middle panel: DNA content of the populations assessed by flow cytometry based on the SYBR Green intensity. Lower panel: Giemsa-stained blood smears. The knockout was induced at 0-6 hpi. Representative data of three biological replicates are shown. (F) Multiplication rates of FKBP35KO and FKBP35WT parasites from generation 1 to generation 2 (G1-G2) and G2-G3. The knockout was induced at 0-6 hpi. Multiplication rates were determined using flow cytometry. Data points represent the multiplication rates from G1-G2 and G2-G3. n=3, error bars represent the standard error of the mean. (G) Schematic showing time point of growth arrest (red arrow) following PfFKBP35 knock-out at different hpi (blue arrow). R, ring stage; T, trophozoite stage; S, schizont stage.

Next, we generated a cell line allowing for the conditional knock-out of the endogenous coding sequence using a dimerizable Cre recombinase (DiCre)-based system (32). Specifically, we introduced two loxP sites – one within a synthetic intron (33) downstream of the start codon and one after the stop codon of the fkbp35 coding region – and added a green fluorescent protein (gfp) sequence downstream of this expression cassette (Figs. 1C, S2A). As a result, these NF54/iKO-FKBP35 parasites express tag-free, wild-type (WT) PfFKBP35 protein under control conditions (FKBP35WT). Upon DiCre activation, the loxP sites are recombined and the fkbp35 locus is excised from the parasite genome (Figs. 1C, S2B). Under these knock-out conditions (FKBP35KO), parasites express cytosolic GFP under control of the endogenous fkbp35 promoter. This setup allows monitoring knock-out efficiency at the single cell level and revealed that DiCre successfully deleted fkbp35 in the vast majority of parasites (Fig. 1D).

In a first attempt, we induced the fkbp35 knock-out in synchronous ring stage NF54/iKO-FKBP35 parasites at 0-6 hours post erythrocyte invasion (hpi), i.e. prior to the expected peak transcription of fkbp35 (34). These FKBP35KO parasites showed a delayed death phenotype: parasites completed intra-erythrocytic development in the first generation (G1) without noticeable effects (Fig. 1E) and entered the next generation (G2) with a multiplication rate similar to that of the FKBP35WT control population (Fig. 1F). In G2, however, FKBP35KO parasites arrested at the late trophozoite/early schizont stage (approximately 30-36 hpi) with many cells failing to enter DNA replication in the S-phase (Fig. 1E). To investigate this phenomenon in more detail, we induced the fkbp35 knock-out at three different time points during the IDC. Similar to DiCre activation at 0-6 hpi, knocking out fkbp35 at 12-18 hpi resulted in parasite death at the trophozoite/schizont stage of the subsequent generation. However, when induced at 24-30 hpi or 34-40 hpi in G1, parasites did not only complete the current IDC but also successfully passed through G2 before they eventually arrested in G3 (Fig. 1G). Considering that low PfFKBP35 levels are sufficient to maintain the essential function of the protein, as observed with the NF54/iKD-FKBP35 parasites (see Fig. S1C-D), it is conceivable that proteins expressed between 0 and 24hpi, i.e. prior to deleting fkbp35 from the genome, allow parasites to enter G3.

The delayed death of FKBP35KO is reminiscent of a phenotype observed in apicoplast-deficient parasites (35, 36). We therefore tested if PfFKBP35 is involved in apicoplast biology, but did not find evidence for an association with this organelle (Fig. S2C). Furthermore, given the reported interaction of PfFKBP35 with heat shock proteins (14, 16), we also tested the parasitès response to heat stress, but could not detect altered heat-susceptibility between FKBP35KO and FKBP35WT cells (Fig. S2D).

PfFKBP35 is crucial for ribosome homeostasis and protein translation

To further elucidate the function of PfFKBP35, we induced the fkbp35 knock-out in early ring stages (0-6 hpi) and compared the proteomes of FKBP35KO and FKBP35WT parasites at the schizont stage of G1 (36-42 hpi) and at the trophozoite stage of the following generation (G2, 24-30 hpi) using a quantitative proteomics approach.

The knock-out of PfFKBP35 causes accumulation of pre-ribosome components in G1

Consistent with the efficient deletion of fkbp35 on the genomic level, FKBP35KO schizonts of G1 showed a 19-fold (18.9 +/−5.7 s.d.) reduction of FKBP35 and GFP levels were increased by 84-fold (83.7 +/−38.5 s.d.) when compared to the control population (Fig. 2A, Table S1). Besides PfFKBP35 and GFP, 50 parasite proteins were significantly deregulated in FKBP35KO schizonts and 34 of them found at elevated levels compared to the FKBP35WT control conditions. A gene ontology analysis using PANTHER (37) indicated enrichment of the “pre-ribosome” (fold enrichment = 30.4, p-value = 1.46E-5, 4/34 proteins) and “nucleolus” (fold enrichment = 6.77, p-value = 8.85E-4, 5/34 proteins) terms among these factors. While the pre-ribosome is a large subunit precursor of mature ribosomes, the nucleolus represents the primary site for the biogenesis of ribosomal subunits (38). In contrast to factors being more abundant in knock-out parasites, none of the 16 factors with lower abundance in FKBP35KO cells were associated with ribosome-related processes (Table S1). Despite being phenotypically silent in G1, the accumulation of pre-ribosome and nucleolar components hints at perturbed ribosome maturation in these cells. Specifically, it is conceivable that the observed buildup of ribosome intermediates results from reduced anabolic activity, i.e. the stalling of nucleolar pre-ribosome assembly.

PfFKBP35 is crucial for ribosome homeostasis and protein translation.

(A) Left panel: Proteome analysis of schizonts (36-42 hpi) in G1. The volcano plot shows change in protein abundance in FKBP35KO compared to FKBP35WT. Ribosomal proteins are highlighted in blue, n=3. Right panel: Relative difference in PfFKBP35 and GFP abundance expressed as signal intensity measured by mass spectrometry. n=3, error bars represent the standard error of the mean. (B) Dose-response effect of the translation inhibitor cycloheximide on FKBP35KO and FKBP35WT parasites. The knockout was induced at 0-6 hpi and parasites were subsequently exposed to cycloheximide for 48 h. Parasite survival was quantified by flow cytometry. n=3, error bars represent the standard error of the mean. (C) Proteome analysis of trophozoites (24-30 hpi) in G2. The volcano plot shows change in protein abundance in FKBP35KO compared to FKBP35WT. Ribosomal proteins are highlighted in blue. n=3. (D) GO enrichment analysis of significantly deregulated proteins in FKBP35KO trophozoites in G2. GO enrichment analysis was performed using the PANTHER “Overrepresentation Test” with the test type “Fisher’s Exact” and the correction method “False Discovery Rate”. (E) Assessment of translation activity using SUnSET. FKBP35KO and FKBP35WT trophozoites of G2 were incubated with puromycin for 1 hour. Incorporated puromycin was detected using an α-puromycin antibody. α-GAPDH served as loading control. n=3, a representative Western blot is shown.

Notwithstanding these effects, FKBP35KO cells were equally sensitive to the translation inhibitor cycloheximide compared to their FKBP35WT counterparts (Fig. 2B), indicating that ribosome function is not compromised at this point in time (39). This is consistent with normal cell cycle progression in G1 and unaltered multiplication rates of FKBP35KO parasites, as described above (see Figs. 1E-F), and shows that the prominent depletion of PfFKBP35 does not have a major impact on parasite viability in G1.

PfFKBP35 is essential for maintaining normal levels of functional ribosomes in G2

In the following generation (G2, 24-30 hpi), the proteome of FKBP35KO trophozoites showed substantial differences. Despite the absence of morphological changes at this point in time, 216 proteins were significantly deregulated (Fig. 2C). Among these, 50 proteins were found at higher abundance compared to FKBP35WT and their gene ontology terms point towards an effect on processes involved in energy metabolism (Fig. 2D).

The majority of deregulated factors (166 of 216) in G2, however, were less abundant in FKBP35KO compared to the FKBP35WT control cells. Very prominently, 20 of the top 50 hit molecules are ribosomal proteins that, on average, show a two-fold reduction (2.0 +/− 0.4 s.d.) in FKBP35KO parasites (Table S1). It is therefore not surprising that PANTHER identified the knock-out of PfFKBP35 to mainly affect ribosome-related processes, such as “cytoplasmic translation”, “ribosome assembly”, and “ribosomal large subunit biogenesis” (Fig. 2D) (37).

To assess if the reduced levels of ribosomal proteins result in lower protein translation rates, we employed the “surface sensing of translation” (SUnSET) technique. SUnSET is based on incorporation of the tyrosyl-tRNA analogue puromycin into nascent peptide chains and thereby allows monitoring newly synthesized proteins (40). As expected, we found that FKBP35KO trophozoites in G2 (24-30 hpi) showed reduced translation rates compared to the FKBP35WT control population (Fig. 2E). Further, protein synthesis was not affected in parasites of the parental line NF54/DiCre following DiCre activation using rapamycin (S2E), demonstrating that translation is impaired in response to knocking out PfFKBP35.

Taken together, these results indicate that, under knock-out conditions, PfFKBP35 levels become limiting for ribosome maturation late during schizogony of G1, without having immediate effects on the steady-state levels of mature ribosomes. The knock-out of PfFKBP35 is thus phenotypically silent in G1. In absence of a continuous supply of mature ribosomes under FKBP35KO conditions, however, FKBP35KO parasites fail at synthesizing protein at levels that would support normal parasite development in the next generation (G2).

The knock-out of PfFKBP35 is transcriptionally silent

To examine if the altered proteome of FKBP35KO parasites is the result of PfFKBP35-controlled transcription, we compared the transcriptional profiles of FKBP35WT and FKBP35KO in a time course RNA-Sequencing (RNA-Seq) experiment spanning two IDCs. Specifically, we induced the knock-out of FKBP35 in synchronous parasites at 0-6 hpi and collected RNA from parasite populations at three and four subsequent time points (TPs) in G1 and G2, respectively (Figs. 3A, S3A). With the exception of TP7, these time points cover a time frame during which FKBP35KO cells cannot be distinguished from their FKBP35-expressing counterparts on a morphological level (compare Fig. 1E).

The knockout of PfFKBP35 remains silent at the transcriptomic level.

(A) Setup of the RNA-Seq experiment. RNA was collected at seven indicated time points (TP). (B) Volcano plots show differential gene expression between FKBP35KO and FKBP35WT parasites. Genes are color-coded according to the fold change and fdr (false discovery rate), n=3. (C) Mapping of FKBP35KO and FKBP35WT transcriptomes to a high-resolution reference (41). Spearman rank coefficients (ρ) show correlation between the sampled transcriptomes and the reference data set (41). The time points with the highest correlation coefficient are indicated.

As expected, the transcriptome of FKBP35KO showed a prominent reduction of steady-state fkbp35 transcripts already in TP1 (11.3-fold +/−2.0 s.d.) compared to the FKBP35WT control population (Fig. S3B, Table S2). During peak transcription at 30-36 hpi (TP2), this difference increased to 240-fold (238.4 +/−20.0 s.d.) and, consistent with the efficient recombination at the fkbp35 locus, these cells started transcribing gfp. Notwithstanding the prominent depletion of fkbp35 transcripts, the transcriptome of FKBP35KO is only marginally affected up to and including 24-30 hpi in G2 (TP6). By contrast, at 30-36 hpi (TP7) the transcription of knock-out parasites shows major changes (Fig. 3B). Considering the appearance of morphological changes in FKBP35KO at this time point, effects on the transcriptome are unlikely to represent specific activities of PfFKBP35 in gene regulation. Indeed, mapping the transcriptome of individual time points to a high resolution reference (41), revealed a significant slow-down of FKBP35KO compared to FKBP35WT parasites at TP7 (Figs. 3C, S3C). This indicates aborted cell cycle progression in FKBP35KO parasites during a phase that is characterized by a rapid increase in biomass – possibly due to a translation and/or size-dependent cell cycle checkpoint (42). At earlier time points (TP1-TP6), the transcriptomes of FKBP35KO and FKBP35WT are comparable and the differences observed on the level of individual genes clearly fail at explaining the proteomic changes observed in FKBP35KO parasites at 24-30 hpi in G2 (TP6), reiterating that the essential activity of PfFKBP35 is linked to post-transcriptional processes.

FKBP-targeting drugs fail at inhibiting the essential function of PfFKBP35

PfFKBP35 emerged as a promising antimalarial drug target (4) and considerable efforts were made to define interactions of this immunophilin with small molecules in the last decade (24–26, 43). Several studies conclusively demonstrated that FK506 – the most well-known ligand of mammalian FKBPs (44) – binds to PfFKBP35 (15, 21, 22) and kills asexual blood stage parasites with a half-maximal inhibitory concentration (IC50) in the low micromolar range at 1.9 µM (3).

Considering the proposed direct interaction between FK506 and FKBP35 in P. falciparum, we expected that FKBP35KO parasites show altered sensitivity to the drug. Consistent with published data, FK506 killed FKBP35WT cells with an IC50 of 1942 nM (+/−193 nM s.d.) when administered at 0-6 hpi (Fig. 4A). Surprisingly however, and despite the prominent depletion of the protein under knock-out conditions, FKBP35KO and FKBP35WT parasites were equally sensitive to FK506 (IC50 = 1741 +/− 187) (Fig. 4A). Similarly, neither the efficient knock-down of PfFKBP35 in NF54/iKD-FKBP35 parasites (compare Fig. S1C), nor its over-expression (OE) in NF54/iOE-FKBP35 cells (Fig. S4A-C), resulted in the expected sensitivity change to FK506 treatment (Fig. 4B). In addition to this apparent independence of the drug to varying PfFKBP35 levels, and in agreement with earlier studies (3, 24, 45), FK506 unfolds its antimalarial activity fairly quickly, i.e. within the same cycle after treatment. Standing in stark contrast to the delayed death phenotype of FKBP35KO parasites (compare Fig. 1E-G), this immediate activity on parasite survival substantiates the notion that FK506 must target essential parasite factors other than PfFKBP35.

CETSA identifies FK506 interaction partners.

(A) Dose-response effect of FK506 on FKBP35KO and FKBP35WT parasites. The knockout was induced at 0-6 hpi and parasites were exposed to FK506 for 48h. n=3, error bars represent the standard error of the mean. (B) Dose-response effect of FK506 on NF54/iKD-FKBP35 and NF54/iOE-FKBP35 parasites in presence and absence of Shield-1 (Shld). Parasites were split at 0-6 hpi and subsequently treated with Shld or the vehicle control EtOH. Data points represent the mean of three and two independent biological replicates for the NF54/iOE-FKBP35 and NF54/iKD-FKBP35 cell line, respectively. Error bars represent the standard error of the mean. (C) CETSA results. The relative shift in protein abundance (area under the curve of heat-challenged samples normalized against the non-denaturing control, ΔAUC) as a function of R2 (goodness of fit as a measure for the dose-response effect). The dashed lines indicate an arbitrary cutoff of 3 times the median absolute deviation (MAD) of each dataset and R2=0.8. Three different temperatures were used for protein denaturation (indicated by colors). (D) Dose-dependent stabilization of P. falciparum and human FKBPs detected using lysate and live cell CETSA. Bullets represent the abundance of non-denatured proteins in response to increasing FK506 concentrations under thermal challenge relative to the DMSO control. Colors indicate the temperatures used for heat challenge; the black curve represents the non-denaturing (37°C) control. (E) Proteins with the most pronounced stability shift under thermal challenge in the lysate CETSA experiment are listed. The EC50 is the concentration at which the half-maximal effect of the FK506-induced thermal stabilization is reached. The mutagenesis index score determined by Zhang et al. (18) is a measure to predict if a protein is essential for asexual growth: 0=essential, 1=non-essential. (F) Fitted live cell CETSA curves of 10 representative ribosomal proteins at 37°C and 55°C. Ribosomal proteins show a FK506 dose-dependent increase in soluble protein fraction. (G) Venn diagram showing the overlap between ribosomal proteins deregulated in the PfFKBP35 knock-out cell line (less abundant in G2 FKBPKO trophozoites determined by proteomics) and the ribosomal proteins stabilized by FK506 in live cell CETSA. The number of factors identified in either or both assays is indicated.

To rule out that the timing of drug exposure is masking expected links between FK506 and PfFKBP35, we tested whether administering FK506 at 30-36 hpi (instead of 0-6 hpi) affects parasite survival in a PfFKBP35-dependent manner. However, we did not observe a change compared to FKBP35WT parasites (Fig. S5A), ruling out the possibility that FKBP35KO parasites were killed by FK506 before PfFKBP35 levels were different in the two populations. Further, to account for a scenario in which the activity of FK506 on PfFKBP35 is masked by a temporal delay between drug exposure and parasite killing, we induced the knock-out of PfFKBP35 at 34-40 hpi (instead of 0-6 hpi) and allowed FKBP35KO parasites to proceed to G2, before exposing them to different FK506 concentrations. Despite these efforts, FKBP35WT and FKBP35KO parasites were equally sensitive to FK506 (Fig. S5B, S5C) and the calcineurin inhibitor cyclosporin A (Fig. S5D-E).

In addition to FK506, we evaluated the PfFKBP35-specific inhibitor D44 (N-(2-Ethylphenyl)- 2-(3h-Imidazo[4,5-B]pyridin-2-Ylsulfanyl)acetamide) but, surprisingly, we failed to reproduce the previously reported dose-response effect (24). We tested D44 from two suppliers using NF54/iKO-FKBP35, NF54 wild type and K1 parasite strains using different assay setups, including the gold standard method that is based on the incorporation of tritium labeled hypoxanthine into DNA of replicating cells (46). Still, we could not detect any activity of D44 on asexual replication of blood stage parasites (Figs. S5F-G).

Taken together, these results are in clear conflict with the perception that FK506 and D44 target the essential activity of PfFKBP35.

Probing FK506 interactions in Cellular Thermal Shift Assays (CETSA)

Considering the unaltered activity of FK506 in FKBP35KO compared to FKBP35WT parasites, we reasoned that FK506 must target proteins other than PfFKBP35. To identify potential targets of the drug in P. falciparum, we used the cellular thermal shift assay followed by mass spectrometry (MS-CETSA). This approach exploits the fact that, once drug-protein complexes are formed, thermostability of the target protein is altered (47). We applied the so-called isothermal dose-response variant of CETSA using FK506 (48). Here, target proteins are identified based on their dose-dependent stabilization by drugs under thermal challenge (different temperatures) when compared to a non-denaturing condition (37 °C), using the change in the area under the curve (ΔAUC) and the dose-response curve goodness of fit (R2) as metrics (Fig. 4C) (48).

CETSA identifies targets of FK506

In a first step, we performed CETSA in combination with drug-exposed protein extracts. Since proteins lose their physiological context during the process of protein extraction, this variant of the CETSA approach is designed to identify direct protein-drug interactions (47, 48). Not surprisingly, we found PfFKBP35 to be stabilized by FK506 in a dose-dependent manner when challenged at 60 °C with an EC50 of 152.0 nM (Fig. 4D). The fact that FK506 exhibits its stabilizing activity already in the low nanomolar concentration range indicates high affinity interactions between the protein-drug pair. This FK506-mediated stabilization of PfFKBP35 is thus achieved at drug concentrations that are magnitudes below the IC50 of FK506 in parasite survival assays (1942 nM +/−193 nM s.d.) (compare Fig. 4A). In line with several previous reports on the PfFKBP35/FK506 pair in vitro, our results show that, indeed, FK506 is binding to PfFKBP35, but in contrast to previous assumptions this interaction fails at explaining the antimalarial activity of FK506.

Besides PfFKBP35, FK506 stabilizes a number of other proteins in a dose-dependent manner (Figs. 4E and S6). A few of these factors show pronounced sigmoidal stabilization profiles and are predicted to be essential (Fig. 4E) (18). Amongst others, N6-methyl transferase MT-A70 (PF3D7_0729500), histone deacetylase HDA2 (PF3D7_1008000), pre-mRNA-splicing factor CEF1 (PF3D7_1033600), FeS cluster assembly protein SufD (PF3D7_1103400) and the H/ACA ribonucleoprotein complex subunit 1 (PF3D7_1309500) showed most pronounced stabilization profiles (Fig. S6 and Table S3). In contrast to PfFKBP35, most of these likely essential candidates are stabilized at FK506 concentrations close to the drug’s IC50 (compare Fig. 4A). It is therefore tempting to speculate that the binding of FK506 to at least one of these factors is responsible for the antimalarial activity of the drug.

The PfFKBP35/FK506 pair interferes with ribosomal complexes

Given the high-affinity interaction of PfFKBP35 with FK506 and its association with ribosomal proteins, we set out to describe the native context of PfFKBP35 using the live cell variant of CETSA. In contrast to focusing exclusively on direct protein-drug interactions, live cell CETSA allows capturing drug-induced shifts in protein stability beyond those of direct binding partners (48) and amongst others allows identifying factors present in the same protein complex as the target protein. To do so, CETSA takes advantage of the fact that factors within complexes have similar denaturation temperatures due to their co-aggregation upon heat-induced denaturation (49). As a consequence, this CETSA variant generally identifies more factors compared to the protein extract-based CETSA approach (Fig. 4C).

As expected, and confirming the results above, live cell CETSA shows that FK506 alters the thermal stability of PfFKBP35 already at very low drug concentrations when challenged at 60 °C with an EC50 of 286.7 nM (Fig. 4D). Interestingly, and in contrast to the protein lysate-based approach, live cell CETSA revealed that more than 50 ribosomal proteins are stabilized in presence of FK506 when challenged at 55°C, with a median EC50 of 45.9 nM (25th percentile = 34.2, 75th percentile = 58.9). Of note, these ribosomal proteins were stabilized at virtually identical FK506 concentrations (Figs. 4F, and S7), indicating that the drug – directly or indirectly – interacts with ribosomal complexes. This is in further agreement with the fact that these ribosomal factors were not identified by protein lysate-based CETSA (compare Figs. 4C, S6).

Worth mentioning, we observed a drug dose-dependent increase in soluble protein abundance of ribosomal factors also under non-denaturing conditions, i.e. at 37 °C, with a median EC50 of 1.94 µM (25th percentile = 1.85, 75th percentile = 2.1) (Figs. 4F, S7). Generally, due to the absence of a denaturing heat challenge, such a drug-dependent enrichment in soluble protein levels does not indicate binding-induced thermal stabilization of proteins (48). Instead, this can be the result of increased protein synthesis, altered protein conformation or, alternatively, occur in response to factors disengaging from larger protein complexes. The fact that FK506 solubilizes a high number of ribosomal proteins within a very narrow concentration window also at 37 °C, implies that these factors are liberated from the same ribosomal complex or from complexes that interact with each other and shows that FK506 negatively affects stability of ribosomal structures in live cells. However, these effects are limited to high, presumably saturating FK506 concentrations, likely explaining why the drug fails at killing parasites in a PfFKBP35-dependent manner. Importantly, the vast majority of ribosomal proteins identified in live cell CETSA are shared with those de-regulated in PfFKBP35 knock out cells (Fig. 4G), providing conclusive evidence for a shared localization of FK506 and its PfFKBP35 target in live parasites.

Together with the above-described importance of PfFKBP35 in ribosome homeostasis, these results strongly suggest that FK506 is recruited to the activity site of PfFKBP35 – most probably the nucleolar maturation sites of ribosomes – where it leads to a destabilization of ribosomal complexes.

Discussion

Members of the FKBP family are involved in regulating diverse cellular processes in eukaryotic as well as in prokaryotic cells (8, 50). In contrast to most other eukaryotes, the human malaria parasite P. falciparum encodes a single FKBP only (16). The enzymatic activity of PfFKBP35, as well as the interactions of the protein with FKBP-targeting drugs, was subject to considerable drug-focused research in vitro. Besides demonstrating the drug’s affinity to PfFKBP35 by co-crystallization, these efforts uncovered that FK506 inhibits enzymatic activity of the protein’s PPIase domain (5). Together with the anti-malarial activity of FK506, these data offered compelling evidence for an essential but yet uncharacterized role of PfFKBP35 during blood stage development of P. falciparum.

Here, we set out to investigate the role of PfFKBP35 in parasite biology. Using an inducible knock-out approach, we show that PfFKBP35 is indeed essential for intra-erythrocytic development of asexually replicating P. falciparum parasites and FKBP35KO cells die during the late trophozoite/early schizont stage. Surprisingly, these parasites display a delayed-death phenotype that is independent of apicoplast function and only manifests in the cycle following knock-out induction without affecting cell cycle progression in the first generation (G1) or multiplication rates from G1 to G2. This contradicts the known effects of FK506 and other PfFKBP35-targeting drugs for two main reasons. First, these drugs exert their parasiticidal activity within the same IDC (24). Second, neither the conditional knock-out, knock-down nor the overexpression of PfFKBP35 altered the parasite’s sensitivity to FK506, even though levels of the PfFKBP35 target are profoundly deregulated in the respective mutant lines. Together, these data show that the parasite-killing activity of FK506 is independent of the essential function of PfFKBP35. We hence aimed at the identification of other FK506 targets using CETSA, as discussed further down.

Knocking out PfFKBP35 early during intra-erythrocytic development has no effect on parasite morphology in G1 and provokes only moderate changes on the protein level compared to FKBP35WT parasites. While the effects in G1 are relatively subtle, it is noticeable that nucleolus-associated factors are found at higher abundance in knock-out cells. In the next cycle, FKBP35KO parasites show a dramatic decrease of factors associated with ribosome- and translation-related processes during trophozoite development (G2, 24-30 hpi). Consistently, protein translation in FKBP35KO parasites is reduced considerably. Importantly, at this point in time, FKBP35KO cells still show ordinary morphology and their transcriptional activity remains largely unaffected. The observed changes in the abundance of ribosomal factors must therefore result from post-transcriptional activity of PfFKBP35. While ribosomal components are present at higher levels in G1, they are drastically reduced in FKBP35KO cells in G2. These opposing effects are peculiar, but likely result from PfFKBP35 activity in a shared cellular process. Specifically, it is conceivable that low PfFKBP35 levels inhibit ribosome maturation in G1, explaining the accumulation of (pre-) ribosomal factors. While the levels of functional ribosomes are sufficient to support continued translation in G1, ribosomes likely become limiting in the next generation (G2), which eventually results in stalled translation activity and concomitant parasite death at the onset of schizogony; i.e. at a time when protein synthesis is required for facilitating the steep increase in biomass.

In contrast to the significant effects on the protein level, the transcription of FKBP35KO remains largely unaffected by the loss of PfFKBP35 and becomes apparent only during late trophozoite/early schizont development in G2 (30-36 hpi); i.e. at a time during which cell cycle progression is slowed down in FKBP35KO parasites. In fact, at this stage the transcriptional signature of FKBP35KO parasites is dominated by the slowed progression through intra-erythrocytic development, rather than resulting from specific gene regulatory events. While this stalling is consistent with the observed effects on ribosome homeostasis and protein translation, it demonstrates that, in contrast to certain FKBP family members in other organisms, PfFKBP35 does not have apparent roles in transcriptional control.

As discussed above, binding of FK506 to PfFKBP35 cannot explain the drug’s antimalarial activity and our attempts to identify targets of FK506 using resistance selection failed. This indicates that parasites do not readily acquire resistance mutations, possibly due to drug action on multiple targets (51). We hence aimed at identifying other FK506 targets using CETSA. In line with previous work on the drug in vitro, CETSA on P. falciparum protein extracts confirmed that FK506 is targeting PfFKBP35 with exceptionally high affinity in the low nanomolar range. Considering this well-characterized FK506-PfFKBP35 interaction (22) and the resulting inhibition of its PPIase activity (5), unaltered drug sensitivity of PfFKBP35 mutants indicates that PPIase activity is dispensable for blood stage parasites. Noteworthy, PPIase-independent functions of FKBPs have been described in other systems. For instance, PPIase activity of the Escherichia coli trigger factor (TF), a ribosome-associated FKBP, is not required for protein folding (52). Similarly, human FKBP52 maintains its function as a chaperone under PPIase-inhibiting conditions (9). While it is conceivable that PPIase activity is dispensable for intra-erythrocytic development of P. falciparum, the enzymatic activity of PfFKBP35 may be compensated by other factors, e.g. by PPIase domain-harboring members of the cyclophilin family (53).

CETSA performed on protein lysates identified a number of parasite factors binding to FK506. Compared to the PfFKBP35/FK506 pair, however, these interactions are of considerably lower affinity. The adenosine-methyltransferase MT-A70 (PF3D7_0729500) and the pre-mRNA-splicing factor CEF1 (PF3D7_1033600), for instance, were stabilized by FK506 with EC50 concentrations of 2.0 and 3.1 µM, respectively. While this affinity is relatively poor compared to the strong interaction formed between PfFKBP35 and FK506, one should be cautious designating the binding of FK506 to these proteins as irrelevant “off target” effects. In fact, their inhibition may well be masked by high-affinity binding of FK506 to PfFKBP35 – a possibility that shall briefly be discussed in the following section.

Assuming that FK506 is inhibiting the essential function of PfFKBP35, one would expect parasites to show a delayed-death phenotype similar to that observed in FKBP35KO parasites. In contrast to this delayed killing, the effect of FK506 on other essential targets likely manifests in the same cycle and would hence mask the effect of PfFKBP35-inhibition. Consistent with this possibility, FK506 kills P. falciparum parasites at 1.9 µM (+/−0.2 µM s.d.), i.e. at concentrations similar to those required for stabilizing MT-A70 and CEF1 under thermal challenge in CETSA. Our attempts to prove that FK506 indeed inhibits essential functions of PfFKBP35 failed: Neither knocking out PfFKBP35 at different time points nor adding FK506 to different stages of FKBP35KO parasites could provide evidence for a scenario in which PfFKBP35-dependent effects are simply masked by “off target” effects (compare Fig. S5). This indicates that either the binding of FK506 does not interfere with the essential role of PfFKBP35, or that PfFKBP35 is inhibited only at high FK506 concentrations that also inhibit other essential factors. In this context it is worth mentioning that parasites remain unaffected by a highly efficient knock-down of endogenous PfFKBP35 (observed with NF54/iKD-FKBP35; compare Figs. S1C-D), which demonstrates that very low protein levels are sufficient to maintain the essential function of PfFKBP35. It is thus conceivable that high FK506 concentrations are required to fully inhibit PfFKBP35 – a view that is further supported by data obtained with the live cell CETSA approach.

Besides identifying direct binding of drugs to their target, live cell CETSA allows capturing indirect effects on factors that act in the same pathway or in the same protein complex as the drug-bound target (48). While this approach confirmed the high affinity binding of FK506 to PfFKBP35, it also revealed that more than 50 ribosomal proteins share a very prominent CETSA signature: First, they are thermally stabilized by FK506 within a narrow concentration window of the drug. Second, higher levels of FK506 increase the solubility of these ribosomal components also at 37 °C. These results indicate that the FK506/PfFKBP35 pair interacts with ribosomal complexes and that FK506 is able to induce the dissociation of these ribosomal structures under physiological conditions. Importantly, and in contrast to the binding of FK506 to PfFKBP35, this effect on ribosomes is only observed at high concentrations of the drug. Together, these data likely explain the discrepancy observed between the high binding affinity of FK506 for PfFKBP35 and the low or even absent effect of the drug on PfFKBP35 function in blood stage parasites.

In line with our data, previous biochemical approaches revealed potential links between PfFKBP35 and ribosomes. Using co-immunoprecipitation, Leneghan and colleagues revealed that PfFKBP35 is interacting with 480 proteins, of which 31 are ribosomal factors (14). Indeed, FKBP family members are essentially involved in controlling ribosomes in other systems. The human FKBP variant HsFKBP25, for instance, is recruited to the pre-ribosome where its chaperone activity appears to be required for the assembly of the ribosomal large subunit (54). While the FKBP variant of yeast, Fpr4, interacts with the ribosome biogenesis factor Nop53 (55), the Escherichia coli trigger factor (TF) is part of the ribosomal complex and crucial for the folding of nascent proteins (56, 57). Considering the marked reduction of ribosomal factors in FKBP35KO parasites, it is tempting to speculate that PfFKBP35 is chaperoning ribosome biogenesis similar to Fpr4 and FKBP25 in Saccharomyces cerevisiae and humans, respectively.

The fact that PfFKBP35 localizes to distinct nuclear foci and its depletion causes dysregulation of ribosomes points towards a role within the nucleolus – the nuclear compartment driving ribosome biogenesis (38). Using IFAs, we found that PfFKBP35 does not co-localize with the nucleolar marker fibrillarin (58) (Fig. S8). However, its localization appears to be intertwined with that of this rRNA methyltransferase. Specifically, PfFKBP35 is mostly found in close vicinity to fibrillarin-rich regions. Nucleoli in eukaryotic cells generally consist of two or, in higher eukaryotes, three sub-compartments. Of those, the granular component (GC) represents the site of pre-ribosome assembly and encompasses fibrillarin-enriched central regions of nucleoli (59). The spatial correlation observed between PfFKBP35 and fibrillarin may thus indicate localization to different sub-nucleolar compartments, with PfFKBP35 occupying the GC area of the nucleolus.

In summary, we demonstrate that limiting PfFKBP35 levels are lethal to P. falciparum and result in a delayed-death phenotype that is characterized by perturbed ribosome homeostasis and defective protein translation, which is likely linked to the reduced biogenesis of functional ribosomes in the nucleolus. We further show that FKBP-binding drugs, including FK506, exert their parasiticidal activity in a PfFKBP35-independent manner, urging strong caution for the future development of FKBP-targeting antimalarials, especially when based on FK506 and structural derivatives thereof. In addition to revealing first insights into the essential function of PfFKBP35 in ribosome homeostasis of P. falciparum, we are convinced that the presented data offer valuable information for target-based efforts in malaria drug discovery.

Methods

Parasite culture

P. falciparum culture and synchronization was performed as described (60, 61). Parasites were cultured in AB+ or B+ human red blood cells (Blood Donation Center, Zürich, Switzerland) at a hematocrit of 5% in culture medium containing 10.44 g/L RPMI-1640, 25 mM HEPES, 100 μM hypoxanthine, 24 mM sodium bicarbonate, 0.5% AlbuMAX II (Gibco #11021-037), 0.1 g/L neomycin, and 2 mM choline chloride (Sigma #C7527). To achieve stabilization of DD-tagged proteins in the cell lines NF54/iKD-FKBP35 and NF54/iOE-FKBP35, parasites were cultured in presence of 625 nM Shield-1 (30, 31). To induce the DiCre-mediated recombination of loxP sites in NF54/iKO-FKBP35 parasites, cultures were split and treated for 4 h with 100 nM rapamycin or the corresponding volume DMSO, which served as vehicle control, giving rise to FKBP35KO and FKBP35WT populations, respectively (32). Cultures were gassed with 3% O2, 4% CO2 and 93% N2 and incubated in an airtight incubation chamber at 37°C.

Cloning of transfection constructs

CRISPR/Cas9-based gene editing of the NF54 parasites was performed using a two-plasmid approach as previously described (62, 63). This system is based on co-transfection of a plasmid that contains the expression cassettes for the Cas9 enzyme, the single guide RNA (sgRNA) and the blasticidin deaminase (BSD) resistance marker (pBF-gC), and a pD-derived donor plasmid that contains the template for the homology-directed repair of the Cas9-induced DNA double strand break (Fig. S2A) (63).

The plasmids pBF-gC_FKBP-3’, pBF-gC_FKBP-5’ and pBF-gC_P230p targeting the 3’ or 5’ end of fkbp35 (PF3D7_1247400) or the p230p (PF3D7_0208900) locus (64), respectively, were generated by ligation of two annealed oligonucleotides (gRNA_top and gRNA_bottom) into the BsaI-digested pBF-gC backbone (63) using T4 DNA ligase (New England Biolabs).

The donor plasmid pD_FKBP35-iKO was generated by assembling six DNA fragments in a Gibson reaction (65) using (i) the plasmid backbone amplified by polymerase chain reaction (PCR) from pUC19 (primers PCRa_F/PCRa_R) (63), (ii/iii) the 5’ and 3’ homology regions amplified from genomic DNA (primers PCRb_F/PCRb_R and PCRc_F/PCRc_R, respectively), (iv) the green fluorescent protein (gfp) sequence amplified from pHcamGDV1-GFP-DD (primers PCRd_F/PCRd_R) (63), (v) the HRPII terminator amplified from genomic DNA (primers PCRe_F/PCRe_R), and (vi) the fkbp35 sequence containing an artificial loxPint (33) amplified from a P. falciparum codon-optimized synthetic sequence (primers PCRf_F/PCRf_R) ordered from Genscript.

The donor plasmid pD_FKBP35-iKD was generated by assembling five DNA fragments in a Gibson reaction (65) using (i) the plasmid backbone amplified from pUC19 (primers PCRa_F/PCRa_R) (63), (ii/iii) the 5’ and 3’ homology regions amplified from genomic DNA (primers PCRg_F/PCRg_R and PCRh_F/PCRh_R, respectively), (iv) the gfp sequence amplified from pHcamGDV1-GFP-DD (primers PCRi_F/PCRi_R) (63), and (v) the destabilization domain (dd) sequence amplified from pHcamGDV1-GFP-DD (primers PCRj_F/PCRj_R) (63).

The donor plasmid pD_FKBP35-iOE was generated by assembling three DNA fragments in a Gibson reaction (65) using (i) the NheI/PstI-digested plasmid pkiwi003 (66), (ii) the fkbp35 coding sequence amplified from genomic DNA (primers PCRk_F/PCRk_R), and (iii) the gfp-dd sequence amplified from pHcamGDV1-GFP-DD (primers PCRl_F/PCRl_R) (63).

Oligonucleotides are listed in Table S4.

Transfection and selection of gene-edited parasites

P. falciparum transfection using the CRISPR/Cas9 two-plasmid approach was performed as described previously (63). Briefly, 50 μg of each plasmid (pBF-gC_FKBP-3’ and pD_FKBP35-iKO, pBF-gC_FKBP-5’ and pD_FKBP35-iKD, pBF-gC_P230p and pD_FKBP35-iOE) were co-transfected into an NF54::DiCre line (67), in which ap2-g was tagged with the fluorophore mScarlet as described previously (68). Transgenic parasites were selected with 2.5 μg/mL blasticidin-S-hydrochloride, which was added 24 h after transfection for 10 days. Transgenic populations were usually obtained 2-3 weeks after transfection and correct editing of the modified loci was confirmed by PCR on gDNA (Figs. S1B, S2B and S4B). Primer sequences used for these PCRs are listed in Table S4. Clonal parasite lines were obtained by limiting dilution cloning (69).

DNA content analysis

NF54/iKO-FKBP35 parasites were fixed in 4% formaldehyde/0.0075% glutaraldehyde for 30 min at room temperature and washed 3 times in PBS and stored at 4°C. Samples were collected from three independent biological replicates. RNA was digested by incubation with 0.1 mg/mL RNAse A in PBS containing 0.1% Triton-X100 for 15 min at room temperature. Nuclei were stained using 1X SYBR Green 1 DNA stain (Invitrogen S7563) for 20 min and washed three times in PBS. The fixation step quenched the GFP signal detected by the flow cytometer (Fig. S9). SYBR Green intensity was measured using a MACS Quant Analyzer 10 and analyzed using the FlowJo_v10.6.1 software.

Western blotting

Saponin lysis of 500 µL infected red blood cells at 3-5% parasitaemia was performed by incubation for 10 min on ice in 3 mL ice-cold 0.15% saponin in PBS. The parasite pellets were washed two times in ice-cold PBS and resuspended in Laemmli buffer (62.5 mM Tris base, 2% SDS, 10% glycerol and 5% 2-mercaptoethanol). Proteins were separated on 4-12% Bis-Tris gels (Novex, Qiagen) using MOPS running buffer (Novex, Qiagen). Then proteins were transferred to a nitrocellulose membrane (GE healthcare #106000169), which was blocked with 5% milk powder in PBS/0.1% Tween (PBS-T) for 1 h. The membrane was probed using the primary antibodies mAb mouse α-GFP (1:1’000, Roche Diagnostics #11814460001), mAb mouse α-PfGAPDH (1:20’000) (70), or mAb mouse α-puromycin (1:5’000, Sigma MABE343) diluted in blocking solution. After 2 h incubation, the membrane was washed 5 times in PBS-T before it was incubated with the secondary antibody goat α-mouse IgG (H&L)-HRP (1:10’000, GE healthcare #NXA931). After 1 h incubation, the membrane was washed 5 times and the signal was detected using the chemiluminescent substrate SuperSignal West Pico Plus (Thermo Scientific, REF 34580) and the imaging system Vilber Fusion FX7 Edge 17.10 SN. Nitrocellulose membranes were stripped by incubating for 10 min in a stripping buffer (1 g/L SDS, 15 g/L glycine, 1% Tween-20, pH 2.2) and washed three times in PBS-T before blocking and re-probing.

Drug dose-response experiments

Dose-response relationships were assessed by exposing synchronous ring stage parasites at 0.5% parasitaemia and 1.25% hematocrit to 12 drug concentrations using a two-step serial dilution in 96-well plates (Corning Incorporated, 96-well cell culture plate, flat bottom, REF 3596) in three independent biological replicates. The plates were gassed and incubated for 48 h at 37 °C. FK506 (MedChemExpress HY-13756), D44 (ChemBridge 7934155 and ChemDiv 7286-2836) and cyclosporin A (Sigma 30024) stocks of 10 mM were prepared in DMSO and working solutions were prepared immediately before the experiment. To measure parasite survival using flow cytometry, cultures were stained in 1X SYBR Green 1 DNA stain (Invitrogen S7563) in a new 96-well plate (Corning Incorporated, 96-well cell culture plate, round bottom, REF 3788) and incubated in the dark for 20 min. The samples were washed once in PBS. Using a MACS Quant Analyzer 10, 50’000 events per condition were measured and analyzed using the FlowJo_v10.6.1 software. Infected RBCs were identified based on SYBR Green 1 DNA stain intensity. The gating strategy is shown in Fig. S9. To quantify parasite survival based on hypoxanthine incorporation into DNA, compounds were dissolved in DMSO (10 mM), diluted in hypoxanthine-free culture medium and titrated in duplicates over a 64-fold range (6 step twofold dilutions) in 96 well plates. 100 μl asexual parasite culture (prepared in hypoxanthine-free medium) were added to each well and mixed with the compound to obtain a final haematocrit of 1.25% and a final parasitemia of 0.3%. After incubation for 48 hrs, 0.25 μCi of [3H]-hypoxanthine was added per well and plates were incubated for an additional 24 hrs. Parasites were then harvested onto glass-fiber filters using a Microbeta FilterMate cell harvester (Perkin Elmer, Waltham, US) and radioactivity was counted using a MicroBeta2 liquid scintillation counter (Perkin Elmer, Waltham, US). The results were recorded and expressed as a percentage of the untreated controls. Curve fitting and IC50 calculations were performed using a non-linear, four-parameter regression model with variable slope (Graph Pad Prism, version 8.2.1).

Fluorescence microscopy

To localize PfFKBP35, NF54/iKD-FKBP35 parasites were fixed in 4% formaldehyde/0.0075% glutaraldehyde for 30 min at room temperature and washed three times in PBS. The fixed cells were permeabilized by incubation with 0.1% Triton-X100-containing PBS for 15 min. After two washing steps in PBS, the cells were incubated in a blocking/aldehyde quenching solution (3% BSA in PBS complemented with 50 mM ammonium chloride). The samples were incubated with the primary antibodies rabbit α-GFP (1:400, Abcam ab6556) and mAb mouse α-fibrillarin (1:100, Santa Cruz Biotechnology sc-166021) for 1 h in 3% BSA in PBS and washed 3 times in PBS. Subsequently, the samples were incubated with the secondary antibodies goat α-rabbit IgG (H&L) Alexa Fluor 488 (1:250, Invitrogen A11008) and goat a-mouse IgG (H&L) Alexa Fluor 594 (1:250, Invitrogen A11032), for 1 h and washed 3 times, before they were mixed with Vectashield containing DAPI (Vector laboratories, H-1200), and mounted on a microscopy slide. Slides were imaged using a Leica THUNDER 3D Assay imaging system.

Proteomics

Parasite cultures were split at 0-6 hpi in G1 and treated for 4 h with 100 nM rapamycin. Saponin lysis of paired FKBP35WT and FKBP35KO populations at 36-42 hpi in G1 and 24-30 hpi in G2 was performed as described above. Samples were collected in three independent biological replicates. The washed parasite pellet was snap frozen in liquid nitrogen. After thawing, parasites were resuspended in 25 µL of lysis buffer (5% SDS, 100 mM tetraethylammonium bromide (TEAB), 10 mM tris(2-carboxyethyl)phosphin (TCEP), pH 8.5) and sonicated (10 cycles, 30 sec on/off at 4 °C, Bioruptor, Diagnode). Lysates were subsequently reduced for 10 minutes at 95 °C. Samples were then cooled down to RT and 0.5 µL of 1M iodoacetamide was added to the samples. Cysteine residues were alkylated for 30 min at 25 °C in the dark. Digestion and peptide purification was performed using S-trap technology (Protifi) according to the manufacturer’s instructions. In brief, samples were acidified by addition of 2.5 µL of 12% phosphoric acid (1:10) and then 165 µL of S-trap buffer (90% methanol, 100 mM TEAB pH 7.1) was added to the samples (6:1). Samples were briefly vortexed and loaded onto S-trap micro spin-columns (Protifi) and centrifuged for 1 min at 4000 g. Flow-through was discarded and spin-columns were then washed 3 times with 150 µL of S-trap buffer (each time samples were centrifuged for 1 min at 4000 g and flow-through was removed). S-trap columns were then moved to the clean tubes and 20 µL of digestion buffer (50 mM TEAB pH 8.0) and trypsin (at 1:25 enzyme to protein ratio) were added to the samples. Digestion was allowed to proceed for 1h at 47 °C. Then, 40 µL of digestion buffer were added to the samples and the peptides were collected by centrifugation at 4000 g for 1 minute. To increase the recovery, S-trap columns were washed with 40 µL of 0.2% formic acid in water (400g, 1 min) and 35 µL of 0.2% formic acid in 50% acetonitrile. Eluted peptides were dried under vacuum and stored at - 20 °C until further analysis.

Peptides were resuspended in 0.1% aqueous formic acid and peptide concentration was adjusted to 0.25 µg/µL. 1 µL of each sample was subjected to LC-MS/MS analysis using an Orbitrap Elicpse Tribrid Mass Spectrometer fitted with an Ultimate 3000 nano system (both from Thermo Fisher Scientific) and a custom-made column heater set to 60 °C. Peptides were resolved using a RP-HPLC column (75 μm × 30 cm) packed in-house with C18 resin (ReproSil-Pur C18–AQ, 1.9 μm resin; Dr. Maisch GmbH) at a flow rate of 0.3 μL/min. The following gradient was used for peptide separation: from 2% buffer B to 12% B over 5 min, to 30% B over 40 min, to 50% B over 15 min, to 95% B over 2 min followed by 11 min at 95% B then back to 2% B. Buffer A was 0.1% formic acid in water and buffer B was 80% acetonitrile, 0.1% formic acid in water.

The mass spectrometer was operated in DDA mode with a cycle time of 3 seconds between master scans. Each master scan was acquired in the Orbitrap at a resolution of 240,000 FWHM (at 200 m/z) and a scan range from 375 to 1600 m/z followed by MS2 scans of the most intense precursors in the linear ion trap at “Rapid” scan rate with isolation width of the quadrupole set to 1.4 m/z. Maximum ion injection time was set to 50 ms (MS1) and 35 ms (MS2) with an AGC target set to 1e6 and 1e4, respectively. Only peptides with charge state 2-5 were included in the analysis. Monoisotopic precursor selection (MIPS) was set to Peptide, and the Intensity Threshold was set to 5e3. Peptides were fragmented by HCD (Higher-energy collisional dissociation) with collision energy set to 35%, and one microscan was acquired for each spectrum. The dynamic exclusion duration was set to 30 seconds.

The acquired raw-files were imported into the Progenesis QI software (v2.0, Nonlinear Dynamics Limited), which was used to extract peptide precursor ion intensities across all samples applying the default parameters. The generated mgf-file was searched using MASCOT against a Plasmodium falciparum (isolate 3D7) database (Uniprot, 11.2019) and 392 commonly observed contaminants using the following search criteria: full tryptic specificity was required (cleavage after lysine or arginine residues, unless followed by proline); 3 missed cleavages were allowed; carbamidomethylation (C) was set as fixed modification; oxidation (M) and acetyl (Protein N-term) were applied as variable modifications; mass tolerance of 10 ppm (precursor) and 0.6 Da (fragments). The database search results were filtered using the ion score to set the false discovery rate (FDR) to 1% on the peptide and protein level, respectively, based on the number of reverse protein sequence hits in the dataset. Quantitative analysis results from label-free quantification were processed using the SafeQuant R package v.2.3.2 (71) to obtain peptide relative abundances. This analysis included global data normalization by equalizing the total peak/reporter areas across all LC-MS runs, data imputation using the knn algorithm, summation of peak areas per protein and LC-MS/MS run, followed by calculation of peptide abundance ratios. Only isoform specific peptide ion signals were considered for quantification. To meet additional assumptions (normality and homoscedasticity) underlying the use of linear regression models and t-Tests, MS-intensity signals were transformed from the linear to the log-scale. The summarized peptide expression values were used for statistical testing of between condition differentially abundant peptides. Here, empirical Bayes moderated t-Tests were applied, as implemented in the R/Bioconductor limma package (72). The resulting per protein and condition comparison p-values were adjusted for multiple testing using the Benjamini-Hochberg method.

Gene ontology enrichment analysis was performed using the PANTHER Overrepresentation Test with the correction method “False Discovery Rate”, the test type “Fisher’s Exact”, and the annotation data set “PANTHER GO-Slim” (37).

IPP complementation

Parasite cultures were split at 0-6 hpi in G1 and treated for 4 h with 100 nM rapamycin. FKBP35KO and FKBP35WT populations were split again and supplemented either with 200 µM isopentenyl pyrophosphate (IPP) (Sigma I0503) or the corresponding volume H2O. The respective culture medium was replaced daily. Parasites treated with 156 ng/mL doxycycline (Sigma D9891) were used to confirm that IPP complementation is able to rescue apicoplast-deficient parasites. DNA content analysis was performed as described above.

Heat shock susceptibility testing

Parasites cultures were split at 0-6 hpi in G1 and treated for 4 h with 100 nM rapamycin. While control populations were always incubated at 37 °C, test populations were incubated at 40 °C for 6 h starting at 24-30 hpi. Parasitaemia was measured in G2 using flow cytometry as described above.

Surface sensing of translation (SUnSET)

Parasite cultures were split at 0-6 hpi in G1 and treated for 4 h with 100 nM rapamycin. At 24-30 hpi in G2, the cultures were incubated with 1 µg/mL puromycin (Sigma P8833) for 1 h at 37 °C (40, 73). Saponin lysis and Western blotting were performed as described above.

Cellular thermal shift assay (CETSA)

The mass spectrometry-based isothermal dose-response cellular thermal shift assay (ITDR-MS-CETSA) was performed according to the protocol developed by Dziekan and colleagues (48). Briefly, live parasites (107 MACS-purified trophozoites per condition) were exposed to a FK506 (MedChemExpress HY-13756) concentration gradient (100 µM to 1.5 nM) and a DMSO control for 1 h. Subsequently, the parasites were washed in PBS and exposed to the denaturing temperatures 50 °C, 55 °C and 60 °C as well as to a non-denaturing temperature 37 °C for 3 min, before they were cooled down to 4 °C for 3 min. Afterwards, the cells were lysed by resuspending in a lysis buffer (50 mM HEPES pH 7.5, 5 mM β-glycerophosphate, 0.1 mM Na3VO4, 10 mM MgCl2, 2 mM TCEP, and EDTA-free protease inhibitor cocktail (Sigma)), three freeze/thaw cycles, and mechanical shearing using a syringe with a 31-gauge needle. Subsequently, insoluble proteins were pelleted by centrifugation (20 min at 4 °C and 20’000 g) and the soluble fraction was collected and subjected to protein digestion, reduction, alkylation, and TMT10 labelling followed by LC/MS analysis as described (48).

To prepare the protein lysate, saponin-lysed parasites were lysed in the afore-mentioned buffer by three freeze/thaw cycles and mechanical shearing. After centrifugation, soluble proteins were exposed to a FK506 concentration gradient (200 µM to 3 nM) and a DMSO control for 1 min before the thermal challenge was performed as described above. Afterwards, the soluble protein fraction was isolated and prepared for by LC/MS analysis as described (48).

Data analysis was performed using the Proteome Discoverer 2.1 software (Thermo Fisher Scientific) and the R package “mineCETSA” (v 1.1.1). Only proteins identified by at least three peptide spectrum matches (PSMs) were included in the analysis (48).

Transcriptomics

Sample preparation

The RNA of 500 µL iRBCs at 3-5% parasitaemia was isolated using 3 mL TRIzol reagent (Invitrogen) followed by purification using the Direct-zol RNA MiniPrep kit (Zymo). Samples were collected from three biological replicates. Stranded RNA sequencing libraries were prepared using the Illumina TruSeq stranded mRNA library preparation kit (REF 20020594) according to the manufacturer’s protocol. The library was sequenced in 100 bp paired-end reads on an Illumina NextSeq 2000 sequencer using the Illumina NextSeq 2000 P3 reagents (REF 20040560), resulting in 5 million reads on average per sample.

RNA-Seq data analysis

Fastqc was run on all raw fastq files. Then, each file was aligned with hisat2 (version 2.0.5) (74) against the 3D7 reference genome (PlasmoDB v58) (75) in which the GFP and the recoded FBPK35 sequences were added at the end of chromosome 12. Read counts for each gene and for each sample were then counted using featureCount (gff file from PlasmoDB v58 modified to include the added GFP and FBPK35). TPM (Transcripts Per Kilobase Million) were then calculated from the raw count matrix.

Principal component analysis (PCA) was performed using the R (version 4.2.1) function prcomp using log transformed TPMs. PCA revealed that the rapamycin-treated replicate B of TP 3 (“3B.R”) was a clear outlier and is different from any other replicate at any time point (Fig. S3D). For this reason, it was excluded from further analysis. The PCA was done again without it and the other samples show the same pattern as the initial PCA (Fig. S3E).

DESeq2 (version 1.36) (76) was used for normalization and differential expression analysis. Paired differential expression tests where done for each time point using DESeq2 LRT (likelihood ratio test) using the full model ∼ Treatment + Replicate (Treatment is DMSO or Rapamycin and Replicate is the clone, added as the data is paired) and the reduced model ∼ Replicate. From this paired test and the average log fold-change calculated from the TPMs, normalized volcano plots were made using the R package EnhancedVolcano (version 1.14) (77).

Cell cycle analysis

Spearman correlation between normalized scaled read counts for each sample for all genes (as one vector) and each time point of the reference microarray time course published by Bozdech et al. (41) (3D7 smoothed, retrieved from PlasmoDB) were tested. Spearman’s coefficients rho for each time point of this RNA-Seq time course and each time point of the reference data set where plotted using ggcorrplot (version 0.1.3) (78). For each time point of this RNA-Seq dataset, the rho coefficients where then plotted against the time points of the reference data set and the FKBP35WT and FKBP35KO data were fitted separately to the linearized sinusoidal using R lm function. The time point matching best to the FKBP35WT and FKBP35KO samples for each time point of the RNA-Seq time course was taken as the maximum of this fitted function. Both fitted functions were F-tested (using R function var.test using both fitted model as applied to the FKBP35KO data) to test whether the FKBP35WT and FKBP35KO data correlated with the reference data set (41) in a significantly different way, which would indicate a cell cycle shift if statistically significant. Only the last time point showed such a significant difference.

Data availability

The mass spectrometry proteomics data have been deposited to the ProteomeXchange Consortium repository with the dataset identifier PXD039018. The sequencing data are available in the Sequence Read Archive (SRA) via the accession number PRJNA914079.

Supporting information

eLife Assessment

This important study addresses both the native role of the Plasmodium falciparum protein PfFKBP35 and whether this protein is the target of FK506, an immunosuppressant with antiplasmodial activity. The genetic evidence for the essentiality of FKBP35 in parasite growth is compelling. However, the conclusion that the role of FKBP35 is to secure ribosome homeostasis and the claim that FK506 exerts its antimalarial activity independently of FKBP35 rely on incomplete evidence.

Reviewer #1 (Public Review):

In this study, the authors investigate the biological function of the FK506-binding protein FKBP35 in the malaria-causing parasite Plasmodium falciparum. Like its homologs in other organisms, PfFKBP35 harbors peptidyl-prolyl isomerase (PPIase) and chaperoning activities, and has been considered a promising drug target due to its high affinity to the macrolide compound FK506. However, PfFKBP35 has not been validated as a drug target using reverse genetics, and the link between PfFKBP35-interacting drugs and their antimalarial activity remains elusive. The manuscript is structured in two parts addressing the biological function of PfFKBP35 and the antimalarial activity of FK506, respectively.

The first part combines conditional genome editing, proteomics and transcriptomics analysis to investigate the effects of FKBP35 depletion in P. falciparum. The work is very well performed and clearly described. The data provide definitive evidence that FKBP35 is essential for P. falciparum blood stage growth. Conditional knockout of PfFKBP35 leads to a delayed death phenotype, associated with defects in ribosome maturation as detected by quantitative proteomics and stalling of protein synthesis in the parasite. The authors propose that FKBP35 regulates ribosome homeostasis but an alternative explanation could be that changes in the ribosome proteome are downstream consequences of the abrogation of FKBP35 essential activities as chaperone and/or PPIase. It is unclear whether FKBP35 has a specific function in P. falciparum as compared to other organisms. The knockdown of PfFKBP35 has no phenotypic consequence, showing that very low amounts of FKBP35 are sufficient for parasite survival and growth. In the absence of quantification of the protein during the course of the experiments, it remains unclear whether the delayed death phenotype in the knockout is due to the delayed depletion of the protein or to a delayed consequence of early protein depletion. This limitation also impacts the interpretation of the drug assays.

In the second part, the authors investigate the activity of FK506 on P. falciparum, and conclude that FK506 exerts its antimalarial effects independently of FKBP35. This conclusion is based on the observation that FK506 has the same activity on FKBP35 wild type and knock-out parasites, suggesting that FK506 activity is independent of FKBP35 levels, and on the fact that FK506 kills the parasite rapidly whereas inducible gene knockout results in delayed death phenotype. However, there are alternative explanations for these observations. As mentioned above, the delayed death phenotype could be due to delayed depletion of the protein upon induction of gene knockout. FK506 could have a similar activity on WT and mutant parasites when added before sufficient depletion of FKBP35 protein. In some experiments, the authors exposed KO parasites to FK506 later, presumably when the KO is effective, and obtained similar results. However, in these conditions, the death induced by the knockout could be a confounding factor when measuring the effects of the drug. Furthermore, the authors show that FK506 binds to FKBP35, and propose that the FK506-FKBP35 complex interferes with ribosome maturation, which would point towards a role of FKBP35 in FK506 action. In summary, the study does not provide sufficient evidence to rule out that FK506 exerts its effects via FKBP35.

Reviewer #2 (Public Review):

The manuscript by Thomen et al. FKBP secures ribosome homeostasis in Plasmodium falciparum and focuses on the importance of PfKBP35 protein, its interaction with the FK506 compound, and the role of PfKBP35 in ribosome biogenesis. The authors showed the interaction of the PfKBP54 with FK506, but the part of the FK506 and PfKBP54 in ribosome biogenesis based on the data is unclear.

The introduction is plotted with two parallel stories about PfKBP35 and FK506, with ribosome biogenesis as the central question at the end. In its current form, the manuscript suffers from two stories that are not entirely interconnected, unfinished, and somewhat confusing. Both stories need additional experiments to make the manuscript(s) more complete. The results from PfFBP35 need more evidence for the proposed ribosome biogenesis pathway control. On the other hand, the results from the drug FK506 point to different targets with lower EC50, and other follow-up experiments are needed to substantiate the authors' claims.

The strengths of the manuscript are the figures and experimental design. The combination of omics methods is informative and gives an opportunity for follow-up experiments.

Reviewer #3 (Public Review):

The study by Thommen et al. sought to identify the native role of the Plasmodium falciparum FKBP35 protein, which has been identified as a potential drug target due to the antiplasmodial activity of the immunosuppressant FK506. This compound has multiple binding proteins in many organisms; however, only one FKBP exists in P. falciparum (FKBP35). Using genetically-modified parasites and mass spectrometry-based cellular thermal shift assays (CETSA), the authors suggest that this protein is in involved in ribosome homeostasis and that the antiplasmodial activity of FK506 is separate from its activity on the FKBP35 protein. The authors first created a conditional knockdown using the destruction domain/shield system, which demonstrated no change in asexual blood stage parasites. A conditional knockout was then generated using the DiCre system. FKBP35KO parasites survived the first generation but died in the second generation. The authors called this "a delayed death phenotype", although it was not secondary to drug treatment, so this may be a misnomer. This slow death was unrelated to apicoplast dysfunction, as demonstrated by lack of alterations in sensitivity to apicoplast inhibitors. Quantitative proteomics on the FKBP35KO vs FKBP35WT parasites demonstrated enrichment of proteins involved in pre-ribosome development and the nucleolus. Interestingly, the KO parasites were not more susceptible to cycloheximide, a translation inhibitor, in the first generation (G1), suggesting that mature ribosomes still exist at this point. The SunSET technique, which incorporates puromycin into nascent peptide chains, also showed that in G1 the FKBP35KO parasites were still able to synthesize proteins. But in the second generation (G2), there was a significant decrease in protein synthesis. Transcriptomics were also performed at multiple time points. The effects of knockout of FKBP35 were transcriptionally silent in G1, and the parasites then slowed their cell cycles as compared to the FKBP35WT parasites.

The authors next sought to evaluate whether killing by FK506 was dependent upon the inhibition of PfKBP35. Interestingly, both FKBP35KO and FKBP35WT parasites were equally susceptible to FK506. This suggested that the antiplasmodial activity of FK506 was related to activity targeting essential functions in the parasite separate from binding to FKBP35. To identify these potential targets, the authors used MS-CETSA on lysates to test for thermal stabilization of proteins after exposure to drug, which suggests drug-protein interactions. As expected, FK506 bound FKBP35 at low nM concentrations. However, given that the parasite IC50 of this compound is in the uM range, the authors searched for proteins stabilized at these concentrations as putative secondary targets. Using live cell MS-CETSA, FK506 bound FKBP35 at low nM concentrations; however, in these experiments over 50 ribosomal proteins were stabilized by the drug at higher concentrations. Of note, there was also an increase in soluble ribosomal factors in the absence of denaturing conditions. The authors suggested that the drug itself led to these smaller factors disengaging from a larger ribosomal complex, leading to an increase in soluble factors. Ultimately, the authors conclude that the native function of FKBP35 is involved in ribosome homeostasis and that the antiplasmodial activity of FK506 is not related to the binding of FKBP35, but instead results from inhibition of essential functions of secondary targets.

Strengths:

This study has many strengths. It addresses an important gap in parasite biology and drug development, by addressing the native role of the potential antiplasmodial drug target FKBP35 and whether the compound FK506 works through inhibition of that putative target. The knockout data provide compelling evidence that the KBP35 protein is essential for asexual parasite growth after one growth cycle. Analysis of the FKBP35KO line also provides evidence that the effects of FK506 are likely not solely due to inhibition of that protein, but instead must have secondary targets whose function is essential. These data are important in the field of drug development as they may guide development away from structure-based FK506 analogs that bind more specifically to the FKBP35 protein.

Weaknesses:

There are also a few notable weaknesses in the evidence that call into question the conclusion in the article title that FKBP35 is definitely involved in ribosomal homeostasis. While the proteomics supports alterations in ribosome biogenesis factors, it is unclear whether this is a direct role of the loss of the FKBP35 protein or is more related to non-specific downstream effects of knocking down the protein. The CETSA data clearly demonstrate that FK506 binds PfKB35 at low nM concentrations, which is different than the IC50 noted in the parasite; however, the evidence that the proteins stabilized by uM concentrations of drug are actual targets is not completely convincing. Especially, given the high uM amounts of drug required to stabilize these proteins. This section of the manuscript would benefit from validation of a least one or two of the putative candidates noted in the text. In the live cell CETSA, it is noted that >50 ribosomal components are stabilized in drug treated but not lysate controls. Similarly, the authors suggest that the -soluble fraction of ribosomal components increases in drug-exposed parasites even at 37{degree sign}C and suggests that this is likely from smaller ribosomal proteins disengaging from larger ribosomal complexes. While the evidence is convincing that this protein may play a role in ribosome homeostasis in some capacity, it is not sure that the title of the paper "FKBP secures ribosome homeostasis" holds true given the lack of mechanistic data. A minor weakness, but one that should nonetheless be addressed, is the use of the term "delayed death phenotype" with regards to the knockout parasite killing. This term is most frequently used in a very specific setting of apicoplast drugs that inhibit apicoplast ribosomes, so the term is misleading. It is also possible that the parasites are able to go through a normal cycle because of the kinetics of the knockout and that the time needed for protein clearance in the parasite to a level that is lethal.

Overall, the authors set out to identify the native role of FKB35 in the P. falciparum parasites and to identify whether this is, in fact, the target of FK506. The data clearly demonstrate that FKBP35 is essential for parasite growth and provide evidence that alterations in its levels have proteomic but not transcriptional changes. However, the conclusion that FKBP35 actually stabilizes ribosomal complexes remains intermediate. The data are also very compelling that FK506 has secondary targets in the parasite aside from FKBP35; however, the high uM concentrations of the drug needed to attain results and the lack of biological validation of the CETSA hits makes it difficult to know whether any of these are actually the target of the compound or instead are nonspecific downstream consequences of treatment.

Author Response

The following is the authors’ response to the original reviews.

eLife assessment

This important study addresses both the native role of the Plasmodium falciparum protein PfFKBP35 and whether this protein is the target of FK506, an immunosuppressant with antiplasmodial activity. The genetic evidence for the essentiality of FKBP35 in parasite growth is compelling. However, the conclusion that the role of FKBP35 is to secure ribosome homeostasis and the claim that FK506 exerts its antimalarial activity independently of FKBP35 rely on incomplete evidence.

We thank the Reviewers and Editors for their careful evaluation of our manuscript and the constructive criticism. We realized that some of our conclusions may be regarded/misunderstood as overstatements. This was by no means our intention and we apologize for the unnecessary inconvenience. The phenotype of FKBP35 knock-out parasites clearly centers on failing ribosomes and protein synthesis, which in our opinion, provides an important leap towards understanding the role of this drug target in P. falciparum biology. It is however correct that, at this point, we can only make evidence-based hypotheses about direct interaction partners and we will emphasize this more clearly in a revised version of the manuscript. In order to prevent misinterpretation of our work, and as detailed in the point-by-point responses to the reviewer comments, we propose changing the manuscript title to “Genetic validation of PfFKBP35 as an antimalarial drug target”. To address the criticism regarding the effects of FK506, we will perform specific additional experiments. We are convinced that this new data set will resolve any remaining ambiguities and allows for a conclusive assessment of FK506 drug activity in P. falciparum.

Reviewer #1 (Public Review):

In this study, the authors investigate the biological function of the FK506-binding protein FKBP35 in the malaria-causing parasite Plasmodium falciparum. Like its homologs in other organisms, PfFKBP35 harbors peptidyl-prolyl isomerase (PPIase) and chaperoning activities, and has been considered a promising drug target due to its high affinity to the macrolide compound FK506. However, PfFKBP35 has not been validated as a drug target using reverse genetics, and the link between PfFKBP35-interacting drugs and their antimalarial activity remains elusive. The manuscript is structured in two parts addressing the biological function of PfFKBP35 and the antimalarial activity of FK506, respectively.

The first part combines conditional genome editing, proteomics and transcriptomics analysis to investigate the effects of FKBP35 depletion in P. falciparum. The work is very well performed and clearly described. The data provide definitive evidence that FKBP35 is essential for P. falciparum blood stage growth. Conditional knockout of PfFKBP35 leads to a delayed death phenotype, associated with defects in ribosome maturation as detected by quantitative proteomics and stalling of protein synthesis in the parasite. The authors propose that FKBP35 regulates ribosome homeostasis but an alternative explanation could be that changes in the ribosome proteome are downstream consequences of the abrogation of FKBP35 essential activities as chaperone and/or PPIase. It is unclear whether FKBP35 has a specific function in P. falciparum as compared to other organisms. The knockdown of PfFKBP35 has no phenotypic consequence, showing that very low amounts of FKBP35 are sufficient for parasite survival and growth. In the absence of quantification of the protein during the course of the experiments, it remains unclear whether the delayed death phenotype in the knockout is due to the delayed depletion of the protein or to a delayed consequence of early protein depletion. This limitation also impacts the interpretation of the drug assays.

We thank the Reviewer for the compliments regarding our experimental setup and the clarity of our manuscript. We agree that the link between FKBP35 knock-out and ribosome homeostasis is indirect and we now emphasize this more clearly in the revised manuscript. To prevent a general misinterpretation of our manuscript, we will adapt the title accordingly.

We would still like to reiterate that the phenotype of FKBP35 knock-out parasites is best described by their defects in maintaining functional ribosomes. It is for several reasons that we believe the links between FKBP35 and ribosome function are purely evidence driven: First, pre-ribosomal and nucleolar factors are the first proteins (in generation 1 schizonts) to be affected upon knock-out of fkbp35 (Figure 2A, Table S1). We realized that Figure 2A falls short in showing this observation, which is why will update the figure accordingly. Second, the dysregulation of ribosomal factors and the general stall in protein synthesis is dominating the phenotype of FKBP35 knock-out parasites in generation 2. We thus believe it is appropriate to say that knock-out cells are most likely killed in response to defective ribosome maintenance – which is a consequence of reduced FKBP35 levels. We are aware that our experiments (and possibly any other reverse genetics approach) cannot rule out that FKBP35 affects ribosomal factors indirectly. Clearly, more work is required to disentangle this question in more detail in the future.

We agree with the Reviewer that it is not possible to tell if the delayed death-like phenotype is due to a “delayed protein depletion”. We would however like to note that the DiCre/loxP approach allows for an immediate knock-out at the genome level and is thus as precise as possible. Further, in addition to the substantial depletion of FKBP35 in knock-out cells during the phenotypically silent generation, knocking out of fkbp35 at earlier time points (TPs 24-30 and 34-40 hpi in the preceding generation) resulted in the very same phenotype cycle (Figure 1). Here, parasite death was delayed substantially longer, i.e. more than one complete cycle. Together with the dysregulation of early ribosome maturation in generation 1, these findings point towards a delayed death phenotype. It is of course still possible to explain the delayed death-like phenotype by remnant activity of proteins synthetized prior to the genomic knock-out. We address this possibility and describe the two scenarios mentioned by the Reviewer in lines 141-144. Disentangling the two possibilities in future experiments will be difficult, not only with regards to FKBP35, but regarding “delayed death” phenotypes in general.

In the second part, the authors investigate the activity of FK506 on P. falciparum, and conclude that FK506 exerts its antimalarial effects independently of FKBP35. This conclusion is based on the observation that FK506 has the same activity on FKBP35 wild type and knock-out parasites, suggesting that FK506 activity is independent of FKBP35 levels, and on the fact that FK506 kills the parasite rapidly whereas inducible gene knockout results in delayed death phenotype. However, there are alternative explanations for these observations. As mentioned above, the delayed death phenotype could be due to delayed depletion of the protein upon induction of gene knockout. FK506 could have a similar activity on WT and mutant parasites when added before sufficient depletion of FKBP35 protein. In some experiments, the authors exposed KO parasites to FK506 later, presumably when the KO is effective, and obtained similar results. However, in these conditions, the death induced by the knockout could be a confounding factor when measuring the effects of the drug. Furthermore, the authors show that FK506 binds to FKBP35, and propose that the FK506-FKBP35 complex interferes with ribosome maturation, which would point towards a role of FKBP35 in FK506 action. In summary, the study does not provide sufficient evidence to rule out that FK506 exerts its effects via FKBP35.

Noteworthy, we were also very much surprised by data indicating that the antimalarial activity of FK506 is independent of FKBP35. It is for this reason that we conducted a comprehensive set of experiments to disprove our initial observations, but couldn`t find any evidence for an FKBP35-dependent mode of action of FK506:

We were not able to see altered FK506 sensitivity in (i) inducible knock-down parasites, (ii) inducible overexpression parasites and (iii) inducible knock-out parasites. Parasites with altered FKBP35 levels (as assessed by Western blot and quantitative proteomics at 36-42 hpi, respectively) were equally sensitive to FK506. Importantly, at no sub-lethal FK506 concentration did lower FKBP35 levels lead to an altered response of FKBP35KO compared to the wild-type control population. Furthermore, (iv) induction of the knock-out in the cycle preceding FK506 exposure also had no effect on parasite sensitivity. As mentioned by the Reviewer, we also exposed the parasites to FK506 at 30-36 hpi and (v) did not see any effect, even though we measured a 19-fold difference in FKBP35 protein levels between the parasite populations at 36-42 hpi. At this point, parasite death induced by the knock-out cannot be a confounding factor (as it was mentioned by the Reviewer), because the FKBP35 knock-out has no effect on parasite survival in generation 1 in the absence of FK506 (Figure 1F). This demonstrates that the observed effect is only due to drug-mediated killing and not due to the FKBP35 knock-out.

To account for a scenario in which the drop in FKBP35 levels only occurs after 36 hpi, we will perform an additional set of experiments, in which we induce the knock-out at 0-6 hpi and treat the parasites at 36-42 hpi (i.e. the time point at which the 19-fold difference in protein levels was measured by quantitative proteomics). This setup will allow determining whether or not the parasite killing activity of FK506 depends on FKBP35 levels.

So far, our experiments cannot support any scenario in which FK506 kills P. falciparum parasites via inhibiting the essential role of FKBP35 and we would therefore want to insist that this statement is based on highly solid evidence. In this context, it is important to note that our conclusion includes two scenarios: “This indicates that either the binding of FK506 does not interfere with the essential role of PfFKBP35, or that PfFKBP35 is inhibited only at high FK506 concentrations that also inhibit other essential factors.” While this phrase is already present in our initial submission, we will emphasize this point more clearly in the revised manuscript. We are convinced that this information is of high importance for ongoing and future drug development.

Reviewer #2 (Public Review):

The manuscript by Thomen et al. FKBP secures ribosome homeostasis in Plasmodium falciparum and focuses on the importance of PfKBP35 protein, its interaction with the FK506 compound, and the role of PfKBP35 in ribosome biogenesis. The authors showed the interaction of the PfKBP54 with FK506, but the part of the FK506 and PfKBP54 in ribosome biogenesis based on the data is unclear.

The introduction is plotted with two parallel stories about PfKBP35 and FK506, with ribosome biogenesis as the central question at the end. In its current form, the manuscript suffers from two stories that are not entirely interconnected, unfinished, and somewhat confusing. Both stories need additional experiments to make the manuscript(s) more complete. The results from PfFBP35 need more evidence for the proposed ribosome biogenesis pathway control. On the other hand, the results from the drug FK506 point to different targets with lower EC50, and other follow-up experiments are needed to substantiate the authors' claims.

The strengths of the manuscript are the figures and experimental design. The combination of omics methods is informative and gives an opportunity for follow-up experiments.

We thank the Reviewer for the evaluation of the manuscript. We apologize for the fact that the Reviewer found the manuscript to be inaccessible. We will use the comments as an incentive to restructure the manuscript and do our best to clarify the presentation, interpretation and conclusion of the presented data in the revised version. We believe that the FKBP35 data are strongly interlinked with the findings on FK506. We will emphasize these links more clearly and are convinced that the complementary nature of the datasets are a particular strength of the presented work.

Reviewer #3 (Public Review):

The study by Thommen et al. sought to identify the native role of the Plasmodium falciparum FKBP35 protein, which has been identified as a potential drug target due to the antiplasmodial activity of the immunosuppressant FK506. This compound has multiple binding proteins in many organisms; however, only one FKBP exists in P. falciparum (FKBP35). Using genetically-modified parasites and mass spectrometry-based cellular thermal shift assays (CETSA), the authors suggest that this protein is in involved in ribosome homeostasis and that the antiplasmodial activity of FK506 is separate from its activity on the FKBP35 protein. The authors first created a conditional knockdown using the destruction domain/shield system, which demonstrated no change in asexual blood stage parasites. A conditional knockout was then generated using the DiCre system. FKBP35KO parasites survived the first generation but died in the second generation. The authors called this "a delayed death phenotype", although it was not secondary to drug treatment, so this may be a misnomer. This slow death was unrelated to apicoplast dysfunction, as demonstrated by lack of alterations in sensitivity to apicoplast inhibitors. Quantitative proteomics on the FKBP35KO vs FKBP35WT parasites demonstrated enrichment of proteins involved in pre-ribosome development and the nucleolus. Interestingly, the KO parasites were not more susceptible to cycloheximide, a translation inhibitor, in the first generation (G1), suggesting that mature ribosomes still exist at this point. The SunSET technique, which incorporates puromycin into nascent peptide chains, also showed that in G1 the FKBP35KO parasites were still able to synthesize proteins. But in the second generation (G2), there was a significant decrease in protein synthesis. Transcriptomics were also performed at multiple time points. The effects of knockout of FKBP35 were transcriptionally silent in G1, and the parasites then slowed their cell cycles as compared to the FKBP35WT parasites.

The authors next sought to evaluate whether killing by FK506 was dependent upon the inhibition of PfKBP35. Interestingly, both FKBP35KO and FKBP35WT parasites were equally susceptible to FK506. This suggested that the antiplasmodial activity of FK506 was related to activity targeting essential functions in the parasite separate from binding to FKBP35. To identify these potential targets, the authors used MS-CETSA on lysates to test for thermal stabilization of proteins after exposure to drug, which suggests drug-protein interactions. As expected, FK506 bound FKBP35 at low nM concentrations. However, given that the parasite IC50 of this compound is in the uM range, the authors searched for proteins stabilized at these concentrations as putative secondary targets. Using live cell MS-CETSA, FK506 bound FKBP35 at low nM concentrations; however, in these experiments over 50 ribosomal proteins were stabilized by the drug at higher concentrations. Of note, there was also an increase in soluble ribosomal factors in the absence of denaturing conditions. The authors suggested that the drug itself led to these smaller factors disengaging from a larger ribosomal complex, leading to an increase in soluble factors. Ultimately, the authors conclude that the native function of FKBP35 is involved in ribosome homeostasis and that the antiplasmodial activity of FK506 is not related to the binding of FKBP35, but instead results from inhibition of essential functions of secondary targets.

Strengths:

This study has many strengths. It addresses an important gap in parasite biology and drug development, by addressing the native role of the potential antiplasmodial drug target FKBP35 and whether the compound FK506 works through inhibition of that putative target. The knockout data provide compelling evidence that the KBP35 protein is essential for asexual parasite growth after one growth cycle. Analysis of the FKBP35KO line also provides evidence that the effects of FK506 are likely not solely due to inhibition of that protein, but instead must have secondary targets whose function is essential. These data are important in the field of drug development as they may guide development away from structure-based FK506 analogs that bind more specifically to the FKBP35 protein.

Weaknesses:

There are also a few notable weaknesses in the evidence that call into question the conclusion in the article title that FKBP35 is definitely involved in ribosomal homeostasis. While the proteomics supports alterations in ribosome biogenesis factors, it is unclear whether this is a direct role of the loss of the FKBP35 protein or is more related to non-specific downstream effects of knocking down the protein. The CETSA data clearly demonstrate that FK506 binds PfKB35 at low nM concentrations, which is different than the IC50 noted in the parasite; however, the evidence that the proteins stabilized by uM concentrations of drug are actual targets is not completely convincing. Especially, given the high uM amounts of drug required to stabilize these proteins. This section of the manuscript would benefit from validation of a least one or two of the putative candidates noted in the text. In the live cell CETSA, it is noted that >50 ribosomal components are stabilized in drug treated but not lysate controls. Similarly, the authors suggest that the -soluble fraction of ribosomal components increases in drug-exposed parasites even at 37{degree sign}C and suggests that this is likely from smaller ribosomal proteins disengaging from larger ribosomal complexes. While the evidence is convincing that this protein may play a role in ribosome homeostasis in some capacity, it is not sure that the title of the paper "FKBP secures ribosome homeostasis" holds true given the lack of mechanistic data. A minor weakness, but one that should nonetheless be addressed, is the use of the term "delayed death phenotype" with regards to the knockout parasite killing. This term is most frequently used in a very specific setting of apicoplast drugs that inhibit apicoplast ribosomes, so the term is misleading. It is also possible that the parasites are able to go through a normal cycle because of the kinetics of the knockout and that the time needed for protein clearance in the parasite to a level that is lethal.

Overall, the authors set out to identify the native role of FKB35 in the P. falciparum parasites and to identify whether this is, in fact, the target of FK506. The data clearly demonstrate that FKBP35 is essential for parasite growth and provide evidence that alterations in its levels have proteomic but not transcriptional changes. However, the conclusion that FKBP35 actually stabilizes ribosomal complexes remains intermediate. The data are also very compelling that FK506 has secondary targets in the parasite aside from FKBP35; however, the high uM concentrations of the drug needed to attain results and the lack of biological validation of the CETSA hits makes it difficult to know whether any of these are actually the target of the compound or instead are nonspecific downstream consequences of treatment.

We appreciate the detailed and valuable suggestions to improve the manuscript. We agree that CETSA could only identify potential targets of FK506 in the micromolar range, while FK506 showed a high affinity for FKBP35, consistent with earlier reports (2). We would however like to point out that FK506 kills P. falciparum at exactly these relatively high concentrations and not at those presumed from the high affinity interactions between FK506 and FKBP35. The relatively high FK506 concentration required to stabilize potential off target proteins is therefore not a concerning observation, but rather corroborates our conclusion that FK506 fails to inhibit the essential function of FKBP35 at concentrations that leave off targets unaffected. As mentioned in response to Reviewer 1, we will describe and discuss these data more clearly in the revised manuscript. We thank the Reviewer for pointing out the potential issues regarding the use of the term “delayed death phenotype”. We now refer to the FKBP35 phenotype as “delayed death-like” in the revised manuscript.

We believe that follow-up work on specific FK506 CETSA hits is out of scope of the current and already quite complex manuscript.

As mentioned in the response to Reviewer 1, we realize that the short title of the manuscript can be regarded as an overstatement. Again, this was clearly not our intention and we apologize that the Reviewers had to indicate this issue. While we believe that the message of the title holds true (see response to Reviewer 1), we recognize the misconception that might arise from it, which is why we propose the new title: “Genetic validation of PfFKBP35 as an antimalarial drug target”.

Reviewer #1 (Recommendations For The Authors):

1. Documentation of FKBP35 protein levels over time in knockout, knockdown and overexpressing parasites is missing here. Since the knockdown of PfFKBP35 has no phenotypic consequence, very low amounts of FKBP35 are probably sufficient for parasite survival and growth. In the absence of quantification of the protein during the course of the experiments, it remains unclear whether the delayed death phenotype in the knockout is due to the delayed depletion of the protein or to a delayed consequence of early protein depletion. This limitation also impacts the interpretation of the drug assays. In particular, the delayed death phenotype could simply reflect delayed protein depletion, contrasting with the immediate inhibition of FKBP35 by FK506. The quantification by mass spectrometry does indicate substantial depletion but provides no information on the kinetics and levels. What is 19 fold compared to the knockdown condition? Also, expression of FKBP35 in overexpressing parasites should be compared side by side with the iKD (in the presence of Shield).

We agree with the Reviewer that low FKBP35 levels are likely sufficient for parasite survival. This is addressed in the manuscript (lines 141-143). Assessing protein levels in the transgenic parasites side by side in time course experiments would be interesting. However, our conclusions are independent of the outcome of such experiments because the relative difference in FKBP35 levels resulting from conditional expression systems did not change the parasites` susceptibility to FK506. We believe that comparing isogenic populations is much more informative than comparing independent cell lines with each other.

1. The authors claim that FK506 fails at inhibiting the essential function of PfFKBP35 (line 103), however this is not directly supported by data. FK506 kills the parasite and so inhibits essential functions. The data indicate that FK506 antimalarial activity does not seem to be influenced by FKBP35 levels, which would support the authors claim. However, as mentioned above, it is important to better define experimentally FKBP35 expression levels. Also, in experiments where FK506 is added late after rapamycin treatment, the authors need to clarify how they could distinguish drug killing and death due to the knockout.

In the experiment described by the Reviewer, the FKBP35 knock-out was induced in young ring stages (0-6 hpi) and FK506 was added at 30-36 hpi, we measured the parasite survival from G1 to G2 (see figure 5A). In the absence of FK506, the FKBP35 knock-out has no effect on parasite survival (Figure 1), demonstrating that the observed effect is only due to drug killing and not due to the KO.

To address the concerns regarding delayed depletion of FKBP35, we have performed an additional set of experiments. This data corroborates that the effect of FK506 is independent of FKBP35 levels. We discuss this topic in more detail in the Public Review. In brief, the additional experiment included exposing knock-out parasites (KO induced 0-6 hpi) with FK506 at 36-42 hpi, i.e. at a time point when FKBP35 protein levels are reduced by more than 90% (19-fold difference compared to the control parasites based on quadruplicate quantitative mass spectrometry data). However, despite the clear difference, the IC50 of FK506 remained the same as determined before (see new figure 4F).

1. Since FK506 is known to inhibit FKBP35 PPIase activity, it could be relevant to compare the effects of FK506 versus KO on ribosomes and translation. This could provide further evidence supporting a FKBP35-independent activity of FK506.

We agree with the Reviewer that this would be very informative. However, it would be difficult to disentangle indirect downstream processes on translation caused by both the FK506 drug treatment and the FKBP35 knock-out in a cellular assay. Establishing a biochemical in vitro assay to study the role of PPIase activity in translation is out of scope of this manuscript.

Minor points

-The title is rather vague, which reflects the fact that the function of PfFKBP35 is not precisely defined in the study.

We thank the Reviewer for this assessment, which is in agreement with Reviewer 3. Based on these concerns, and in order to prevent misinterpretation of our manuscript, we propose changing the title to “Genetic validation of PfFKBP35 as an antimalarial drug target” (see public response above).

-The transcriptomics data (Fig 3) provide little information on the function of FKBP35 and could be included as supplemental material. On the contrary, data in FigS5 convey important information and should be moved to the main figures.

We believe that the transcriptomics data are important to characterize the effect of limiting FKBP35 levels in G2, as they show that, unlike certain homologs of other organisms (3), FKBP35 has no role in transcriptional control and its knock-out does not have any downstream consequences on the transcriptional level (except for the death-related stall in cell cycle progression). We would therefore would like to keep this dataset represented in the main figures. The updated Figure 4F now includes more information about the effect of adding FK506 at different time points, which was only addressed in Figure S5 in the previous version of the manuscript. We believe that the key message of Figure S5 is now covered in Figure 4.

-Line 30: "action" rather than "role"

We corrected this.

Reviewer #2 (Recommendations For The Authors):

I have no comments on data, code, or other issues.

General comments:

The introduction is plotted with two parallel stories about PfKBP35 and FK506, with ribosome biogenesis as the central question at the end. In its current form, the manuscript suffers from two stories that are not entirely interconnected, unfinished, and somewhat confusing. I recommend focusing only on one story - either characterizing PfBP35 and its role in Plasmodium falciparum biology - future investigation of PfBP35 control of cellular processes or focusing on the actual targets of the FK506 drug (identified in figure 4). Both stories need additional experiments to make the manuscript(s) more complete and ready for publication. The results from PfFBP35 need more evidence for the proposed ribosome biogenesis pathway control. On the other hand, the results from the drug FK506 point to different targets with lower EC50, and other follow-up experiments are needed to substantiate the authors' claims.

The strengths of the manuscript are the figures and experimental design. The combination of omics methods is informative and gives an opportunity for follow-up experiments.

Detailed points and suggestions for authors:

Line 99

There is no such thing as "protein translation"; it is mRNA translation or protein synthesis, which needs to be updated throughout the manuscript.

We thank Reviewer 2 for pointing out this error that we now corrected.

Line 174

The statement needs a reference(s).

We added an appropriate review reference.

Lines 229-235

While transcriptomics and proteomics data can argue that FKBP35 maybe acts at the post-transcriptional level, its function, as well as presented data, could point to post-translational mechanisms as well, cell cycle checkpoint misregulation, and multiple other pathways that control cell size, cell proliferation, translation, and ribosome biogenesis. More solid and direct evidence on ribosome biogenesis (rRNA processing, polysome profiles, or similar experiments) would be needed to show the function of FKBP35 in this cellular process.

We have changed the term “post-transcriptional processes” to “transcription-independent processes”. As detailed in the Public Review, we agree with the Reviewer and lowered our statements regarding the function of FKBP35 throughout the manuscript.

Lines 237-313

The authors showed again the interaction of PfFKBP35 with the FK506 drug, but the phenotype differs from the protein deletion. Moreover, EC50s for multiple other proteins (i.e., PF3D7_1138700 or PF3D7_1325900, among others) are lower than for PfFKBP35 but are never further tested. This would be necessary to characterize FK506 drug targets, and it would be a different study.

We believe that characterizing putative targets of FK506 is out of the scope of this already complex study and should be addressed, as suggested by the Reviewer, in a future and independent efforts.

Lines 293 - 301

The point of lower EC50 for PfFKBP35 and FK506 in in vitro cell lysate experiment compared to in vivo IC50 data is not surprising, given that drug delivery is not an issue in a lysate experiment. It is unclear why the authors pick some proteins and not others for further characterization of FK506 binding. There is no explanation for this selection. They did not follow up on the best targets of FK406 drugs from Fig 4 (above comment).

As mentioned above, validation of FK506 targets is out of scope of this study.

Lines 313 -352

An alternative scenario for the FK506 drug data in CETSA experiments is that they bind directly to ribosomes interacting with rRNA, as many macrolides do. One should note that these are not ribosomal factors (line 334) but ribosomal proteins mentioned in Fig.4 F, mainly associated with large ribosomal subunit.

We agree with Reviewer 2 that FK506 could bind indirectly to ribosomal proteins. This scenario is already described in the initial version of the manuscript (see lines 285-287: “Of note, these ribosomal proteins were stabilized at virtually identical FK506 concentrations (Figs. 4D, and S7), indicating that the drug – directly or indirectly – interacts with ribosomal complexes.”).

We thank the Reviewer for pointing out that we are indeed talking about “ribosomal proteins” rather than “ribosomal factors”. We now corrected this.

Reviewer #3 (Recommendations For The Authors):

Please see Public review for suggestions about experimental validation of the link to ribosome homeostasis.

We would like to thank Reviewer 3 for the detailed suggestions.

References

1. Kennedy K, Cobbold SA, Hanssen E, Birnbaum J, Spillman NJ, McHugh E, et al. Delayed death in the malaria parasite Plasmodium falciparum is caused by disruption of prenylation-dependent intracellular trafficking. PLoS Biol. 2019;17(7):e3000376.

2. Kotaka M, Ye H, Alag R, Hu G, Bozdech Z, Preiser PR, et al. Crystal structure of the FK506 binding domain of Plasmodium falciparum FKBP35 in complex with FK506. Biochemistry. 2008;47(22):5951-61.

3. Kasahara K, Nakayama R, Shiwa Y, Kanesaki Y, Ishige T, Yoshikawa H, et al. Fpr1, a primary target of rapamycin, functions as a transcription factor for ribosomal protein genes cooperatively with Hmo1 in Saccharomyces cerevisiae. PLoS Genet. 2020;16(6):e1008865.

Version 2

Abstract

Abstract

Plasmodium falciparum accounts for the majority of over 600’000 malaria-associated deaths annually. Parasites resistant to nearly all antimalarials have emerged and the need for drugs with alternative modes of action is thus undoubted. The FK506-binding protein PfFKBP35 has gained attention as a promising drug target due to its high affinity to the macrolide compound FK506 (tacrolimus). Whilst there is considerable interest in targeting PfFKBP35 with small molecules, a genetic validation of this factor as a drug target is missing and its function in parasite biology remains elusive. Here, we show that limiting PfFKBP35 levels are lethal to P. falciparum and result in a delayed death-like phenotype that is characterized by defective ribosome homeostasis and stalled protein synthesis. Our data furthermore suggest that FK506, unlike the action of this drug in model organisms, exerts its anti-proliferative activity in a PfFKBP35-independent manner and, using cellular thermal shift assays, we identify putative FK506-targets beyond PfFKBP35. In addition to revealing first insights into the function of PfFKBP35, our results show that FKBP-binding drugs can adopt non-canonical modes of action – with major implications for the development of FK506-derived molecules active against Plasmodium parasites and other eukaryotic pathogens.

Introduction

Despite considerable progress in recent years, malaria remains one of the major global health threats (1). The apicomplexan parasite Plasmodium falciparum is responsible for the vast majority of the over 600’000 malaria deaths in 2020. Most of this burden is carried by infants and children under the age of five in sub-Saharan Africa. Malaria deaths increased by 12% compared to 2019, which can at least in part be explained by the COVID-19 related suspension of malaria control and treatment measures (1). In addition, the emergence of parasites resistant to most antimalarials, including artemisinin-based combination therapies, endangers current and future disease elimination efforts and underscores the need for developing drugs with alternative modes of action (2).

FK506 (tacrolimus), a well-characterized immunosuppressant, shows considerable activity against asexual blood stage parasites (3). P. falciparum encodes a single FK506-binding protein (FKBP) dubbed PfFKBP35 (PF3D7_1247400), which is considered to be a promising drug target (4, 5). FKBPs belong to the immunophilin family and are conserved throughout the eukaryotic kingdom (6, 7). In contrast to the malaria parasite, most other eukaryotes encode several FKBPs that, besides showing high affinity for the FK506 drug, exert peptidyl-prolyl isomerase (PPIase) activity. The PPIase moiety catalyzes cis-trans isomerization of proline residues – an event thought to be rate-limiting for the folding of many proteins (8). Independent of their PPIase activity, FKBPs act as chaperones to prevent protein aggregation under stress conditions (9, 10). Furthermore, some FKBP variants play a role in cellular signaling or gene regulation (11, 12). Human FKBP12 for instance is best-known for its ability to form a complex with the immunosuppressant drugs FK506 and rapamycin. These complexes inhibit the phosphatase activity of calcineurin and the kinase activity of mTOR (mechanistic target of rapamycin), respectively (13). Amongst many other effects, this inhibition results in decreased T-cell activation. Of note, the regulatory role of HsFKBP12 depends on the presence of rapamycin or FK506. In absence of drugs, this immunophilin neither interacts with mTOR nor with calcineurin and does hence not affect activity of these key components of eukaryotic cells. Due to the conserved nature of FKBPs, rapamycin and FK506 have also gained attention as potential antimicrobial drugs (3, 7).

Similar to its homologs in model organisms, PfFKBP35 harbors PPIase and chaperoning activity in vitro (5). Despite sharing key features with FKBPs of other eukaryotes, its role in P. falciparum remains elusive. PfFKBP35 is believed to interact with heat shock proteins (HSPs) and co-immunoprecipitates with cytoskeletal factors of P. falciparum (14). While PfFKBP35 is inhibiting the phosphatase activity of recombinant calcineurin in a FK506-independent manner (15-17), it does not co-localize with calcineurin in vivo (16). The protein is expressed throughout the 48-hour intra-erythrocytic development of P. falciparum and is found within the cytosol of ring stage parasites. In the older trophozoite and the replicating schizont stages, the protein was reported to also localize to the nucleus (16). While a transposon-based mutagenesis screen suggests that PfFKBP35 is essential for intra-erythrocytic replication (18), neither its cellular function nor its essentiality was confirmed experimentally in live parasites. The unknown role in parasite biology notwithstanding, PfFKBP35 is considered to be a viable drug target and, given the fact that P. vivax and P. knowlesi encode FKBP homologs that exhibit comparable PPIase activities (19, 20), future FKBP-targeted therapies may be effective across different Plasmodium species.

Crystal structures of PfFKBP35 in complex with FK506 and rapamycin revealed high affinity interactions with the FK506-binding domain of the protein (15, 21, 22). Consistent with these observations, in vitro enzyme activity assays showed that FK506 inhibits the PPIase activity of recombinant PfFKBP35 in a dose-dependent manner (5). Previous research aimed at exploiting structural differences between human and parasite-encoded FKBPs to circumvent the immunosuppressive activity of drug-bound HsFKBP12 (23-26). These efforts include investigation of non-covalent FKBP inhibitors such as adamantyl derivatives, macrocycles, the small molecule D44 (23-25, 27, 28) and the synthetic ligand for FKBP (SLF) as well as derivatives thereof designed to covalently bind PfFKBP35 (29). While D44 shows promising antimalarial activity, it fails to alter thermostability of the FK506-binding domains of PfFKBP35 and HsFKBP12 in vitro (29), suggesting that this compound does not directly interact with PfFKBP35. Up until now, PfFKBP35 has not been validated as a drug target using reverse genetics, and the link between PfFKBP35-interacting drugs and their antimalarial activity remains elusive (3).

Here, we generated inducible PfFKBP35 knock-out, knock-down as well as over-expression cell lines and demonstrate that PfFKBP35 is essential for asexual replication of blood forms. We show that the knock-out of PfFKBP35 results in reduced levels of ribosomal proteins and stalled protein synthesis. However, these effects are not reflected in the parasite transcriptome, indicating a vital role at the post-transcriptional level. While the mechanistic link between PfFKBP35 and ribosome homeostasis remains elusive, our data underpins the potential of PfFKBP35 as a drug target and, using cellular thermal shift assays (CETSA), we corroborate the high affinity binding of FK506 to PfFKBP35. Importantly, however, our data does not support scenarios in which FK506 inhibits the essential function of PfFKBP35, suggesting that it exerts its antimalarial activity in a PfFKBP35-independent manner, which has major implications for the development of future FK506-based antimalarials and/or PfFKBP35- inhibiting drugs.

Results

PfFKBP35 is essential for asexual replication

In a first step, using CRISPR/Cas9-mediated gene editing, we engineered a P. falciparum cell line allowing for a conditional knock-down of PfFKBP35 using the DD/Shield system (30, 31) (Figs. 1A and S1A-B). Despite the efficient depletion of PfFKBP35 under knock-down conditions, parasites did not show apparent growth defects (Figs. S1C-D). Immunofluorescence assays (IFA) showed that PfFKBP35 localizes to distinct foci within the parasite nucleus (Fig. 1B).

PfFKBP35 depletion causes delayed parasite death.

(A) Schematic of the N-terminally modified fkbp35 locus in NF54/iKD-FKBP35 parasites. dd, destabilization domain. gfp, green fluorescent protein. (B) Localization of PfFKBP35 in NF54/iKD-FKBP35 parasites at the ring, trophozoite and schizont stage assessed by immunofluorescence. DNA was stained with DAPI. Representative images are shown. Scale bar: 5 µm. DIC, differential interference contrast.(C) Schematic of the modified fkbp35 locus in NF54/iKO-FKBP35 parasites before and after loxP recombination. gfp, green fluorescent protein. (D) Flow cytometry plot showing GFP expression in G1 schizonts of FKBP35KO and FKBP35WT. DNA was stained with DRAQ5. (E) Time course showing the development of NF54/iKO-FKBP35 parasites. Upper panel: Schematic of the experimental setup. Middle panel: DNA content of the populations assessed by flow cytometry based on the SYBR Green intensity. Lower panel: Giemsa-stained blood smears. The knockout was induced at 0-6 hpi. Representative data of three biological replicates are shown. (F) Multiplication rates of FKBP35KO and FKBP35WT parasites from generation 1 to generation 2 (G1-G2) and G2-G3. The knockout was induced at 0-6 hpi. Multiplication rates were determined using flow cytometry. Data points represent the multiplication rates from G1-G2 and G2-G3. n=3, error bars represent the standard error of the mean. (G) Schematic showing time point of growth arrest (red arrow) following PfFKBP35 knock-out at different hpi (blue arrow). R, ring stage; T, trophozoite stage; S, schizont stage.

Next, we generated a cell line allowing for the conditional knock-out of the endogenous coding sequence using a dimerizable Cre recombinase (DiCre)-based system (32). Specifically, we introduced two loxP sites – one within a synthetic intron (33) downstream of the start codon and one after the stop codon of the fkbp35 coding region – and added a green fluorescent protein (gfp) sequence downstream of this expression cassette (Figs. 1C, S2A). As a result, these NF54/iKO-FKBP35 parasites express tag-free, wild-type (WT) PfFKBP35 protein under control conditions (FKBP35WT). Upon DiCre activation, the loxP sites are recombined and the fkbp35 locus is excised from the parasite genome (Figs. 1C, S2B). Under these knock-out conditions (FKBP35KO), parasites express cytosolic GFP under control of the endogenous fkbp35 promoter. This setup allows monitoring knock-out efficiency at the single cell level and revealed that DiCre successfully deleted fkbp35 in the vast majority of parasites (Fig. 1D).

In a first attempt, we induced the fkbp35 knock-out in synchronous ring stage NF54/iKO-FKBP35 parasites at 0-6 hours post erythrocyte invasion (hpi), i.e. prior to the expected peak transcription of fkbp35 (34). These FKBP35KO parasites showed a delayed death-like phenotype. In the first generation (G1), parasites completed intra-erythrocytic development without noticeable effects (Fig. 1E) and entered the next generation (G2) with a multiplication rate similar to that of the FKBP35WT control population (Fig. 1F). In G2, however, FKBP35KO parasites arrested at the late trophozoite/early schizont stage (approximately 30-36 hpi) with many cells failing to enter DNA replication in the S-phase (Fig. 1E). To investigate this phenomenon in more detail, we induced the fkbp35 knock-out at three different time points during the IDC. Similar to DiCre activation at 0-6 hpi, knocking out fkbp35 at 12-18 hpi resulted in parasite death at the trophozoite/schizont stage of the subsequent generation. However, when induced at 24-30 hpi or 34-40 hpi in G1, parasites did not only complete the current IDC but also successfully passed through G2 before they eventually arrested in G3 (Fig. 1G). Considering that low PfFKBP35 levels are sufficient to maintain the essential function of the protein, as observed with the NF54/iKD-FKBP35 parasites (see Fig. S1C-D), it is conceivable that proteins expressed between 0 and 24hpi, i.e. prior to deleting fkbp35 from the genome, allow parasites to enter G3.

The delayed death of FKBP35KO is reminiscent of a phenotype observed in apicoplast-deficient parasites (35, 36). We therefore tested if PfFKBP35 is involved in apicoplast biology, i.e. the synthesis and/or export of isoprenoid precursors, but did not find evidence for an association with this organelle (Fig. S2C). Furthermore, given the reported interaction of PfFKBP35 with heat shock proteins (14, 16), we also tested the parasitès response to heat stress, but could not detect altered heat-susceptibility between FKBP35KO and FKBP35WT cells (Fig. S2D).

PfFKBP35 is crucial for ribosome homeostasis and protein synthesis

To further elucidate the function of PfFKBP35, we induced the fkbp35 knock-out in early ring stages (0-6 hpi) and compared the proteomes of FKBP35KO and FKBP35WT parasites at the schizont stage of G1 (36-42 hpi) and at the trophozoite stage of the following generation (G2, 24-30 hpi) using a quantitative proteomics approach.

The knock-out of PfFKBP35 causes accumulation of pre-ribosome components in G1

Consistent with the efficient deletion of fkbp35 on the genomic level, FKBP35KO schizonts of G1 showed a 19-fold (18.9 +/-5.7 s.d.) reduction of FKBP35 and GFP levels were increased by 84-fold (83.7 +/-38.5 s.d.) when compared to the control population (Fig. 2A, Table S1). Besides PfFKBP35 and GFP, 50 parasite proteins were significantly deregulated in FKBP35KO schizonts and 34 of them found at elevated levels compared to the FKBP35WT control conditions. A gene ontology analysis using PANTHER (37) indicated enrichment of the “pre-ribosome” (fold enrichment = 30.4, p-value = 1.46E-5, 4/34 proteins) and “nucleolus” (fold enrichment = 6.77, p-value = 8.85E-4, 5/34 proteins) terms among these factors. While the pre-ribosome is a large subunit precursor of mature ribosomes, the nucleolus represents the primary site for the biogenesis of ribosomal subunits (38). In contrast to factors being more abundant in knock-out parasites, none of the 16 factors with lower abundance in FKBP35KO cells were associated with ribosome-related processes (Table S1). Despite being phenotypically silent in G1, the accumulation of pre-ribosome and nucleolar components hints at perturbed ribosome maturation in these cells. Specifically, it is conceivable that the observed buildup of ribosome intermediates results from reduced anabolic activity, i.e. the stalling of nucleolar pre-ribosome assembly (39).

PfFKBP35 is crucial for ribosome homeostasis and protein synthesis.

(A) Left panel: Proteome analysis of schizonts (36-42 hpi) in G1. The volcano plot shows change in protein abundance in FKBP35KO compared to FKBP35WT. Ribosomal proteins are highlighted in blue, proteins associated with the GO terms “pre-ribosome” or “nucleolus” in red. n=3. Right panel: Relative difference in PfFKBP35 and GFP abundance expressed as signal intensity measured by mass spectrometry. n=3, error bars represent the standard error of the mean. (B) Dose-response effect of the translation inhibitor cycloheximide on FKBP35KO and FKBP35WT parasites. The knockout was induced at 0-6 hpi and parasites were subsequently exposed to cycloheximide for 48 h. Parasite survival was quantified by flow cytometry. n=3, error bars represent the standard error of the mean. (C) Proteome analysis of trophozoites (24- 30 hpi) in G2. The volcano plot shows change in protein abundance in FKBP35KO compared to FKBP35WT. Ribosomal proteins are highlighted in blue. n=3. (D) GO enrichment analysis of significantly deregulated proteins in FKBP35KO trophozoites in G2. GO enrichment analysis was performed using the PANTHER “Overrepresentation Test” with the test type “Fisher’s Exact” and the correction method “False Discovery Rate”. (E) Assessment of translation activity using SUnSET. FKBP35KO and FKBP35WT trophozoites of G2 were incubated with puromycin for 1 hour. Incorporated puromycin was detected using an α-puromycin antibody. α-GAPDH served as loading control. n=3, a representative Western blot is shown.

Notwithstanding these effects, FKBP35KO cells were equally sensitive to the translation inhibitor cycloheximide compared to their FKBP35WT counterparts (Fig. 2B), indicating that ribosome function is not compromised at this point in time (40). This is consistent with normal cell cycle progression in G1 and unaltered multiplication rates of FKBP35KO parasites, as described above (see Figs. 1E-F), and shows that the prominent depletion of PfFKBP35 does not have a major impact on parasite viability in G1.

PfFKBP35 is essential for maintaining normal levels of functional ribosomes in G2

In the following generation (G2, 24-30 hpi), the proteome of FKBP35KO trophozoites showed substantial differences. Despite the absence of morphological changes at this point in time, 216 proteins were significantly deregulated (Fig. 2C). Among these, 50 proteins were found at higher abundance compared to FKBP35WT and their gene ontology terms point towards an effect on processes involved in energy metabolism (Fig. 2D).

The majority of deregulated factors (166 of 216) in G2, however, were less abundant in FKBP35KO compared to the FKBP35WT control cells. Very prominently, 20 of the top 50 hit molecules are ribosomal proteins that, on average, show a two-fold reduction (2.0 +/- 0.4 s.d.) in FKBP35KO parasites (Table S1). It is therefore not surprising that PANTHER identified the knock-out of PfFKBP35 to mainly affect ribosome-related processes, such as “cytoplasmic translation”, “ribosome assembly”, and “ribosomal large subunit biogenesis” (Fig. 2D) (37).

To assess if the reduced levels of ribosomal proteins result in lower protein synthesis rates, we employed the “surface sensing of translation” (SUnSET) technique. SUnSET is based on incorporation of the tyrosyl-tRNA analogue puromycin into nascent peptide chains and thereby allows monitoring newly synthesized proteins (41). As expected, we found that FKBP35KO trophozoites in G2 (24-30 hpi) showed reduced translation rates compared to the FKBP35WT control population (Fig. 2E), while protein synthesis was not affected under control conditions (rapamycin-induced DiCre activation in the NF54/DiCre parental line (Fig. S2E)).

Taken together, these results indicate that, under knock-out conditions, PfFKBP35 levels become limiting for ribosome maturation late during schizogony of G1, without having immediate effects on the steady-state levels of mature ribosomes. The knock-out of PfFKBP35 is thus phenotypically silent in G1. In absence of a continuous supply of mature ribosomes under FKBP35KO conditions, however, FKBP35KO parasites fail at maintaining translation at a level that would support normal parasite development in the next generation (G2).

The knock-out of PfFKBP35 is transcriptionally silent

To examine if the altered proteome of FKBP35KO parasites is the result of PfFKBP35- controlled transcription, we compared the transcriptional profiles of FKBP35WT and FKBP35KO in a time course RNA-Sequencing (RNA-Seq) experiment spanning two IDCs. Specifically, we induced the knock-out of FKBP35 in synchronous parasites at 0-6 hpi and collected RNA from parasite populations at three and four subsequent time points (TPs) in G1 and G2, respectively (Figs. 3A, S3A). With the exception of TP7, these time points cover a time frame during which FKBP35KO cells cannot be distinguished from their FKBP35-expressing counterparts on a morphological level (compare Fig. 1E).

The knockout of PfFKBP35 remains silent at the transcriptomic level.

(A) Setup of the RNA-Seq experiment. RNA was collected at seven indicated time points (TP). (B) Volcano plots show differential gene expression between FKBP35KO and FKBP35WT parasites. Genes are color-coded according to the fold change and fdr (false discovery rate), n=3. (C) Mapping of FKBP35KO and FKBP35WT transcriptomes to a high-resolution reference (42). Spearman rank coefficients (ρ) show correlation between the sampled transcriptomes and the reference data set (42). The time points with the highest correlation coefficient are indicated.

As expected, the transcriptome of FKBP35KO showed a prominent reduction of steady-state fkbp35 transcripts already in TP1 (11.3-fold +/-2.0 s.d.) compared to the FKBP35WT control population (Fig. S3B, Table S2). During peak transcription at 30-36 hpi (TP2), this difference increased to 240-fold (238.4 +/-20.0 s.d.) and, consistent with the efficient recombination at the fkbp35 locus, these cells started transcribing gfp. Notwithstanding the prominent depletion of fkbp35 transcripts, the transcriptome of FKBP35KO is only marginally affected up to and including 24-30 hpi in G2 (TP6). By contrast, at 30-36 hpi (TP7) the transcription of knock-out parasites shows major changes (Fig. 3B). Considering the appearance of morphological changes in FKBP35KO at this time point, effects on the transcriptome are unlikely to represent specific activities of PfFKBP35 in gene regulation. Indeed, mapping the transcriptome of individual time points to a high resolution reference (42), revealed a significant slow-down of FKBP35KO compared to FKBP35WT parasites at TP7 (Figs. 3C, S3C). This indicates aborted cell cycle progression in FKBP35KO parasites during a phase that is characterized by a rapid increase in biomass – possibly due to a translation and/or size-dependent cell cycle checkpoint (43). At earlier time points (TP1-TP6), the transcriptomes of FKBP35KO and FKBP35WT are highly comparable and the differences observed on the level of individual genes clearly fail at explaining the proteomic changes observed in FKBP35KO parasites at 24-30 hpi in G2 (TP6), reiterating that the essential activity of PfFKBP35 is linked to transcription-independent processes.

Probing FK506 interactions in Cellular Thermal Shift Assays (CETSA)

PfFKBP35 emerged as a promising antimalarial drug target (4) and considerable efforts were made to define interactions of this immunophilin with small molecules in the last decade (24-26, 44). Several studies conclusively demonstrated that FK506 – the most well-known ligand of mammalian FKBPs (45) – binds to PfFKBP35 (15, 21, 22) and kills asexual blood stage parasites with a half-maximal inhibitory concentration (IC50) in the low micromolar range at 1.9 µM (3).

To investigate interactions of FK506 with P. falciparum proteins, we used the cellular thermal shift assay followed by mass spectrometry (MS-CETSA). This approach exploits the fact that, once drug-protein complexes are formed, thermostability of the target protein is altered (46). We applied the so-called isothermal dose-response variant of CETSA using FK506 (47). Here, target proteins are identified based on their dose-dependent stabilization by drugs under thermal challenge (different temperatures) when compared to a non-denaturing condition (37 °C), using the change in the area under the curve (ΔAUC) and the dose-response curve goodness of fit (R2) as metrics (Fig. 4A) (47).

CETSA identifies FK506 interaction partners.

(A) CETSA results. The relative shift in protein abundance (area under the curve of heat-challenged samples normalized against the non-denaturing control, ΔAUC) as a function of R2 (goodness of fit as a measure for the dose-response effect). The dashed lines indicate an arbitrary cutoff of 3 times the median absolute deviation (MAD) of each dataset and R2=0.8. Three different temperatures were used for protein denaturation (indicated by colors). (B) Dose-dependent stabilization of P. falciparum and human FKBPs detected using lysate and live cell CETSA. Bullets represent the abundance of non-denatured proteins in response to increasing FK506 concentrations under thermal challenge relative to the DMSO control. Colors indicate the temperatures used for heat challenge; the black curve represents the non-denaturing (37°C) control. (C) Proteins with the most pronounced stability shift under thermal challenge in the lysate CETSA experiment are listed. The EC50 is the concentration at which the half-maximal effect of the FK506-induced thermal stabilization is reached. The mutagenesis index score determined by Zhang et al. (18) is a measure to predict if a protein is essential for asexual growth: 0=essential, 1=non-essential. (D) Fitted live cell CETSA curves of 10 representative ribosomal proteins at 37°C and 55°C. Ribosomal proteins show a FK506 dose-dependent increase in soluble protein fraction. (E) Venn diagram showing the overlap between ribosomal proteins deregulated in the PfFKBP35 knock-out cell line (less abundant in G2 FKBPKO trophozoites determined by proteomics) and the ribosomal proteins stabilized by FK506 in live cell CETSA. The number of factors identified in either or both assays is indicated. (F) Dose-response effect of FK506 on FKBP35KO and FKBP35WT parasites. The knockout was induced at 0-6 hpi and parasites were exposed to FK506 at 4-10 hpi and 36-42 hpi for 48h. n=3, error bars represent the standard error of the mean. (G) Dose-response effect of FK506 on NF54/iKD-FKBP35 and NF54/iOE-FKBP35 parasites in presence and absence of Shield-1 (Shld). Parasites were split at 0-6 hpi and subsequently treated with Shld or the vehicle control EtOH. Data points represent the mean of three and two independent biological replicates for the NF54/iOE-FKBP35 and NF54/iKD-FKBP35 cell line, respectively. Error bars represent the standard error of the mean.

CETSA identifies targets of FK506

In a first step, we performed CETSA in combination with drug-exposed protein extracts. Since proteins lose their physiological context during the process of protein extraction, this variant of the CETSA approach is designed to identify direct protein-drug interactions (46, 47). Not surprisingly, we found PfFKBP35 to be stabilized by FK506 in a dose-dependent manner when challenged at 60 °C with a half maximal effective concentration (EC50) of 152.0 nM (Fig. 4B). The fact that FK506 exhibits its stabilizing activity already in the low nanomolar concentration range indicates high affinity interactions between the protein-drug pair.

Besides PfFKBP35, FK506 stabilizes a number of other proteins in a dose-dependent manner (Figs. 4C and S6). A few of these factors show pronounced sigmoidal stabilization profiles and are predicted to be essential (Fig. 4C) (18). Amongst others, N6-methyl transferase MT-A70 (PF3D7_0729500), histone deacetylase HDA2 (PF3D7_1008000), pre-mRNA-splicing factor CEF1 (PF3D7_1033600), FeS cluster assembly protein SufD (PF3D7_1103400) and the H/ACA ribonucleoprotein complex subunit 1 (PF3D7_1309500) showed most pronounced stabilization profiles (Fig. S6 and Table S3). In contrast to PfFKBP35, most of these likely essential candidates are stabilized at FK506 concentrations close to the drug`s IC50 (3). It can therefore not be excluded that the binding of FK506 to at least one of these factors is contributes to the antimalarial activity of the drug.

The PfFKBP35/FK506 pair interferes with ribosomal complexes

Given the high-affinity interaction of PfFKBP35 with FK506 and its association with ribosomal proteins, we set out to describe the native context of PfFKBP35 using the “live cell variant” of CETSA. In contrast to focusing exclusively on direct protein-drug interactions, live cell CETSA allows capturing drug-induced shifts in protein stability beyond those of direct binding partners (47) and amongst others allows identifying factors present in the same protein complex as the target protein. To do so, CETSA takes advantage of the fact that factors within complexes have similar denaturation temperatures due to their co-aggregation upon heat-induced denaturation (48). As a consequence, this CETSA variant generally identifies more factors compared to the protein extract-based CETSA approach (Fig. 4A).

As expected, and confirming the results above, live cell CETSA shows that FK506 alters the thermal stability of PfFKBP35 already at very low drug concentrations when challenged at 60 °C with an EC50 of 286.7 nM (Fig. 4B). Interestingly, and in contrast to the protein lysate-based approach, live cell CETSA revealed that more than 50 ribosomal proteins are stabilized in presence of FK506 when challenged at 55 °C, with a median EC50 of 45.9 nM (25th percentile = 34.2, 75th percentile = 58.9). Of note, these ribosomal proteins were stabilized at virtually identical FK506 concentrations (Figs. 4D, and S7), indicating that the drug – directly or indirectly – interacts with ribosomal complexes. This is in further agreement with the fact that these ribosomal factors were not identified by protein lysate-based CETSA (compare Figs. 4A, S6).

Worth mentioning, we observed a drug dose-dependent increase in soluble protein abundance of ribosomal proteins also under non-denaturing conditions, i.e. at 37 °C, with a median EC50 of 1.94 µM (25th percentile = 1.85, 75th percentile = 2.1) (Figs. 4D, S7). Generally, due to the absence of a denaturing heat challenge, such a drug-dependent enrichment in soluble protein levels does not indicate binding-induced thermal stabilization of proteins (47). Instead, this can be the result of increased protein synthesis, altered protein conformation or, alternatively, occur in response to factors disengaging from larger protein complexes. The fact that FK506 solubilizes a high number of ribosomal proteins within a very narrow concentration window also at 37 °C, implies that these factors are liberated from the same ribosomal complex or from complexes that interact with each other and shows that FK506 negatively affects stability of ribosomal structures in live cells. However, these effects are limited to high, presumably saturating FK506 concentrations. Importantly, the vast majority of ribosomal proteins identified in live cell CETSA are shared with those de-regulated in PfFKBP35 knock out cells (Fig. 4E), pointing towards a shared localization of FK506 and its PfFKBP35 target in live parasites.

Together with the above-described importance of PfFKBP35 in ribosome homeostasis, these results strongly suggest that FK506 is recruited to the activity site of PfFKBP35 – most probably the nucleolar maturation sites of ribosomes – where it interacts with ribosomal complexes.

FKBP-targeting drugs fail at inhibiting the essential function of PfFKBP35

Considering the proposed direct interaction between FK506 and FKBP35 in P. falciparum, we expected that FKBP35KO parasites show altered sensitivity to the drug. Consistent with published data, FK506 killed FKBP35WT cells with an IC50 of 1942 nM (+/-193 nM s.d.) when administered at 4-10 hpi (Fig. 4F). Surprisingly however, and despite the prominent depletion of the protein under knock-out conditions, FKBP35KO and FKBP35WT parasites were equally sensitive to FK506 (IC50 = 1741 nM +/- 187 s.d.) (Fig. 4F). Similarly, neither the efficient knock-down of PfFKBP35 in NF54/iKD-FKBP35 parasites (compare Fig. S1C), nor its over-expression (OE) in NF54/iOE-FKBP35 cells (Fig. S4A-C), resulted in the expected sensitivity change to FK506 treatment (Fig. 4G). In addition to this apparent independence of the drug to varying PfFKBP35 levels, and in agreement with earlier studies (3, 24, 49), FK506 unfolds its antimalarial activity within the same cycle after treatment. Standing in stark contrast to the delayed death-like phenotype of FKBP35KO parasites (compare Fig. 1E-G), this immediate activity on parasite survival substantiates the notion that FK506 must target essential parasite factors other than PfFKBP35.

To rule out that the timing of drug exposure is masking expected links between FK506 and PfFKBP35, we tested whether administering FK506 at 30-36 hpi and 36-42 hpi (instead of 4- 10 hpi) affects parasite survival in a PfFKBP35-dependent manner. However, we did not observe a change compared to FKBP35WT parasites (Fig. S5A), ruling out the possibility that FKBP35KO parasites were killed by FK506 before PfFKBP35 levels were different in the two populations. Further, to account for a scenario in which the activity of FK506 on PfFKBP35 is masked by a temporal delay between drug exposure and parasite killing, we induced the knock-out of PfFKBP35 at 34-40 hpi (instead of 0-6 hpi) and allowed FKBP35KO parasites to proceed to G2, before exposing them to different FK506 concentrations. Despite these efforts, FKBP35WT and FKBP35KO parasites were equally sensitive to FK506 (Fig. S5B, S5C) and the calcineurin inhibitor cyclosporin A (Fig. S5D-E).

In addition to FK506, we evaluated the PfFKBP35-specific inhibitor D44 (N-(2-Ethylphenyl)- 2-(3h-Imidazo[4,5-B]pyridin-2-Ylsulfanyl)acetamide) but, surprisingly, we failed to reproduce the previously reported dose-response effect (24). We tested D44 from two suppliers using NF54/iKO-FKBP35, NF54 wild type and K1 parasite strains using different assay setups, including the gold standard method that is based on the incorporation of tritium labeled hypoxanthine into DNA of replicating cells (50). Still, we could not detect any activity of D44 on asexual replication of blood stage parasites (Figs. S5F-G).

Taken together, these results are in clear conflict with the perception that FK506 and D44 target the essential activity of PfFKBP35.

Discussion

Members of the FKBP family are involved in regulating diverse cellular processes in eukaryotic as well as in prokaryotic cells (8, 51). In contrast to most other eukaryotes, the human malaria parasite P. falciparum encodes a single FKBP only (16). The enzymatic activity of PfFKBP35, as well as the interactions of the protein with FKBP-targeting drugs, was subject to considerable drug-focused research in vitro. Besides demonstrating the drug`s affinity to PfFKBP35 by co-crystallization, these efforts uncovered that FK506 inhibits enzymatic activity of the protein`s PPIase domain (5). Together with the anti-malarial activity of FK506, these data offered compelling evidence for an essential but yet uncharacterized role of PfFKBP35 during blood stage development of P. falciparum.

Here, we set out to investigate the role of PfFKBP35 in parasite biology. Using an inducible knock-out approach, we show that PfFKBP35 is indeed essential for intra-erythrocytic development of asexually replicating P. falciparum parasites and CETSA experiments confirmed high affinity binding of FK506 to its PfFKBP35 target. While the knock-out of PfFKBP35 does not markedly affect transcription, FKBP35KO cells show reduced levels of ribosomal proteins and impaired protein synthesis on the global scale. Surprisingly however, these effects, as well as the consequent death during the late trophozoite/early schizont stage, only manifest in the cycle following knock-out induction without affecting cell cycle progression in the first generation (G1) or multiplication rates from G1 to G2. In contrast to the delayed death-like phenotype of FKBP35KO cells, FK506 unfolds its parasiticidal activity without a lag phase, i.e. within the IDC of treatment (24), and the dose-response relationship between FK506 and parasite death is irresponsive to altered PfFKBP35 levels. Taken together, our data strongly indicate that the parasite-killing activity of FK506 is PfFKBP35-independent.

Knocking out PfFKBP35 early during intra-erythrocytic development has only minor effects on the parasite proteome in G1. Being relatively subtle, these changes may still point towards an important function of PfFKBP35: Factors associated with the nucleolus – the site of ribosome biogenesis – are found at higher abundance in knock-out cells. In the next cycle, FKBP35KO parasites show a dramatic decrease of factors associated with ribosome-and translation-related processes and, likely a consequence thereof, globally reduced protein synthesis during trophozoite development (G2, 24-30 hpi). Importantly, at this point in time, FKBP35KO cells still show ordinary morphology and their transcriptional activity remains largely unaffected. The observed changes in the abundance of ribosomal factors must therefore result from transcription-independent activity of PfFKBP35. While ribosomal components are present at higher levels in G1, they are drastically reduced in FKBP35KO cells in G2. These opposing effects likely result from PfFKBP35 activity in a shared cellular process. Specifically, it is conceivable that low PfFKBP35 levels inhibit ribosome maturation in G1, explaining the accumulation of (pre-) ribosomal factors. While the levels of functional ribosomes are sufficient to support continued translation in G1, ribosomes likely become limiting in the next generation (G2), which eventually results in stalled translation activity and concomitant parasite death at the onset of schizogony.

In contrast to the significant effects on the protein levels in G2, the transcription of FKBP35KO remains largely unaffected by the loss of PfFKBP35 and becomes apparent only during late trophozoite/early schizont development in G2 (30-36 hpi); i.e. at a time during which cell cycle progression is slowed down in FKBP35KO parasites. In fact, at this stage the transcriptional signature of FKBP35KO parasites is dominated by the slowed progression through intra-erythrocytic development, rather than resulting from specific gene regulatory events. While this stalling is consistent with the observed effects on ribosome homeostasis and protein synthesis, it demonstrates that, in contrast to certain FKBP family members in other organisms, PfFKBP35 does not have apparent roles in transcriptional control.

Regarding the well-characterized and fast-acting parasiticidal activity of FK506 on P. falciparum, it was somewhat surprising to observe a delayed death-like phenotype for PfFKBP35 knock-out parasites (24). We hence set out to investigate unifying explanations. The delayed death of FKBP35KO parasites is most easily be explained by residual activity of PfFKBP35 proteins synthesized prior to inducing the genomic knock-out of fkbp35. In such a scenario, low PfFKBP35 in FKBP35KO would be sufficient to allow completion of the IDC, before eventually dropping below a level that is required for supporting parasite survival in the next cycle. Despite intense efforts, we failed at confirming this hypothesis and our data strongly favors an alternative scenario in which FK506 kills parasites in a PfFKBP35-idenpendent manner. Aside from the fact that FK506 exerts its parasiticidal activity within the same IDC, neither the conditional knock-out, knock-down nor the overexpression of PfFKBP35 altered the parasite’s sensitivity to FK506; and this irrespective of the profoundly deregulated PfFKBP35 levels in the respective mutant lines (Fig. 4F). For instance, administering FK506 36 hours post induction of the knock-out in early ring stages (36-42 hpi) – i.e. at a time point where FKBP35KO cells are only left with about 5% of the regular PfFKBP35 levels (19-fold reduction compared to FKBP35WT parasites, Fig. 2A) – does not affect the dose-response of parasites to the drug. Further, our attempts to demonstrate that the timing of drug exposure is masking the expected links between FK506 and its PfFKBP35 drug target failed (compare Figs. 4A and S5A-C). Our data is thus incompatible with the notion that the parasite-killing activity of FK506 results from inhibiting the essential function of PfFKBP35.

Since binding of FK506 to PfFKBP35 alone cannot explain the drug’s antimalarial activity, we aimed identifying additional FK506 targets. CETSA performed on protein lysates identified a number of parasite factors binding to FK506. Compared to the PfFKBP35/FK506 pair, however, these interactions are of considerably lower affinity. The adenosine-methyltransferase MT-A70 (PF3D7_0729500) and the pre-mRNA-splicing factor CEF1 (PF3D7_1033600), for instance, were stabilized by FK506 with EC50 concentrations of 2.0 and 3.1 µM, respectively. While this affinity is relatively poor compared to the strong interaction formed between PfFKBP35 and FK506, one should be cautious designating the binding of FK506 to these proteins as irrelevant “off-target” effects. In fact, their inhibition may hypothetically mask effects of FK506 on PfFKBP35 – a possibility that shall briefly be discussed in the following section.

Assuming that FK506 is inhibiting the essential function of PfFKBP35, one would expect parasites to show a delayed death-like phenotype similar to that observed in FKBP35KO parasites. The presence and inhibition of other essential targets may, however, mask such an effect. Consistent with this possibility, FK506 kills P. falciparum parasites at 1.9 µM (+/-0.2 µM s.d.), i.e. at concentrations similar to those required for stabilizing MT-A70 and CEF1 under thermal challenge in CETSA. However, numerous attempts to prove that FK506 indeed inhibits essential PfFKBP35 functions failed: Neither knocking out PfFKBP35 at different time points nor adding FK506 to different stages of FKBP35KO parasites could provide evidence for a PfFKBP35-dependent effect. This indicates that either the binding of FK506 does not interfere with the essential role of PfFKBP35, or that PfFKBP35 is inhibited only at high FK506 concentrations that also inhibit other essential factors. In this context it is worth mentioning that parasites remain unaffected by a highly efficient knock-down of endogenous PfFKBP35 (observed with NF54/iKD-FKBP35; compare Figs. S1C-D), which demonstrates that very low protein levels are sufficient to maintain the essential function of PfFKBP35. It is thus conceivable that high FK506 concentrations are required to fully inhibit PfFKBP35 – a view that is further supported by data obtained with the live cell CETSA approach discussed hereafter.

Besides identifying direct binding of drugs to their target, live cell CETSA allows capturing indirect effects on factors that act in the same pathway or in the same protein complex as the drug-bound target (47). While this approach confirmed the high affinity binding of FK506 to PfFKBP35, it also revealed that more than 50 ribosomal proteins share a very prominent CETSA signature: First, they are thermally stabilized by FK506 within a narrow concentration window of the drug. Second, higher levels of FK506 increase the solubility of these ribosomal components also at 37 ℃. These results indicate that the FK506/PfFKBP35 pair interacts with ribosomal complexes and that FK506 is able to induce the dissociation of these ribosomal structures under physiological conditions. Importantly, and in contrast to the binding of FK506 to PfFKBP35, this effect on ribosomes is only observed at high concentrations of the drug. Together, these data likely explain the discrepancy observed between the high binding affinity of FK506 for PfFKBP35 and the low or even absent effect of the drug on PfFKBP35 function in blood stage parasites.

In line with our data, previous biochemical approaches revealed potential links between PfFKBP35 and ribosomes. Using co-immunoprecipitation, Leneghan and colleagues revealed that PfFKBP35 is interacting with 480 proteins, of which 31 are ribosomal factors (14). Indeed, FKBP family members are essentially involved in controlling ribosomes in other systems. The human FKBP variant HsFKBP25, for instance, is recruited to the pre-ribosome where its chaperone activity appears to be required for the assembly of the ribosomal large subunit (52). While the FKBP variant of yeast, Fpr4, interacts with the ribosome biogenesis factor Nop53 (53), the Escherichia coli trigger factor (TF) is part of the ribosomal complex and crucial for the folding of nascent proteins (54, 55). Considering the marked reduction of ribosomal proteins in FKBP35KO parasites, it is tempting to speculate that PfFKBP35 is chaperoning ribosome biogenesis similar to Fpr4 and FKBP25 in Saccharomyces cerevisiae and humans, respectively.

The fact that PfFKBP35 localizes to distinct nuclear foci and its depletion causes dysregulation of ribosomes points towards a role within the nucleolus – the nuclear compartment driving ribosome biogenesis (38). Using IFAs, we found that PfFKBP35 does not co-localize with the nucleolar marker fibrillarin (56) (Fig. S8). However, its localization appears to be intertwined with that of this rRNA methyltransferase. Specifically, PfFKBP35 is mostly found in close vicinity to fibrillarin-rich regions. Nucleoli in eukaryotic cells generally consist of two or, in higher eukaryotes, three sub-compartments. Of those, the granular component (GC) represents the site of pre-ribosome assembly and encompasses fibrillarin-enriched central regions of nucleoli (57). The spatial correlation observed between PfFKBP35 and fibrillarin may thus indicate localization to different sub-nucleolar compartments, with PfFKBP35 occupying the GC area of the nucleolus.

In summary, we demonstrate that limiting PfFKBP35 levels are lethal to P. falciparum and result in a delayed death-like phenotype that is characterized by perturbed ribosome homeostasis and defective protein synthesis, which is likely linked to the reduced biogenesis of functional ribosomes in the nucleolus. We further found no evidence that FKBP-binding drugs, including FK506, exert their parasiticidal activity in a PfFKBP35-dependent manner, urging strong caution for the future development of FKBP-targeting antimalarials, especially when based on FK506 and structural derivatives thereof. In addition to revealing first insights into the essential function of PfFKBP35 in ribosome homeostasis of P. falciparum, we are convinced that the presented data offer valuable information for target-based efforts in malaria drug discovery.

Methods

Parasite culture

P. falciparum culture and synchronization was performed as described (58, 59). Parasites were cultured in AB+ or B+ human red blood cells (Blood Donation Center, Zürich, Switzerland) at a hematocrit of 5% in culture medium containing 10.44 g/L RPMI-1640, 25 mM HEPES, 100 μM hypoxanthine, 24 mM sodium bicarbonate, 0.5% AlbuMAX II (Gibco #11021-037), 0.1 g/L neomycin, and 2 mM choline chloride (Sigma #C7527). To achieve stabilization of DD-tagged proteins in the cell lines NF54/iKD-FKBP35 and NF54/iOE-FKBP35, parasites were cultured in presence of 625 nM Shield-1 (30, 31). To induce the DiCre-mediated recombination of loxP sites in NF54/iKO-FKBP35 parasites, cultures were split and treated for 4 h with 100 nM rapamycin or the corresponding volume DMSO, which served as vehicle control, giving rise to FKBP35KO and FKBP35WT populations, respectively (32). Cultures were gassed with 3% O2, 4% CO2 and 93% N2 and incubated in an airtight incubation chamber at 37°C.

Cloning of transfection constructs

CRISPR/Cas9-based gene editing of the NF54 parasites was performed using a two-plasmid approach as previously described (60, 61). This system is based on co-transfection of a plasmid that contains the expression cassettes for the Cas9 enzyme, the single guide RNA (sgRNA) and the blasticidin deaminase (BSD) resistance marker (pBF-gC), and a pD-derived donor plasmid that contains the template for the homology-directed repair of the Cas9-induced DNA double strand break (Fig. S2A) (61).

The plasmids pBF-gC_FKBP-3’, pBF-gC_FKBP-5’ and pBF-gC_P230p targeting the 3’ or 5’ end of fkbp35 (PF3D7_1247400) or the p230p (PF3D7_0208900) locus (62), respectively, were generated by ligation of two annealed oligonucleotides (gRNA_top and gRNA_bottom) into the BsaI-digested pBF-gC backbone (61) using T4 DNA ligase (New England Biolabs).

The donor plasmid pD_FKBP35-iKO was generated by assembling six DNA fragments in a Gibson reaction (63) using (i) the plasmid backbone amplified by polymerase chain reaction (PCR) from pUC19 (primers PCRa_F/PCRa_R) (61), (ii/iii) the 5’ and 3’ homology regions amplified from genomic DNA (primers PCRb_F/PCRb_R and PCRc_F/PCRc_R, respectively), (iv) the green fluorescent protein (gfp) sequence amplified from pHcamGDV1- GFP-DD (primers PCRd_F/PCRd_R) (61), (v) the HRPII terminator amplified from genomic DNA (primers PCRe_F/PCRe_R), and (vi) the fkbp35 sequence containing an artificial loxPint (33) amplified from a P. falciparum codon-optimized synthetic sequence (primers PCRf_F/PCRf_R) ordered from Genscript.

The donor plasmid pD_FKBP35-iKD was generated by assembling five DNA fragments in a Gibson reaction (63) using (i) the plasmid backbone amplified from pUC19 (primers PCRa_F/PCRa_R) (61), (ii/iii) the 5’ and 3’ homology regions amplified from genomic DNA (primers PCRg_F/PCRg_R and PCRh_F/PCRh_R, respectively), (iv) the gfp sequence amplified from pHcamGDV1-GFP-DD (primers PCRi_F/PCRi_R) (61), and (v) the destabilization domain (dd) sequence amplified from pHcamGDV1-GFP-DD (primers PCRj_F/PCRj_R) (61).

The donor plasmid pD_FKBP35-iOE was generated by assembling three DNA fragments in a Gibson reaction (63) using (i) the NheI/PstI-digested plasmid pkiwi003 (64), (ii) the fkbp35 coding sequence amplified from genomic DNA (primers PCRk_F/PCRk_R), and (iii) the gfp-dd sequence amplified from pHcamGDV1-GFP-DD (primers PCRl_F/PCRl_R) (61).

Oligonucleotides are listed in Table S4.

Transfection and selection of gene-edited parasites

P. falciparum transfection using the CRISPR/Cas9 two-plasmid approach was performed as described previously (61). Briefly, 50 μg of each plasmid (pBF-gC_FKBP-3’ and pD_FKBP35- iKO, pBF-gC_FKBP-5’ and pD_FKBP35-iKD, pBF-gC_P230p and pD_FKBP35-iOE) were co-transfected into an NF54::DiCre line (65), in which ap2-g was tagged with the fluorophore mScarlet as described previously (66). Transgenic parasites were selected with 2.5 μg/mL blasticidin-S-hydrochloride, which was added 24 h after transfection for 10 days. Transgenic populations were usually obtained 2-3 weeks after transfection and correct editing of the modified loci was confirmed by PCR on gDNA (Figs. S1B, S2B and S4B). Primer sequences used for these PCRs are listed in Table S4. Clonal parasite lines were obtained by limiting dilution cloning (67).

DNA content analysis

NF54/iKO-FKBP35 parasites were fixed in 4% formaldehyde/0.0075% glutaraldehyde for 30 min at room temperature and washed 3 times in PBS and stored at 4°C. Samples were collected from three independent biological replicates. RNA was digested by incubation with 0.1 mg/mL RNAse A in PBS containing 0.1% Triton-X100 for 15 min at room temperature. Nuclei were stained using 1X SYBR Green 1 DNA stain (Invitrogen S7563) for 20 min and washed three times in PBS. The fixation step quenched the GFP signal detected by the flow cytometer (Fig. S9). SYBR Green intensity was measured using a MACS Quant Analyzer 10 and analyzed using the FlowJo_v10.6.1 software.

Western blotting

Saponin lysis of 500 µL infected red blood cells at 3-5% parasitaemia was performed by incubation for 10 min on ice in 3 mL ice-cold 0.15% saponin in PBS. The parasite pellets were washed two times in ice-cold PBS and resuspended in Laemmli buffer (62.5 mM Tris base, 2% SDS, 10% glycerol and 5% 2-mercaptoethanol). Proteins were separated on 4-12% Bis-Tris gels (Novex, Qiagen) using MOPS running buffer (Novex, Qiagen). Then proteins were transferred to a nitrocellulose membrane (GE healthcare #106000169), which was blocked with 5% milk powder in PBS/0.1% Tween (PBS-T) for 1 h. The membrane was probed using the primary antibodies mAb mouse α-GFP (1:1’000, Roche Diagnostics #11814460001), mAb mouse α-PfGAPDH (1:20’000) (68), or mAb mouse α-puromycin (1:5’000, Sigma MABE343) diluted in blocking solution. After 2 h incubation, the membrane was washed 5 times in PBS-T before it was incubated with the secondary antibody goat α-mouse IgG (H&L)-HRP (1:10’000, GE healthcare #NXA931). After 1 h incubation, the membrane was washed 5 times and the signal was detected using the chemiluminescent substrate SuperSignal West Pico Plus (Thermo Scientific, REF 34580) and the imaging system Vilber Fusion FX7 Edge 17.10 SN. Nitrocellulose membranes were stripped by incubating for 10 min in a stripping buffer (1 g/L SDS, 15 g/L glycine, 1% Tween-20, pH 2.2) and washed three times in PBS-T before blocking and re-probing.

Drug dose-response experiments

Dose-response relationships were assessed by exposing synchronous ring stage parasites at 0.5% parasitaemia and 1.25% hematocrit to 12 drug concentrations using a two-step serial dilution in 96-well plates (Corning Incorporated, 96-well cell culture plate, flat bottom, REF 3596) in three independent biological replicates. The plates were gassed and incubated for 48 h at 37 °C. FK506 (MedChemExpress HY-13756), D44 (ChemBridge 7934155 and ChemDiv 7286-2836) and cyclosporin A (Sigma 30024) stocks of 10 mM were prepared in DMSO and working solutions were prepared immediately before the experiment. To measure parasite survival using flow cytometry, cultures were stained in 1X SYBR Green 1 DNA stain (Invitrogen S7563) in a new 96-well plate (Corning Incorporated, 96-well cell culture plate, round bottom, REF 3788) and incubated in the dark for 20 min. The samples were washed once in PBS. Using a MACS Quant Analyzer 10, 50’000 events per condition were measured and analyzed using the FlowJo_v10.6.1 software. Infected RBCs were identified based on SYBR Green 1 DNA stain intensity. The gating strategy is shown in Fig. S9. To quantify parasite survival based on hypoxanthine incorporation into DNA, compounds were dissolved in DMSO (10 mM), diluted in hypoxanthine-free culture medium and titrated in duplicates over a 64-fold range (6 step twofold dilutions) in 96 well plates. 100 μl asexual parasite culture (prepared in hypoxanthine-free medium) were added to each well and mixed with the compound to obtain a final haematocrit of 1.25% and a final parasitemia of 0.3%. After incubation for 48 hrs, 0.25 μCi of [3H]-hypoxanthine was added per well and plates were incubated for an additional 24 hrs. Parasites were then harvested onto glass-fiber filters using a Microbeta FilterMate cell harvester (Perkin Elmer, Waltham, US) and radioactivity was counted using a MicroBeta2 liquid scintillation counter (Perkin Elmer, Waltham, US). The results were recorded and expressed as a percentage of the untreated controls. Curve fitting and IC50 calculations were performed using a non-linear, four-parameter regression model with variable slope (Graph Pad Prism, version 8.2.1).

Fluorescence microscopy

To localize PfFKBP35, NF54/iKD-FKBP35 parasites were fixed in 4% formaldehyde/0.0075% glutaraldehyde for 30 min at room temperature and washed three times in PBS. The fixed cells were permeabilized by incubation with 0.1% Triton-X100-containing PBS for 15 min. After two washing steps in PBS, the cells were incubated in a blocking/aldehyde quenching solution (3% BSA in PBS complemented with 50 mM ammonium chloride). The samples were incubated with the primary antibodies rabbit α-GFP (1:400, Abcam ab6556) and mAb mouse α-fibrillarin (1:100, Santa Cruz Biotechnology sc-166021) for 1 h in 3% BSA in PBS and washed 3 times in PBS. Subsequently, the samples were incubated with the secondary antibodies goat α-rabbit IgG (H&L) Alexa Fluor 488 (1:250, Invitrogen A11008) and goat α- mouse IgG (H&L) Alexa Fluor 594 (1:250, Invitrogen A11032), for 1 h and washed 3 times, before they were mixed with Vectashield containing DAPI (Vector laboratories, H-1200), and mounted on a microscopy slide. Slides were imaged using a Leica THUNDER 3D Assay imaging system.

Proteomics

Parasite cultures were split at 0-6 hpi in G1 and treated for 4 h with 100 nM rapamycin. Saponin lysis of paired FKBP35WT and FKBP35KO populations at 36-42 hpi in G1 and 24-30 hpi in G2 was performed as described above. Samples were collected in three independent biological replicates. The washed parasite pellet was snap frozen in liquid nitrogen. After thawing, parasites were resuspended in 25 µL of lysis buffer (5% SDS, 100 mM tetraethylammonium bromide (TEAB), 10 mM tris(2-carboxyethyl)phosphin (TCEP), pH 8.5) and sonicated (10 cycles, 30 sec on/off at 4 °C, Bioruptor, Diagnode). Lysates were subsequently reduced for 10 minutes at 95 °C. Samples were then cooled down to RT and 0.5 µL of 1M iodoacetamide was added to the samples. Cysteine residues were alkylated for 30 min at 25 °C in the dark. Digestion and peptide purification was performed using S-trap technology (Protifi) according to the manufacturer’s instructions. In brief, samples were acidified by addition of 2.5 µL of 12% phosphoric acid (1:10) and then 165 µL of S-trap buffer (90% methanol, 100 mM TEAB pH 7.1) was added to the samples (6:1). Samples were briefly vortexed and loaded onto S-trap micro spin-columns (Protifi) and centrifuged for 1 min at 4000 g. Flow-through was discarded and spin-columns were then washed 3 times with 150 µL of S-trap buffer (each time samples were centrifuged for 1 min at 4000 g and flow-through was removed). S-trap columns were then moved to the clean tubes and 20 µL of digestion buffer (50 mM TEAB pH 8.0) and trypsin (at 1:25 enzyme to protein ratio) were added to the samples. Digestion was allowed to proceed for 1h at 47 °C. Then, 40 µL of digestion buffer were added to the samples and the peptides were collected by centrifugation at 4000 g for 1 minute. To increase the recovery, S-trap columns were washed with 40 µL of 0.2% formic acid in water (400g, 1 min) and 35 µL of 0.2% formic acid in 50% acetonitrile. Eluted peptides were dried under vacuum and stored at - 20 °C until further analysis.

Peptides were resuspended in 0.1% aqueous formic acid and peptide concentration was adjusted to 0.25 µg/µL. 1 µL of each sample was subjected to LC-MS/MS analysis using an Orbitrap Elicpse Tribrid Mass Spectrometer fitted with an Ultimate 3000 nano system (both from Thermo Fisher Scientific) and a custom-made column heater set to 60 °C. Peptides were resolved using a RP-HPLC column (75 μm × 30 cm) packed in-house with C18 resin (ReproSil-Pur C18–AQ, 1.9 μm resin; Dr. Maisch GmbH) at a flow rate of 0.3 μL/min. The following gradient was used for peptide separation: from 2% buffer B to 12% B over 5 min, to 30% B over 40 min, to 50% B over 15 min, to 95% B over 2 min followed by 11 min at 95% B then back to 2% B. Buffer A was 0.1% formic acid in water and buffer B was 80% acetonitrile, 0.1% formic acid in water.

The mass spectrometer was operated in DDA mode with a cycle time of 3 seconds between master scans. Each master scan was acquired in the Orbitrap at a resolution of 240,000 FWHM (at 200 m/z) and a scan range from 375 to 1600 m/z followed by MS2 scans of the most intense precursors in the linear ion trap at “Rapid” scan rate with isolation width of the quadrupole set to 1.4 m/z. Maximum ion injection time was set to 50 ms (MS1) and 35 ms (MS2) with an AGC target set to 1e6 and 1e4, respectively. Only peptides with charge state 2-5 were included in the analysis. Monoisotopic precursor selection (MIPS) was set to Peptide, and the Intensity Threshold was set to 5e3. Peptides were fragmented by HCD (Higher-energy collisional dissociation) with collision energy set to 35%, and one microscan was acquired for each spectrum. The dynamic exclusion duration was set to 30 seconds.

The acquired raw-files were imported into the Progenesis QI software (v2.0, Nonlinear Dynamics Limited), which was used to extract peptide precursor ion intensities across all samples applying the default parameters. The generated mgf-file was searched using MASCOT against a Plasmodium falciparum (isolate 3D7) database (Uniprot, 11.2019) and 392 commonly observed contaminants using the following search criteria: full tryptic specificity was required (cleavage after lysine or arginine residues, unless followed by proline); 3 missed cleavages were allowed; carbamidomethylation (C) was set as fixed modification; oxidation (M) and acetyl (Protein N-term) were applied as variable modifications; mass tolerance of 10 ppm (precursor) and 0.6 Da (fragments). The database search results were filtered using the ion score to set the false discovery rate (FDR) to 1% on the peptide and protein level, respectively, based on the number of reverse protein sequence hits in the dataset. Quantitative analysis results from label-free quantification were processed using the SafeQuant R package v.2.3.2 (69) to obtain peptide relative abundances. This analysis included global data normalization by equalizing the total peak/reporter areas across all LC-MS runs, data imputation using the knn algorithm, summation of peak areas per protein and LC-MS/MS run, followed by calculation of peptide abundance ratios. Only isoform specific peptide ion signals were considered for quantification. To meet additional assumptions (normality and homoscedasticity) underlying the use of linear regression models and t-Tests, MS-intensity signals were transformed from the linear to the log-scale. The summarized peptide expression values were used for statistical testing of between condition differentially abundant peptides. Here, empirical Bayes moderated t-Tests were applied, as implemented in the R/Bioconductor limma package (70). The resulting per protein and condition comparison p-values were adjusted for multiple testing using the Benjamini-Hochberg method.

Gene ontology enrichment analysis was performed using the PANTHER Overrepresentation Test with the correction method “False Discovery Rate”, the test type “Fisher’s Exact”, and the annotation data set “PANTHER GO-Slim” (37).

IPP complementation

Parasite cultures were split at 0-6 hpi in G1 and treated for 4 h with 100 nM rapamycin. FKBP35KO and FKBP35WT populations were split again and supplemented either with 200 µM isopentenyl pyrophosphate (IPP) (Sigma I0503) or the corresponding volume H2O. The respective culture medium was replaced daily. Parasites treated with 156 ng/mL doxycycline (Sigma D9891) were used to confirm that IPP complementation is able to rescue apicoplast-deficient parasites. DNA content analysis was performed as described above.

Heat shock susceptibility testing

Parasites cultures were split at 0-6 hpi in G1 and treated for 4 h with 100 nM rapamycin. While control populations were always incubated at 37 °C, test populations were incubated at 40 °C for 6 h starting at 24-30 hpi. Parasitaemia was measured in G2 using flow cytometry as described above.

Surface sensing of translation (SUnSET)

Parasite cultures were split at 0-6 hpi in G1 and treated for 4 h with 100 nM rapamycin. At 24- 30 hpi in G2, the cultures were incubated with 1 µg/mL puromycin (Sigma P8833) for 1 h at 37 °C (41, 71). Saponin lysis and Western blotting were performed as described above.

Cellular thermal shift assay (CETSA)

The mass spectrometry-based isothermal dose-response cellular thermal shift assay (ITDR-MS-CETSA) was performed according to the protocol developed by Dziekan and colleagues (47). Briefly, live parasites (107 MACS-purified trophozoites per condition) were exposed to a FK506 (MedChemExpress HY-13756) concentration gradient (100 µM to 1.5 nM) and a DMSO control for 1 h. Subsequently, the parasites were washed in PBS and exposed to the denaturing temperatures 50 °C, 55 °C and 60 °C as well as to a non-denaturing temperature 37 °C for 3 min, before they were cooled down to 4 °C for 3 min. Afterwards, the cells were lysed by resuspending in a lysis buffer (50 mM HEPES pH 7.5, 5 mM β-glycerophosphate, 0.1 mM Na3VO4, 10 mM MgCl2, 2 mM TCEP, and EDTA-free protease inhibitor cocktail (Sigma)), three freeze/thaw cycles, and mechanical shearing using a syringe with a 31-gauge needle. Subsequently, insoluble proteins were pelleted by centrifugation (20 min at 4 °C and 20’000 g) and the soluble fraction was collected and subjected to protein digestion, reduction, alkylation, and TMT10 labelling followed by LC/MS analysis as described (47).

To prepare the protein lysate, saponin-lysed parasites were lysed in the afore-mentioned buffer by three freeze/thaw cycles and mechanical shearing. After centrifugation, soluble proteins were exposed to a FK506 concentration gradient (200 µM to 3 nM) and a DMSO control for 1 min before the thermal challenge was performed as described above. Afterwards, the soluble protein fraction was isolated and prepared for by LC/MS analysis as described (47).

Data analysis was performed using the Proteome Discoverer 2.1 software (Thermo Fisher Scientific) and the R package “mineCETSA” (v 1.1.1). Only proteins identified by at least three peptide spectrum matches (PSMs) were included in the analysis (47).

Transcriptomics

Sample preparation

The RNA of 500 µL iRBCs at 3-5% parasitaemia was isolated using 3 mL TRIzol reagent (Invitrogen) followed by phenol-chloroform phase separation. The aqueous phase was purified using the Direct-zol RNA MicroPrep kit (Zymo). Samples were collected from three biological replicates. Stranded RNA sequencing libraries were prepared using the Illumina TruSeq stranded mRNA library preparation kit (REF 20020594) according to the manufacturer’s protocol. The library was sequenced in 100 bp paired-end reads on an Illumina NextSeq 2000 sequencer using the Illumina NextSeq 2000 P3 reagents (REF 20040560), resulting in 5 million reads on average per sample.

RNA-Seq data analysis

Fastqc was run on all raw fastq files. Then, each file was aligned with hisat2 (version 2.0.5) (72) against the 3D7 reference genome (PlasmoDB v58) (73) in which the GFP and the recoded FBPK35 sequences were added at the end of chromosome 12. Read counts for each gene and for each sample were then counted using featureCount (gff file from PlasmoDB v58 modified to include the added GFP and FBPK35). TPM (Transcripts Per Kilobase Million) were then calculated from the raw count matrix.

Principal component analysis (PCA) was performed using the R (version 4.2.1) function prcomp using log transformed TPMs. PCA revealed that the rapamycin-treated replicate B of TP 3 (“3B.R”) was a clear outlier and is different from any other replicate at any time point (Fig. S3D). For this reason, it was excluded from further analysis. The PCA was done again without it and the other samples show the same pattern as the initial PCA (Fig. S3E).

DESeq2 (version 1.36) (74) was used for normalization and differential expression analysis. Paired differential expression tests where done for each time point using DESeq2 LRT (likelihood ratio test) using the full model ∼ Treatment + Replicate (Treatment is DMSO or Rapamycin and Replicate is the clone, added as the data is paired) and the reduced model ∼ Replicate. From this paired test and the average log fold-change calculated from the TPMs, normalized volcano plots were made using the R package EnhancedVolcano (version 1.14) (75).

Cell cycle analysis

Spearman correlation between normalized scaled read counts for each sample for all genes (as one vector) and each time point of the reference microarray time course published by Bozdech et al. (42) (3D7 smoothed, retrieved from PlasmoDB) were tested. Spearman’s coefficients rho for each time point of this RNA-Seq time course and each time point of the reference data set where plotted using ggcorrplot (version 0.1.3) (76). For each time point of this RNA-Seq dataset, the rho coefficients where then plotted against the time points of the reference data set and the FKBP35WT and FKBP35KO data were fitted separately to the linearized sinusoidal A ∗ using R lm function. The time point matching best to the FKBP35WT and FKBP35KO samples for each time point of the RNA-Seq time course was taken as the maximum of this fitted function. Both fitted functions were F-tested (using R function var.test using both fitted model as applied to the FKBP35KO data) to test whether the FKBP35WT and FKBP35KO data correlated with the reference data set (42) in a significantly different way, which would indicate a cell cycle shift if statistically significant. Only the last time point showed such a significant difference.

Data availability

The mass spectrometry proteomics data have been deposited to the ProteomeXchange Consortium repository with the dataset identifier PXD039018. The sequencing data are available in the Sequence Read Archive (SRA) via the accession number PRJNA914079.

Supporting information

eLife Assessment

This important study addresses both the native role of the Plasmodium falciparum protein PfFKBP35 and whether this protein is the target of FK506, an immunosuppressant with antiplasmodial activity. The demonstration of the essentiality of FKBP35 in parasite growth relies on compelling genetic evidence. However, it remains unclear whether FK506 exerts its antimalarial activity through an FKBP35-independent mechanism.

Reviewer #1 (Public Review):

In this study, the authors investigate the biological function of the FK506-binding protein FKBP35 in the malaria-causing parasite Plasmodium falciparum. Like its homologs in other organisms, PfFKBP35 harbors peptidyl-prolyl isomerase (PPIase) and chaperoning activities, and has been considered a promising drug target due to its high affinity to the macrolide compound FK506. However, PfFKBP35 has not been validated as a drug target using reverse genetics, and the link between PfFKBP35-interacting drugs and their antimalarial activity remains elusive. The manuscript is structured in two parts addressing the biological function of PfFKBP35 and the antimalarial activity of FK506, respectively.

The first part combines conditional genome editing, proteomics and transcriptomics analysis to investigate the effects of FKBP35 depletion in P. falciparum. The work is very well performed and clearly described. The data provide definitive evidence that FKBP35 is essential for P. falciparum blood stage growth. Conditional knockout of PfFKBP35 leads to a delayed death phenotype, associated with defects in ribosome maturation as detected by quantitative proteomics and stalling of protein synthesis in the parasite. The authors propose that FKBP35 regulates ribosome homeostasis but an alternative explanation could be that changes in the ribosome proteome are downstream consequences of the abrogation of FKBP35 essential activities as chaperone and/or PPIase. It is unclear whether FKBP35 has a specific function in P. falciparum as compared to other organisms. The knockdown of PfFKBP35 has no phenotypic consequence, showing that very low amounts of FKBP35 are sufficient for parasite survival and growth. In the absence of quantification of the protein during the course of the experiments, it remains unclear whether the delayed death phenotype in the knockout is due to the delayed depletion of the protein or to a delayed consequence of early protein depletion. This limitation also impacts the interpretation of the drug assays.

In the second part, the authors investigate the activity of FK506 on P. falciparum, and conclude that FK506 exerts its antimalarial effects independently of FKBP35. This conclusion is based on the observation that FK506 has the same activity on FKBP35 wild type and knock-out parasites, suggesting that FK506 activity is independent of FKBP35 levels, and on the fact that FK506 kills the parasite rapidly whereas inducible gene knockout results in delayed death phenotype. However, there are alternative explanations for these observations. As mentioned above, the delayed death phenotype could be due to delayed depletion of the protein upon induction of gene knockout. FK506 could have a similar activity on WT and mutant parasites when added before sufficient depletion of FKBP35 protein. In some experiments, the authors exposed KO parasites to FK506 later, presumably when the KO is effective, and obtained similar results. However, in these conditions, the death induced by the knockout could be a confounding factor when measuring the effects of the drug. Furthermore, the authors show that FK506 binds to FKBP35, and propose that the FK506-FKBP35 complex interferes with ribosome maturation, which would point towards a role of FKBP35 in FK506 action. In summary, the study does not provide sufficient evidence to rule out that FK506 exerts its effects via FKBP35.

Reviewer #2 (Public Review):

The introduction is plotted with two parallel stories about PfKBP35 and FK506, with ribosome biogenesis as the central question at the end. In its current form, the manuscript suffers from two stories that are not entirely interconnected, unfinished, and somewhat confusing. I recommend focusing only on one story - either characterizing PfBP35 and its role in Plasmodium falciparum biology - future investigation of PfBP35 control of cellular processes or focusing on the actual targets of the FK506 drug (identified in figure 4). Both stories need additional experiments to make the manuscript(s) more complete. The results from PfFBP35 need more evidence for the proposed ribosome biogenesis pathway control. On the other hand, the results from the drug FK506 point to different targets with lower EC50, and other follow-up experiments are needed to substantiate the authors' claims. The strengths of the manuscript are the figures and experimental design. The combination of omics methods is informative and gives an opportunity for follow-up experiments.

Reviewer #3 (Public Review):

The study by Thommen et al. sought to identify the native role of the Plasmodium falciparum FKBP35 protein, which has been identified as a potential drug target due to the antiplasmodial activity of the immunosuppressant FK506. This compound has multiple binding proteins in many organisms; however, only one FKBP exists in P. falciparum (FKBP35). Using genetically-modified parasites and mass spectrometry-based cellular thermal shift assays (CETSA), the authors suggest that this protein is in involved in ribosome homeostasis and that the antiplasmodial activity of FK506 is separate from its activity on the FKBP35 protein. The authors first created a conditional knockdown using the destruction domain/shield system, which demonstrated no change in asexual blood stage parasites. A conditional knockout was then generated using the DiCre system. FKBP35KO parasites survived the first generation but died in the second generation. The authors called this "a delayed death phenotype", although it was not secondary to drug treatment, so this may be a misnomer. This slow death was unrelated to apicoplast dysfunction, as demonstrated by lack of alterations in sensitivity to apicoplast inhibitors. Quantitative proteomics on the FKBP35KO vs FKBP35WT parasites demonstrated enrichment of proteins involved in pre-ribosome development and the nucleolus. Interestingly, the KO parasites were not more susceptible to cycloheximide, a translation inhibitor, in the first generation (G1), suggesting that mature ribosomes still exist at this point. The SunSET technique, which incorporates puromycin into nascent peptide chains, also showed that in G1 the FKBP35KO parasites were still able to synthesize proteins. But in the second generation (G2), there was a significant decrease in protein synthesis. Transcriptomics were also performed at multiple time points. The effects of knockout of FKBP35 were transcriptionally silent in G1, and the parasites then slowed their cell cycles as compared to the FKBP35WT parasites.

The authors next sought to evaluate whether killing by FK506 was dependent upon the inhibition of PfKBP35. Interestingly, both FKBP35KO and FKBP35WT parasites were equally susceptible to FK506. This suggested that the antiplasmodial activity of FK506 was related to activity targeting essential functions in the parasite separate from binding to FKBP35. To identify these potential targets, the authors used MS-CETSA on lysates to test for thermal stabilization of proteins after exposure to drug, which suggests drug-protein interactions. As expected, FK506 bound FKBP35 at low nM concentrations. However, given that the parasite IC50 of this compound is in the uM range, the authors searched for proteins stabilized at these concentrations as putative secondary targets. Using live cell MS-CETSA, FK506 bound FKBP35 at low nM concentrations; however, in these experiments over 50 ribosomal proteins were stabilized by the drug at higher concentrations. Of note, there was also an increase in soluble ribosomal factors in the absence of denaturing conditions. The authors suggested that the drug itself led to these smaller factors disengaging from a larger ribosomal complex, leading to an increase in soluble factors. Ultimately, the authors conclude that the native function of FKBP35 is involved in ribosome homeostasis and that the antiplasmodial activity of FK506 is not related to the binding of FKBP35, but instead results from inhibition of essential functions of secondary targets.

Strengths

This study has many strengths. It addresses an important gap in parasite biology and drug development, by addressing the native role of the potential antiplasmodial drug target FKBP35 and whether the compound FK506 works through inhibition of that putative target. The knockout data provide compelling evidence that the KBP35 protein is essential for asexual parasite growth after one growth cycle. Analysis of the FKBP35KO line also provides evidence that the effects of FK506 are likely not solely due to inhibition of that protein, but instead must have secondary targets whose function is essential. These data are important in the field of drug development as they may guide development away from structure-based FK506 analogs that bind more specifically to the FKBP35 protein.

Weaknesses:

There are also a few notable weaknesses in the evidence that call into question the conclusion in the article title that FKBP35 is definitely involved in ribosomal homeostasis. While the proteomics supports alterations in ribosome biogenesis factors, it is unclear whether this is a direct role of the loss of the FKBP35 protein or is more related to non-specific downstream effects of knocking down the protein. The CETSA data clearly demonstrate that FK506 binds PfKB35 at low nM concentrations, which is different than the IC50 noted in the parasite; however, the evidence that the proteins stabilized by uM concentrations of drug are actual targets is not completely convincing. Especially, given the high uM amounts of drug required to stabilize these proteins. This section of the manuscript would benefit from validation of a least one or two of the putative candidates noted in the text. In the live cell CETSA, it is noted that >50 ribosomal components are stabilized in drug treated but not lysate controls. Similarly, the authors suggest that the -soluble fraction of ribosomal components increases in drug-exposed parasites even at 37{degree sign}C and suggests that this is likely from smaller ribosomal proteins disengaging from larger ribosomal complexes. While the evidence is convincing that this protein may play a role in ribosome homeostasis in some capacity, it is not sure that the title of the paper "FKBP secures ribosome homeostasis" holds true given the lack of mechanistic data. A minor weakness, but one that should nonetheless be addressed, is the use of the term "delayed death phenotype" with regards to the knockout parasite killing. This term is most frequently used in a very specific setting of apicoplast drugs that inhibit apicoplast ribosomes, so the term is misleading. It is also possible that the parasites are able to go through a normal cycle because of the kinetics of the knockout and that the time needed for protein clearance in the parasite to a level that is lethal.

Overall, the authors set out to identify the native role of FKB35 in the P. falciparum parasites and to identify whether this is, in fact, the target of FK506. The data clearly demonstrate that FKBP35 is essential for parasite growth and provide evidence that alterations in its levels have proteomic but not transcriptional changes. However, the conclusion that FKBP35 actually stabilizes ribosomal complexes remains intermediate. The data are also very compelling that FK506 has secondary targets in the parasite aside from FKBP35; however, the high uM concentrations of the drug needed to attain results and the lack of biological validation of the CETSA hits makes it difficult to know whether any of these are actually the target of the compound or instead are nonspecific downstream consequences of treatment.

Author Response

The following is the authors’ response to the previous reviews

We thank the Reviewers and Editors for the evaluation of our revised manuscript.

We especially value the careful assessment of Reviewer 1; at the same time, we clearly disagree with the reviewer’s statement that the revised manuscript “is essentially unchanged”. As appreciated by the other Reviewers, we performed a key experiment (in our opinion the only conclusive experiment) to further solidify that FK506-treatment kills parasites in a FK506-independent manner. Of note, however, Reviewer 1 made us aware of an error in the legend of Figure 4F, which likely contributed to the confusion regarding the antiplasmodial effect of FK506: Unfortunately, we missed updating this legend to appropriately imbed the new experiment. We therefore incorrectly stated that parasites were exposed to FK506 for 48 hours after FK506 treatment at 4-10 hpi and 36-42 hpi in G1. In contrast to the experiments described in the initial submission, parasite survival was not measured 48 h later, but in G2 ring stage parasites, i.e. at a time point during which parasitemia is not affected by the knockout of PfFKBP35. We have now corrected this. As pointed out correctly by Reviewer 1, it would otherwise not be possible to disentangle the effects of the gene knockout and the drug. The setup we now present in Figure 4F, however, is clearly able to do so.

We apologize for the inaccuracy and hope this resolves the ambiguities regarding the FKBP35-independent antimalarial effect of FK506. In line with the comments of Reviewers 2 and 3, we believe that our findings on FK506 activity are of particular importance for the malaria research community. We therefore hope that the final eLife assessment will reflect this.

Version 3

Abstract

Abstract

Plasmodium falciparum accounts for the majority of over 600’000 malaria-associated deaths annually. Parasites resistant to nearly all antimalarials have emerged and the need for drugs with alternative modes of action is thus undoubted. The FK506-binding protein PfFKBP35 has gained attention as a promising drug target due to its high affinity to the macrolide compound FK506 (tacrolimus). Whilst there is considerable interest in targeting PfFKBP35 with small molecules, a genetic validation of this factor as a drug target is missing and its function in parasite biology remains elusive. Here, we show that limiting PfFKBP35 levels are lethal to P. falciparum and result in a delayed death-like phenotype that is characterized by defective ribosome homeostasis and stalled protein synthesis. Our data furthermore suggest that FK506, unlike the action of this drug in model organisms, exerts its anti-proliferative activity in a PfFKBP35-independent manner and, using cellular thermal shift assays, we identify putative FK506-targets beyond PfFKBP35. In addition to revealing first insights into the function of PfFKBP35, our results show that FKBP-binding drugs can adopt non-canonical modes of action – with major implications for the development of FK506-derived molecules active against Plasmodium parasites and other eukaryotic pathogens.

Introduction

Despite considerable progress in recent years, malaria remains one of the major global health threats (1). The apicomplexan parasite Plasmodium falciparum is responsible for the vast majority of the over 600’000 malaria deaths in 2020. Most of this burden is carried by infants and children under the age of five in sub-Saharan Africa. Malaria deaths increased by 12% compared to 2019, which can at least in part be explained by the COVID-19 related suspension of malaria control and treatment measures (1). In addition, the emergence of parasites resistant to most antimalarials, including artemisinin-based combination therapies, endangers current and future disease elimination efforts and underscores the need for developing drugs with alternative modes of action (2).

FK506 (tacrolimus), a well-characterized immunosuppressant, shows considerable activity against asexual blood stage parasites (3). P. falciparum encodes a single FK506-binding protein (FKBP) dubbed PfFKBP35 (PF3D7_1247400), which is considered to be a promising drug target (4, 5). FKBPs belong to the immunophilin family and are conserved throughout the eukaryotic kingdom (6, 7). In contrast to the malaria parasite, most other eukaryotes encode several FKBPs that, besides showing high affinity for the FK506 drug, exert peptidyl-prolyl isomerase (PPIase) activity. The PPIase moiety catalyzes cis-trans isomerization of proline residues – an event thought to be rate-limiting for the folding of many proteins (8). Independent of their PPIase activity, FKBPs act as chaperones to prevent protein aggregation under stress conditions (9, 10). Furthermore, some FKBP variants play a role in cellular signaling or gene regulation (11, 12). Human FKBP12 for instance is best-known for its ability to form a complex with the immunosuppressant drugs FK506 and rapamycin. These complexes inhibit the phosphatase activity of calcineurin and the kinase activity of mTOR (mechanistic target of rapamycin), respectively (13). Amongst many other effects, this inhibition results in decreased T-cell activation. Of note, the regulatory role of HsFKBP12 depends on the presence of rapamycin or FK506. In absence of drugs, this immunophilin neither interacts with mTOR nor with calcineurin and does hence not affect activity of these key components of eukaryotic cells. Due to the conserved nature of FKBPs, rapamycin and FK506 have also gained attention as potential antimicrobial drugs (3, 7).

Similar to its homologs in model organisms, PfFKBP35 harbors PPIase and chaperoning activity in vitro (5). Despite sharing key features with FKBPs of other eukaryotes, its role in P. falciparum remains elusive. PfFKBP35 is believed to interact with heat shock proteins (HSPs) and co-immunoprecipitates with cytoskeletal factors of P. falciparum (14). While PfFKBP35 is inhibiting the phosphatase activity of recombinant calcineurin in a FK506-independent manner (15-17), it does not co-localize with calcineurin in vivo (16). The protein is expressed throughout the 48-hour intra-erythrocytic development of P. falciparum and is found within the cytosol of ring stage parasites. In the older trophozoite and the replicating schizont stages, the protein was reported to also localize to the nucleus (16). While a transposon-based mutagenesis screen suggests that PfFKBP35 is essential for intra-erythrocytic replication (18), neither its cellular function nor its essentiality was confirmed experimentally in live parasites. The unknown role in parasite biology notwithstanding, PfFKBP35 is considered to be a viable drug target and, given the fact that P. vivax and P. knowlesi encode FKBP homologs that exhibit comparable PPIase activities (19, 20), future FKBP-targeted therapies may be effective across different Plasmodium species.

Crystal structures of PfFKBP35 in complex with FK506 and rapamycin revealed high affinity interactions with the FK506-binding domain of the protein (15, 21, 22). Consistent with these observations, in vitro enzyme activity assays showed that FK506 inhibits the PPIase activity of recombinant PfFKBP35 in a dose-dependent manner (5). Previous research aimed at exploiting structural differences between human and parasite-encoded FKBPs to circumvent the immunosuppressive activity of drug-bound HsFKBP12 (23-26). These efforts include investigation of non-covalent FKBP inhibitors such as adamantyl derivatives, macrocycles, the small molecule D44 (23-25, 27, 28) and the synthetic ligand for FKBP (SLF) as well as derivatives thereof designed to covalently bind PfFKBP35 (29). While D44 shows promising antimalarial activity, it fails to alter thermostability of the FK506-binding domains of PfFKBP35 and HsFKBP12 in vitro (29), suggesting that this compound does not directly interact with PfFKBP35. Up until now, PfFKBP35 has not been validated as a drug target using reverse genetics, and the link between PfFKBP35-interacting drugs and their antimalarial activity remains elusive (3).

Here, we generated inducible PfFKBP35 knock-out, knock-down as well as over-expression cell lines and demonstrate that PfFKBP35 is essential for asexual replication of blood forms. We show that the knock-out of PfFKBP35 results in reduced levels of ribosomal proteins and stalled protein synthesis. However, these effects are not reflected in the parasite transcriptome, indicating a vital role at the post-transcriptional level. While the mechanistic link between PfFKBP35 and ribosome homeostasis remains elusive, our data underpins the potential of PfFKBP35 as a drug target and, using cellular thermal shift assays (CETSA), we corroborate the high affinity binding of FK506 to PfFKBP35. Importantly, however, our data does not support scenarios in which FK506 inhibits the essential function of PfFKBP35, suggesting that it exerts its antimalarial activity in a PfFKBP35-independent manner, which has major implications for the development of future FK506-based antimalarials and/or PfFKBP35- inhibiting drugs.

Results

PfFKBP35 is essential for asexual replication

In a first step, using CRISPR/Cas9-mediated gene editing, we engineered a P. falciparum cell line allowing for a conditional knock-down of PfFKBP35 using the DD/Shield system (30, 31) (Figs. 1A and S1A-B). Despite the efficient depletion of PfFKBP35 under knock-down conditions, parasites did not show apparent growth defects (Figs. S1C-D). Immunofluorescence assays (IFA) showed that PfFKBP35 localizes to distinct foci within the parasite nucleus (Fig. 1B).

PfFKBP35 depletion causes delayed parasite death.

(A) Schematic of the N-terminally modified fkbp35 locus in NF54/iKD-FKBP35 parasites. dd, destabilization domain. gfp, green fluorescent protein. (B) Localization of PfFKBP35 in NF54/iKD-FKBP35 parasites at the ring, trophozoite and schizont stage assessed by immunofluorescence. DNA was stained with DAPI. Representative images are shown. Scale bar: 5 µm. DIC, differential interference contrast.(C) Schematic of the modified fkbp35 locus in NF54/iKO-FKBP35 parasites before and after loxP recombination. gfp, green fluorescent protein. (D) Flow cytometry plot showing GFP expression in G1 schizonts of FKBP35KO and FKBP35WT. DNA was stained with DRAQ5. (E) Time course showing the development of NF54/iKO-FKBP35 parasites. Upper panel: Schematic of the experimental setup. Middle panel: DNA content of the populations assessed by flow cytometry based on the SYBR Green intensity. Lower panel: Giemsa-stained blood smears. The knockout was induced at 0-6 hpi. Representative data of three biological replicates are shown. (F) Multiplication rates of FKBP35KO and FKBP35WT parasites from generation 1 to generation 2 (G1-G2) and G2-G3. The knockout was induced at 0-6 hpi. Multiplication rates were determined using flow cytometry. Data points represent the multiplication rates from G1-G2 and G2-G3. n=3, error bars represent the standard error of the mean. (G) Schematic showing time point of growth arrest (red arrow) following PfFKBP35 knock-out at different hpi (blue arrow). R, ring stage; T, trophozoite stage; S, schizont stage.

Next, we generated a cell line allowing for the conditional knock-out of the endogenous coding sequence using a dimerizable Cre recombinase (DiCre)-based system (32). Specifically, we introduced two loxP sites – one within a synthetic intron (33) downstream of the start codon and one after the stop codon of the fkbp35 coding region – and added a green fluorescent protein (gfp) sequence downstream of this expression cassette (Figs. 1C, S2A). As a result, these NF54/iKO-FKBP35 parasites express tag-free, wild-type (WT) PfFKBP35 protein under control conditions (FKBP35WT). Upon DiCre activation, the loxP sites are recombined and the fkbp35 locus is excised from the parasite genome (Figs. 1C, S2B). Under these knock-out conditions (FKBP35KO), parasites express cytosolic GFP under control of the endogenous fkbp35 promoter. This setup allows monitoring knock-out efficiency at the single cell level and revealed that DiCre successfully deleted fkbp35 in the vast majority of parasites (Fig. 1D).

In a first attempt, we induced the fkbp35 knock-out in synchronous ring stage NF54/iKO-FKBP35 parasites at 0-6 hours post erythrocyte invasion (hpi), i.e. prior to the expected peak transcription of fkbp35 (34). These FKBP35KO parasites showed a delayed death-like phenotype. In the first generation (G1), parasites completed intra-erythrocytic development without noticeable effects (Fig. 1E) and entered the next generation (G2) with a multiplication rate similar to that of the FKBP35WT control population (Fig. 1F). In G2, however, FKBP35KO parasites arrested at the late trophozoite/early schizont stage (approximately 30-36 hpi) with many cells failing to enter DNA replication in the S-phase (Fig. 1E). To investigate this phenomenon in more detail, we induced the fkbp35 knock-out at three different time points during the IDC. Similar to DiCre activation at 0-6 hpi, knocking out fkbp35 at 12-18 hpi resulted in parasite death at the trophozoite/schizont stage of the subsequent generation. However, when induced at 24-30 hpi or 34-40 hpi in G1, parasites did not only complete the current IDC but also successfully passed through G2 before they eventually arrested in G3 (Fig. 1G). Considering that low PfFKBP35 levels are sufficient to maintain the essential function of the protein, as observed with the NF54/iKD-FKBP35 parasites (see Fig. S1C-D), it is conceivable that proteins expressed between 0 and 24hpi, i.e. prior to deleting fkbp35 from the genome, allow parasites to enter G3.

The delayed death of FKBP35KO is reminiscent of a phenotype observed in apicoplast-deficient parasites (35, 36). We therefore tested if PfFKBP35 is involved in apicoplast biology, i.e. the synthesis and/or export of isoprenoid precursors, but did not find evidence for an association with this organelle (Fig. S2C). Furthermore, given the reported interaction of PfFKBP35 with heat shock proteins (14, 16), we also tested the parasitès response to heat stress, but could not detect altered heat-susceptibility between FKBP35KO and FKBP35WT cells (Fig. S2D).

PfFKBP35 is crucial for ribosome homeostasis and protein synthesis

To further elucidate the function of PfFKBP35, we induced the fkbp35 knock-out in early ring stages (0-6 hpi) and compared the proteomes of FKBP35KO and FKBP35WT parasites at the schizont stage of G1 (36-42 hpi) and at the trophozoite stage of the following generation (G2, 24-30 hpi) using a quantitative proteomics approach.

The knock-out of PfFKBP35 causes accumulation of pre-ribosome components in G1

Consistent with the efficient deletion of fkbp35 on the genomic level, FKBP35KO schizonts of G1 showed a 19-fold (18.9 +/-5.7 s.d.) reduction of FKBP35 and GFP levels were increased by 84-fold (83.7 +/-38.5 s.d.) when compared to the control population (Fig. 2A, Table S1). Besides PfFKBP35 and GFP, 50 parasite proteins were significantly deregulated in FKBP35KO schizonts and 34 of them found at elevated levels compared to the FKBP35WT control conditions. A gene ontology analysis using PANTHER (37) indicated enrichment of the “pre-ribosome” (fold enrichment = 30.4, p-value = 1.46E-5, 4/34 proteins) and “nucleolus” (fold enrichment = 6.77, p-value = 8.85E-4, 5/34 proteins) terms among these factors. While the pre-ribosome is a large subunit precursor of mature ribosomes, the nucleolus represents the primary site for the biogenesis of ribosomal subunits (38). In contrast to factors being more abundant in knock-out parasites, none of the 16 factors with lower abundance in FKBP35KO cells were associated with ribosome-related processes (Table S1). Despite being phenotypically silent in G1, the accumulation of pre-ribosome and nucleolar components hints at perturbed ribosome maturation in these cells. Specifically, it is conceivable that the observed buildup of ribosome intermediates results from reduced anabolic activity, i.e. the stalling of nucleolar pre-ribosome assembly (39).

PfFKBP35 is crucial for ribosome homeostasis and protein synthesis.

(A) Left panel: Proteome analysis of schizonts (36-42 hpi) in G1. The volcano plot shows change in protein abundance in FKBP35KO compared to FKBP35WT. Ribosomal proteins are highlighted in blue, proteins associated with the GO terms “pre-ribosome” or “nucleolus” in red. n=3. Right panel: Relative difference in PfFKBP35 and GFP abundance expressed as signal intensity measured by mass spectrometry. n=3, error bars represent the standard error of the mean. (B) Dose-response effect of the translation inhibitor cycloheximide on FKBP35KO and FKBP35WT parasites. The knockout was induced at 0-6 hpi and parasites were subsequently exposed to cycloheximide for 48 h. Parasite survival was quantified by flow cytometry. n=3, error bars represent the standard error of the mean. (C) Proteome analysis of trophozoites (24- 30 hpi) in G2. The volcano plot shows change in protein abundance in FKBP35KO compared to FKBP35WT. Ribosomal proteins are highlighted in blue. n=3. (D) GO enrichment analysis of significantly deregulated proteins in FKBP35KO trophozoites in G2. GO enrichment analysis was performed using the PANTHER “Overrepresentation Test” with the test type “Fisher’s Exact” and the correction method “False Discovery Rate”. (E) Assessment of translation activity using SUnSET. FKBP35KO and FKBP35WT trophozoites of G2 were incubated with puromycin for 1 hour. Incorporated puromycin was detected using an α-puromycin antibody. α-GAPDH served as loading control. n=3, a representative Western blot is shown.

Notwithstanding these effects, FKBP35KO cells were equally sensitive to the translation inhibitor cycloheximide compared to their FKBP35WT counterparts (Fig. 2B), indicating that ribosome function is not compromised at this point in time (40). This is consistent with normal cell cycle progression in G1 and unaltered multiplication rates of FKBP35KO parasites, as described above (see Figs. 1E-F), and shows that the prominent depletion of PfFKBP35 does not have a major impact on parasite viability in G1.

PfFKBP35 is essential for maintaining normal levels of functional ribosomes in G2

In the following generation (G2, 24-30 hpi), the proteome of FKBP35KO trophozoites showed substantial differences. Despite the absence of morphological changes at this point in time, 216 proteins were significantly deregulated (Fig. 2C). Among these, 50 proteins were found at higher abundance compared to FKBP35WT and their gene ontology terms point towards an effect on processes involved in energy metabolism (Fig. 2D).

The majority of deregulated factors (166 of 216) in G2, however, were less abundant in FKBP35KO compared to the FKBP35WT control cells. Very prominently, 20 of the top 50 hit molecules are ribosomal proteins that, on average, show a two-fold reduction (2.0 +/- 0.4 s.d.) in FKBP35KO parasites (Table S1). It is therefore not surprising that PANTHER identified the knock-out of PfFKBP35 to mainly affect ribosome-related processes, such as “cytoplasmic translation”, “ribosome assembly”, and “ribosomal large subunit biogenesis” (Fig. 2D) (37).

To assess if the reduced levels of ribosomal proteins result in lower protein synthesis rates, we employed the “surface sensing of translation” (SUnSET) technique. SUnSET is based on incorporation of the tyrosyl-tRNA analogue puromycin into nascent peptide chains and thereby allows monitoring newly synthesized proteins (41). As expected, we found that FKBP35KO trophozoites in G2 (24-30 hpi) showed reduced translation rates compared to the FKBP35WT control population (Fig. 2E), while protein synthesis was not affected under control conditions (rapamycin-induced DiCre activation in the NF54/DiCre parental line (Fig. S2E)).

Taken together, these results indicate that, under knock-out conditions, PfFKBP35 levels become limiting for ribosome maturation late during schizogony of G1, without having immediate effects on the steady-state levels of mature ribosomes. The knock-out of PfFKBP35 is thus phenotypically silent in G1. In absence of a continuous supply of mature ribosomes under FKBP35KO conditions, however, FKBP35KO parasites fail at maintaining translation at a level that would support normal parasite development in the next generation (G2).

The knock-out of PfFKBP35 is transcriptionally silent

To examine if the altered proteome of FKBP35KO parasites is the result of PfFKBP35- controlled transcription, we compared the transcriptional profiles of FKBP35WT and FKBP35KO in a time course RNA-Sequencing (RNA-Seq) experiment spanning two IDCs. Specifically, we induced the knock-out of FKBP35 in synchronous parasites at 0-6 hpi and collected RNA from parasite populations at three and four subsequent time points (TPs) in G1 and G2, respectively (Figs. 3A, S3A). With the exception of TP7, these time points cover a time frame during which FKBP35KO cells cannot be distinguished from their FKBP35-expressing counterparts on a morphological level (compare Fig. 1E).

The knockout of PfFKBP35 remains silent at the transcriptomic level.

(A) Setup of the RNA-Seq experiment. RNA was collected at seven indicated time points (TP). (B) Volcano plots show differential gene expression between FKBP35KO and FKBP35WT parasites. Genes are color-coded according to the fold change and fdr (false discovery rate), n=3. (C) Mapping of FKBP35KO and FKBP35WT transcriptomes to a high-resolution reference (42). Spearman rank coefficients (ρ) show correlation between the sampled transcriptomes and the reference data set (42). The time points with the highest correlation coefficient are indicated.

As expected, the transcriptome of FKBP35KO showed a prominent reduction of steady-state fkbp35 transcripts already in TP1 (11.3-fold +/-2.0 s.d.) compared to the FKBP35WT control population (Fig. S3B, Table S2). During peak transcription at 30-36 hpi (TP2), this difference increased to 240-fold (238.4 +/-20.0 s.d.) and, consistent with the efficient recombination at the fkbp35 locus, these cells started transcribing gfp. Notwithstanding the prominent depletion of fkbp35 transcripts, the transcriptome of FKBP35KO is only marginally affected up to and including 24-30 hpi in G2 (TP6). By contrast, at 30-36 hpi (TP7) the transcription of knock-out parasites shows major changes (Fig. 3B). Considering the appearance of morphological changes in FKBP35KO at this time point, effects on the transcriptome are unlikely to represent specific activities of PfFKBP35 in gene regulation. Indeed, mapping the transcriptome of individual time points to a high resolution reference (42), revealed a significant slow-down of FKBP35KO compared to FKBP35WT parasites at TP7 (Figs. 3C, S3C). This indicates aborted cell cycle progression in FKBP35KO parasites during a phase that is characterized by a rapid increase in biomass – possibly due to a translation and/or size-dependent cell cycle checkpoint (43). At earlier time points (TP1-TP6), the transcriptomes of FKBP35KO and FKBP35WT are highly comparable and the differences observed on the level of individual genes clearly fail at explaining the proteomic changes observed in FKBP35KO parasites at 24-30 hpi in G2 (TP6), reiterating that the essential activity of PfFKBP35 is linked to transcription-independent processes.

Probing FK506 interactions in Cellular Thermal Shift Assays (CETSA)

PfFKBP35 emerged as a promising antimalarial drug target (4) and considerable efforts were made to define interactions of this immunophilin with small molecules in the last decade (24-26, 44). Several studies conclusively demonstrated that FK506 – the most well-known ligand of mammalian FKBPs (45) – binds to PfFKBP35 (15, 21, 22) and kills asexual blood stage parasites with a half-maximal inhibitory concentration (IC50) in the low micromolar range at 1.9 µM (3).

To investigate interactions of FK506 with P. falciparum proteins, we used the cellular thermal shift assay followed by mass spectrometry (MS-CETSA). This approach exploits the fact that, once drug-protein complexes are formed, thermostability of the target protein is altered (46). We applied the so-called isothermal dose-response variant of CETSA using FK506 (47). Here, target proteins are identified based on their dose-dependent stabilization by drugs under thermal challenge (different temperatures) when compared to a non-denaturing condition (37 °C), using the change in the area under the curve (ΔAUC) and the dose-response curve goodness of fit (R2) as metrics (Fig. 4A) (47).

CETSA identifies FK506 interaction partners.

(A) CETSA results. The relative shift in protein abundance (area under the curve of heat-challenged samples normalized against the non-denaturing control, ΔAUC) as a function of R2 (goodness of fit as a measure for the dose-response effect). The dashed lines indicate an arbitrary cutoff of 3 times the median absolute deviation (MAD) of each dataset and R2=0.8. Three different temperatures were used for protein denaturation (indicated by colors). (B) Dose-dependent stabilization of P. falciparum and human FKBPs detected using lysate and live cell CETSA. Bullets represent the abundance of non-denatured proteins in response to increasing FK506 concentrations under thermal challenge relative to the DMSO control. Colors indicate the temperatures used for heat challenge; the black curve represents the non-denaturing (37°C) control. (C) Proteins with the most pronounced stability shift under thermal challenge in the lysate CETSA experiment are listed. The EC50 is the concentration at which the half-maximal effect of the FK506-induced thermal stabilization is reached. The mutagenesis index score determined by Zhang et al. (18) is a measure to predict if a protein is essential for asexual growth: 0=essential, 1=non-essential. (D) Fitted live cell CETSA curves of 10 representative ribosomal proteins at 37°C and 55°C. Ribosomal proteins show a FK506 dose-dependent increase in soluble protein fraction. (E) Venn diagram showing the overlap between ribosomal proteins deregulated in the PfFKBP35 knock-out cell line (less abundant in G2 FKBPKO trophozoites determined by proteomics) and the ribosomal proteins stabilized by FK506 in live cell CETSA. The number of factors identified in either or both assays is indicated. (F) Dose-response effect of FK506 on FKBP35KO and FKBP35WT parasites. The knockout was induced at 0-6 hpi and parasites were exposed to FK506 at 4-10 hpi and 36-42 hpi for 48h. n=3, error bars represent the standard error of the mean. (G) Dose-response effect of FK506 on NF54/iKD-FKBP35 and NF54/iOE-FKBP35 parasites in presence and absence of Shield-1 (Shld). Parasites were split at 0-6 hpi and subsequently treated with Shld or the vehicle control EtOH. Data points represent the mean of three and two independent biological replicates for the NF54/iOE-FKBP35 and NF54/iKD-FKBP35 cell line, respectively. Error bars represent the standard error of the mean.

CETSA identifies targets of FK506

In a first step, we performed CETSA in combination with drug-exposed protein extracts. Since proteins lose their physiological context during the process of protein extraction, this variant of the CETSA approach is designed to identify direct protein-drug interactions (46, 47). Not surprisingly, we found PfFKBP35 to be stabilized by FK506 in a dose-dependent manner when challenged at 60 °C with a half maximal effective concentration (EC50) of 152.0 nM (Fig. 4B). The fact that FK506 exhibits its stabilizing activity already in the low nanomolar concentration range indicates high affinity interactions between the protein-drug pair.

Besides PfFKBP35, FK506 stabilizes a number of other proteins in a dose-dependent manner (Figs. 4C and S6). A few of these factors show pronounced sigmoidal stabilization profiles and are predicted to be essential (Fig. 4C) (18). Amongst others, N6-methyl transferase MT-A70 (PF3D7_0729500), histone deacetylase HDA2 (PF3D7_1008000), pre-mRNA-splicing factor CEF1 (PF3D7_1033600), FeS cluster assembly protein SufD (PF3D7_1103400) and the H/ACA ribonucleoprotein complex subunit 1 (PF3D7_1309500) showed most pronounced stabilization profiles (Fig. S6 and Table S3). In contrast to PfFKBP35, most of these likely essential candidates are stabilized at FK506 concentrations close to the drug`s IC50 (3). It can therefore not be excluded that the binding of FK506 to at least one of these factors is contributes to the antimalarial activity of the drug.

The PfFKBP35/FK506 pair interferes with ribosomal complexes

Given the high-affinity interaction of PfFKBP35 with FK506 and its association with ribosomal proteins, we set out to describe the native context of PfFKBP35 using the “live cell variant” of CETSA. In contrast to focusing exclusively on direct protein-drug interactions, live cell CETSA allows capturing drug-induced shifts in protein stability beyond those of direct binding partners (47) and amongst others allows identifying factors present in the same protein complex as the target protein. To do so, CETSA takes advantage of the fact that factors within complexes have similar denaturation temperatures due to their co-aggregation upon heat-induced denaturation (48). As a consequence, this CETSA variant generally identifies more factors compared to the protein extract-based CETSA approach (Fig. 4A).

As expected, and confirming the results above, live cell CETSA shows that FK506 alters the thermal stability of PfFKBP35 already at very low drug concentrations when challenged at 60 °C with an EC50 of 286.7 nM (Fig. 4B). Interestingly, and in contrast to the protein lysate-based approach, live cell CETSA revealed that more than 50 ribosomal proteins are stabilized in presence of FK506 when challenged at 55 °C, with a median EC50 of 45.9 nM (25th percentile = 34.2, 75th percentile = 58.9). Of note, these ribosomal proteins were stabilized at virtually identical FK506 concentrations (Figs. 4D, and S7), indicating that the drug – directly or indirectly – interacts with ribosomal complexes. This is in further agreement with the fact that these ribosomal factors were not identified by protein lysate-based CETSA (compare Figs. 4A, S6).

Worth mentioning, we observed a drug dose-dependent increase in soluble protein abundance of ribosomal proteins also under non-denaturing conditions, i.e. at 37 °C, with a median EC50 of 1.94 µM (25th percentile = 1.85, 75th percentile = 2.1) (Figs. 4D, S7). Generally, due to the absence of a denaturing heat challenge, such a drug-dependent enrichment in soluble protein levels does not indicate binding-induced thermal stabilization of proteins (47). Instead, this can be the result of increased protein synthesis, altered protein conformation or, alternatively, occur in response to factors disengaging from larger protein complexes. The fact that FK506 solubilizes a high number of ribosomal proteins within a very narrow concentration window also at 37 °C, implies that these factors are liberated from the same ribosomal complex or from complexes that interact with each other and shows that FK506 negatively affects stability of ribosomal structures in live cells. However, these effects are limited to high, presumably saturating FK506 concentrations. Importantly, the vast majority of ribosomal proteins identified in live cell CETSA are shared with those de-regulated in PfFKBP35 knock out cells (Fig. 4E), pointing towards a shared localization of FK506 and its PfFKBP35 target in live parasites.

Together with the above-described importance of PfFKBP35 in ribosome homeostasis, these results strongly suggest that FK506 is recruited to the activity site of PfFKBP35 – most probably the nucleolar maturation sites of ribosomes – where it interacts with ribosomal complexes.

FKBP-targeting drugs fail at inhibiting the essential function of PfFKBP35

Considering the proposed direct interaction between FK506 and FKBP35 in P. falciparum, we expected that FKBP35KO parasites show altered sensitivity to the drug. Consistent with published data, FK506 killed FKBP35WT cells with an IC50 of 1942 nM (+/-193 nM s.d.) when administered at 4-10 hpi (Fig. 4F). Surprisingly however, and despite the prominent depletion of the protein under knock-out conditions, FKBP35KO and FKBP35WT parasites were equally sensitive to FK506 (IC50 = 1741 nM +/- 187 s.d.) (Fig. 4F). Similarly, neither the efficient knock-down of PfFKBP35 in NF54/iKD-FKBP35 parasites (compare Fig. S1C), nor its over-expression (OE) in NF54/iOE-FKBP35 cells (Fig. S4A-C), resulted in the expected sensitivity change to FK506 treatment (Fig. 4G). In addition to this apparent independence of the drug to varying PfFKBP35 levels, and in agreement with earlier studies (3, 24, 49), FK506 unfolds its antimalarial activity within the same cycle after treatment. Standing in stark contrast to the delayed death-like phenotype of FKBP35KO parasites (compare Fig. 1E-G), this immediate activity on parasite survival substantiates the notion that FK506 must target essential parasite factors other than PfFKBP35.

To rule out that the timing of drug exposure is masking expected links between FK506 and PfFKBP35, we tested whether administering FK506 at 30-36 hpi and 36-42 hpi (instead of 4- 10 hpi) affects parasite survival in a PfFKBP35-dependent manner. However, we did not observe a change compared to FKBP35WT parasites (Fig. S5A), ruling out the possibility that FKBP35KO parasites were killed by FK506 before PfFKBP35 levels were different in the two populations. Further, to account for a scenario in which the activity of FK506 on PfFKBP35 is masked by a temporal delay between drug exposure and parasite killing, we induced the knock-out of PfFKBP35 at 34-40 hpi (instead of 0-6 hpi) and allowed FKBP35KO parasites to proceed to G2, before exposing them to different FK506 concentrations. Despite these efforts, FKBP35WT and FKBP35KO parasites were equally sensitive to FK506 (Fig. S5B, S5C) and the calcineurin inhibitor cyclosporin A (Fig. S5D-E).

In addition to FK506, we evaluated the PfFKBP35-specific inhibitor D44 (N-(2-Ethylphenyl)- 2-(3h-Imidazo[4,5-B]pyridin-2-Ylsulfanyl)acetamide) but, surprisingly, we failed to reproduce the previously reported dose-response effect (24). We tested D44 from two suppliers using NF54/iKO-FKBP35, NF54 wild type and K1 parasite strains using different assay setups, including the gold standard method that is based on the incorporation of tritium labeled hypoxanthine into DNA of replicating cells (50). Still, we could not detect any activity of D44 on asexual replication of blood stage parasites (Figs. S5F-G).

Taken together, these results are in clear conflict with the perception that FK506 and D44 target the essential activity of PfFKBP35.

Discussion

Members of the FKBP family are involved in regulating diverse cellular processes in eukaryotic as well as in prokaryotic cells (8, 51). In contrast to most other eukaryotes, the human malaria parasite P. falciparum encodes a single FKBP only (16). The enzymatic activity of PfFKBP35, as well as the interactions of the protein with FKBP-targeting drugs, was subject to considerable drug-focused research in vitro. Besides demonstrating the drug`s affinity to PfFKBP35 by co-crystallization, these efforts uncovered that FK506 inhibits enzymatic activity of the protein`s PPIase domain (5). Together with the anti-malarial activity of FK506, these data offered compelling evidence for an essential but yet uncharacterized role of PfFKBP35 during blood stage development of P. falciparum.

Here, we set out to investigate the role of PfFKBP35 in parasite biology. Using an inducible knock-out approach, we show that PfFKBP35 is indeed essential for intra-erythrocytic development of asexually replicating P. falciparum parasites and CETSA experiments confirmed high affinity binding of FK506 to its PfFKBP35 target. While the knock-out of PfFKBP35 does not markedly affect transcription, FKBP35KO cells show reduced levels of ribosomal proteins and impaired protein synthesis on the global scale. Surprisingly however, these effects, as well as the consequent death during the late trophozoite/early schizont stage, only manifest in the cycle following knock-out induction without affecting cell cycle progression in the first generation (G1) or multiplication rates from G1 to G2. In contrast to the delayed death-like phenotype of FKBP35KO cells, FK506 unfolds its parasiticidal activity without a lag phase, i.e. within the IDC of treatment (24), and the dose-response relationship between FK506 and parasite death is irresponsive to altered PfFKBP35 levels. Taken together, our data strongly indicate that the parasite-killing activity of FK506 is PfFKBP35-independent.

Knocking out PfFKBP35 early during intra-erythrocytic development has only minor effects on the parasite proteome in G1. Being relatively subtle, these changes may still point towards an important function of PfFKBP35: Factors associated with the nucleolus – the site of ribosome biogenesis – are found at higher abundance in knock-out cells. In the next cycle, FKBP35KO parasites show a dramatic decrease of factors associated with ribosome-and translation-related processes and, likely a consequence thereof, globally reduced protein synthesis during trophozoite development (G2, 24-30 hpi). Importantly, at this point in time, FKBP35KO cells still show ordinary morphology and their transcriptional activity remains largely unaffected. The observed changes in the abundance of ribosomal factors must therefore result from transcription-independent activity of PfFKBP35. While ribosomal components are present at higher levels in G1, they are drastically reduced in FKBP35KO cells in G2. These opposing effects likely result from PfFKBP35 activity in a shared cellular process. Specifically, it is conceivable that low PfFKBP35 levels inhibit ribosome maturation in G1, explaining the accumulation of (pre-) ribosomal factors. While the levels of functional ribosomes are sufficient to support continued translation in G1, ribosomes likely become limiting in the next generation (G2), which eventually results in stalled translation activity and concomitant parasite death at the onset of schizogony.

In contrast to the significant effects on the protein levels in G2, the transcription of FKBP35KO remains largely unaffected by the loss of PfFKBP35 and becomes apparent only during late trophozoite/early schizont development in G2 (30-36 hpi); i.e. at a time during which cell cycle progression is slowed down in FKBP35KO parasites. In fact, at this stage the transcriptional signature of FKBP35KO parasites is dominated by the slowed progression through intra-erythrocytic development, rather than resulting from specific gene regulatory events. While this stalling is consistent with the observed effects on ribosome homeostasis and protein synthesis, it demonstrates that, in contrast to certain FKBP family members in other organisms, PfFKBP35 does not have apparent roles in transcriptional control.

Regarding the well-characterized and fast-acting parasiticidal activity of FK506 on P. falciparum, it was somewhat surprising to observe a delayed death-like phenotype for PfFKBP35 knock-out parasites (24). We hence set out to investigate unifying explanations. The delayed death of FKBP35KO parasites is most easily be explained by residual activity of PfFKBP35 proteins synthesized prior to inducing the genomic knock-out of fkbp35. In such a scenario, low PfFKBP35 in FKBP35KO would be sufficient to allow completion of the IDC, before eventually dropping below a level that is required for supporting parasite survival in the next cycle. Despite intense efforts, we failed at confirming this hypothesis and our data strongly favors an alternative scenario in which FK506 kills parasites in a PfFKBP35-idenpendent manner. Aside from the fact that FK506 exerts its parasiticidal activity within the same IDC, neither the conditional knock-out, knock-down nor the overexpression of PfFKBP35 altered the parasite’s sensitivity to FK506; and this irrespective of the profoundly deregulated PfFKBP35 levels in the respective mutant lines (Fig. 4F). For instance, administering FK506 36 hours post induction of the knock-out in early ring stages (36-42 hpi) – i.e. at a time point where FKBP35KO cells are only left with about 5% of the regular PfFKBP35 levels (19-fold reduction compared to FKBP35WT parasites, Fig. 2A) – does not affect the dose-response of parasites to the drug. Further, our attempts to demonstrate that the timing of drug exposure is masking the expected links between FK506 and its PfFKBP35 drug target failed (compare Figs. 4A and S5A-C). Our data is thus incompatible with the notion that the parasite-killing activity of FK506 results from inhibiting the essential function of PfFKBP35.

Since binding of FK506 to PfFKBP35 alone cannot explain the drug’s antimalarial activity, we aimed identifying additional FK506 targets. CETSA performed on protein lysates identified a number of parasite factors binding to FK506. Compared to the PfFKBP35/FK506 pair, however, these interactions are of considerably lower affinity. The adenosine-methyltransferase MT-A70 (PF3D7_0729500) and the pre-mRNA-splicing factor CEF1 (PF3D7_1033600), for instance, were stabilized by FK506 with EC50 concentrations of 2.0 and 3.1 µM, respectively. While this affinity is relatively poor compared to the strong interaction formed between PfFKBP35 and FK506, one should be cautious designating the binding of FK506 to these proteins as irrelevant “off-target” effects. In fact, their inhibition may hypothetically mask effects of FK506 on PfFKBP35 – a possibility that shall briefly be discussed in the following section.

Assuming that FK506 is inhibiting the essential function of PfFKBP35, one would expect parasites to show a delayed death-like phenotype similar to that observed in FKBP35KO parasites. The presence and inhibition of other essential targets may, however, mask such an effect. Consistent with this possibility, FK506 kills P. falciparum parasites at 1.9 µM (+/-0.2 µM s.d.), i.e. at concentrations similar to those required for stabilizing MT-A70 and CEF1 under thermal challenge in CETSA. However, numerous attempts to prove that FK506 indeed inhibits essential PfFKBP35 functions failed: Neither knocking out PfFKBP35 at different time points nor adding FK506 to different stages of FKBP35KO parasites could provide evidence for a PfFKBP35-dependent effect. This indicates that either the binding of FK506 does not interfere with the essential role of PfFKBP35, or that PfFKBP35 is inhibited only at high FK506 concentrations that also inhibit other essential factors. In this context it is worth mentioning that parasites remain unaffected by a highly efficient knock-down of endogenous PfFKBP35 (observed with NF54/iKD-FKBP35; compare Figs. S1C-D), which demonstrates that very low protein levels are sufficient to maintain the essential function of PfFKBP35. It is thus conceivable that high FK506 concentrations are required to fully inhibit PfFKBP35 – a view that is further supported by data obtained with the live cell CETSA approach discussed hereafter.

Besides identifying direct binding of drugs to their target, live cell CETSA allows capturing indirect effects on factors that act in the same pathway or in the same protein complex as the drug-bound target (47). While this approach confirmed the high affinity binding of FK506 to PfFKBP35, it also revealed that more than 50 ribosomal proteins share a very prominent CETSA signature: First, they are thermally stabilized by FK506 within a narrow concentration window of the drug. Second, higher levels of FK506 increase the solubility of these ribosomal components also at 37 ℃. These results indicate that the FK506/PfFKBP35 pair interacts with ribosomal complexes and that FK506 is able to induce the dissociation of these ribosomal structures under physiological conditions. Importantly, and in contrast to the binding of FK506 to PfFKBP35, this effect on ribosomes is only observed at high concentrations of the drug. Together, these data likely explain the discrepancy observed between the high binding affinity of FK506 for PfFKBP35 and the low or even absent effect of the drug on PfFKBP35 function in blood stage parasites.

In line with our data, previous biochemical approaches revealed potential links between PfFKBP35 and ribosomes. Using co-immunoprecipitation, Leneghan and colleagues revealed that PfFKBP35 is interacting with 480 proteins, of which 31 are ribosomal factors (14). Indeed, FKBP family members are essentially involved in controlling ribosomes in other systems. The human FKBP variant HsFKBP25, for instance, is recruited to the pre-ribosome where its chaperone activity appears to be required for the assembly of the ribosomal large subunit (52). While the FKBP variant of yeast, Fpr4, interacts with the ribosome biogenesis factor Nop53 (53), the Escherichia coli trigger factor (TF) is part of the ribosomal complex and crucial for the folding of nascent proteins (54, 55). Considering the marked reduction of ribosomal proteins in FKBP35KO parasites, it is tempting to speculate that PfFKBP35 is chaperoning ribosome biogenesis similar to Fpr4 and FKBP25 in Saccharomyces cerevisiae and humans, respectively.

The fact that PfFKBP35 localizes to distinct nuclear foci and its depletion causes dysregulation of ribosomes points towards a role within the nucleolus – the nuclear compartment driving ribosome biogenesis (38). Using IFAs, we found that PfFKBP35 does not co-localize with the nucleolar marker fibrillarin (56) (Fig. S8). However, its localization appears to be intertwined with that of this rRNA methyltransferase. Specifically, PfFKBP35 is mostly found in close vicinity to fibrillarin-rich regions. Nucleoli in eukaryotic cells generally consist of two or, in higher eukaryotes, three sub-compartments. Of those, the granular component (GC) represents the site of pre-ribosome assembly and encompasses fibrillarin-enriched central regions of nucleoli (57). The spatial correlation observed between PfFKBP35 and fibrillarin may thus indicate localization to different sub-nucleolar compartments, with PfFKBP35 occupying the GC area of the nucleolus.

In summary, we demonstrate that limiting PfFKBP35 levels are lethal to P. falciparum and result in a delayed death-like phenotype that is characterized by perturbed ribosome homeostasis and defective protein synthesis, which is likely linked to the reduced biogenesis of functional ribosomes in the nucleolus. We further found no evidence that FKBP-binding drugs, including FK506, exert their parasiticidal activity in a PfFKBP35-dependent manner, urging strong caution for the future development of FKBP-targeting antimalarials, especially when based on FK506 and structural derivatives thereof. In addition to revealing first insights into the essential function of PfFKBP35 in ribosome homeostasis of P. falciparum, we are convinced that the presented data offer valuable information for target-based efforts in malaria drug discovery.

Methods

Parasite culture

P. falciparum culture and synchronization was performed as described (58, 59). Parasites were cultured in AB+ or B+ human red blood cells (Blood Donation Center, Zürich, Switzerland) at a hematocrit of 5% in culture medium containing 10.44 g/L RPMI-1640, 25 mM HEPES, 100 μM hypoxanthine, 24 mM sodium bicarbonate, 0.5% AlbuMAX II (Gibco #11021-037), 0.1 g/L neomycin, and 2 mM choline chloride (Sigma #C7527). To achieve stabilization of DD-tagged proteins in the cell lines NF54/iKD-FKBP35 and NF54/iOE-FKBP35, parasites were cultured in presence of 625 nM Shield-1 (30, 31). To induce the DiCre-mediated recombination of loxP sites in NF54/iKO-FKBP35 parasites, cultures were split and treated for 4 h with 100 nM rapamycin or the corresponding volume DMSO, which served as vehicle control, giving rise to FKBP35KO and FKBP35WT populations, respectively (32). Cultures were gassed with 3% O2, 4% CO2 and 93% N2 and incubated in an airtight incubation chamber at 37°C.

Cloning of transfection constructs

CRISPR/Cas9-based gene editing of the NF54 parasites was performed using a two-plasmid approach as previously described (60, 61). This system is based on co-transfection of a plasmid that contains the expression cassettes for the Cas9 enzyme, the single guide RNA (sgRNA) and the blasticidin deaminase (BSD) resistance marker (pBF-gC), and a pD-derived donor plasmid that contains the template for the homology-directed repair of the Cas9-induced DNA double strand break (Fig. S2A) (61).

The plasmids pBF-gC_FKBP-3’, pBF-gC_FKBP-5’ and pBF-gC_P230p targeting the 3’ or 5’ end of fkbp35 (PF3D7_1247400) or the p230p (PF3D7_0208900) locus (62), respectively, were generated by ligation of two annealed oligonucleotides (gRNA_top and gRNA_bottom) into the BsaI-digested pBF-gC backbone (61) using T4 DNA ligase (New England Biolabs).

The donor plasmid pD_FKBP35-iKO was generated by assembling six DNA fragments in a Gibson reaction (63) using (i) the plasmid backbone amplified by polymerase chain reaction (PCR) from pUC19 (primers PCRa_F/PCRa_R) (61), (ii/iii) the 5’ and 3’ homology regions amplified from genomic DNA (primers PCRb_F/PCRb_R and PCRc_F/PCRc_R, respectively), (iv) the green fluorescent protein (gfp) sequence amplified from pHcamGDV1- GFP-DD (primers PCRd_F/PCRd_R) (61), (v) the HRPII terminator amplified from genomic DNA (primers PCRe_F/PCRe_R), and (vi) the fkbp35 sequence containing an artificial loxPint (33) amplified from a P. falciparum codon-optimized synthetic sequence (primers PCRf_F/PCRf_R) ordered from Genscript.

The donor plasmid pD_FKBP35-iKD was generated by assembling five DNA fragments in a Gibson reaction (63) using (i) the plasmid backbone amplified from pUC19 (primers PCRa_F/PCRa_R) (61), (ii/iii) the 5’ and 3’ homology regions amplified from genomic DNA (primers PCRg_F/PCRg_R and PCRh_F/PCRh_R, respectively), (iv) the gfp sequence amplified from pHcamGDV1-GFP-DD (primers PCRi_F/PCRi_R) (61), and (v) the destabilization domain (dd) sequence amplified from pHcamGDV1-GFP-DD (primers PCRj_F/PCRj_R) (61).

The donor plasmid pD_FKBP35-iOE was generated by assembling three DNA fragments in a Gibson reaction (63) using (i) the NheI/PstI-digested plasmid pkiwi003 (64), (ii) the fkbp35 coding sequence amplified from genomic DNA (primers PCRk_F/PCRk_R), and (iii) the gfp-dd sequence amplified from pHcamGDV1-GFP-DD (primers PCRl_F/PCRl_R) (61).

Oligonucleotides are listed in Table S4.

Transfection and selection of gene-edited parasites

P. falciparum transfection using the CRISPR/Cas9 two-plasmid approach was performed as described previously (61). Briefly, 50 μg of each plasmid (pBF-gC_FKBP-3’ and pD_FKBP35- iKO, pBF-gC_FKBP-5’ and pD_FKBP35-iKD, pBF-gC_P230p and pD_FKBP35-iOE) were co-transfected into an NF54::DiCre line (65), in which ap2-g was tagged with the fluorophore mScarlet as described previously (66). Transgenic parasites were selected with 2.5 μg/mL blasticidin-S-hydrochloride, which was added 24 h after transfection for 10 days. Transgenic populations were usually obtained 2-3 weeks after transfection and correct editing of the modified loci was confirmed by PCR on gDNA (Figs. S1B, S2B and S4B). Primer sequences used for these PCRs are listed in Table S4. Clonal parasite lines were obtained by limiting dilution cloning (67).

DNA content analysis

NF54/iKO-FKBP35 parasites were fixed in 4% formaldehyde/0.0075% glutaraldehyde for 30 min at room temperature and washed 3 times in PBS and stored at 4°C. Samples were collected from three independent biological replicates. RNA was digested by incubation with 0.1 mg/mL RNAse A in PBS containing 0.1% Triton-X100 for 15 min at room temperature. Nuclei were stained using 1X SYBR Green 1 DNA stain (Invitrogen S7563) for 20 min and washed three times in PBS. The fixation step quenched the GFP signal detected by the flow cytometer (Fig. S9). SYBR Green intensity was measured using a MACS Quant Analyzer 10 and analyzed using the FlowJo_v10.6.1 software.

Western blotting

Saponin lysis of 500 µL infected red blood cells at 3-5% parasitaemia was performed by incubation for 10 min on ice in 3 mL ice-cold 0.15% saponin in PBS. The parasite pellets were washed two times in ice-cold PBS and resuspended in Laemmli buffer (62.5 mM Tris base, 2% SDS, 10% glycerol and 5% 2-mercaptoethanol). Proteins were separated on 4-12% Bis-Tris gels (Novex, Qiagen) using MOPS running buffer (Novex, Qiagen). Then proteins were transferred to a nitrocellulose membrane (GE healthcare #106000169), which was blocked with 5% milk powder in PBS/0.1% Tween (PBS-T) for 1 h. The membrane was probed using the primary antibodies mAb mouse α-GFP (1:1’000, Roche Diagnostics #11814460001), mAb mouse α-PfGAPDH (1:20’000) (68), or mAb mouse α-puromycin (1:5’000, Sigma MABE343) diluted in blocking solution. After 2 h incubation, the membrane was washed 5 times in PBS-T before it was incubated with the secondary antibody goat α-mouse IgG (H&L)-HRP (1:10’000, GE healthcare #NXA931). After 1 h incubation, the membrane was washed 5 times and the signal was detected using the chemiluminescent substrate SuperSignal West Pico Plus (Thermo Scientific, REF 34580) and the imaging system Vilber Fusion FX7 Edge 17.10 SN. Nitrocellulose membranes were stripped by incubating for 10 min in a stripping buffer (1 g/L SDS, 15 g/L glycine, 1% Tween-20, pH 2.2) and washed three times in PBS-T before blocking and re-probing.

Drug dose-response experiments

Dose-response relationships were assessed by exposing synchronous ring stage parasites at 0.5% parasitaemia and 1.25% hematocrit to 12 drug concentrations using a two-step serial dilution in 96-well plates (Corning Incorporated, 96-well cell culture plate, flat bottom, REF 3596) in three independent biological replicates. The plates were gassed and incubated for 48 h at 37 °C. FK506 (MedChemExpress HY-13756), D44 (ChemBridge 7934155 and ChemDiv 7286-2836) and cyclosporin A (Sigma 30024) stocks of 10 mM were prepared in DMSO and working solutions were prepared immediately before the experiment. To measure parasite survival using flow cytometry, cultures were stained in 1X SYBR Green 1 DNA stain (Invitrogen S7563) in a new 96-well plate (Corning Incorporated, 96-well cell culture plate, round bottom, REF 3788) and incubated in the dark for 20 min. The samples were washed once in PBS. Using a MACS Quant Analyzer 10, 50’000 events per condition were measured and analyzed using the FlowJo_v10.6.1 software. Infected RBCs were identified based on SYBR Green 1 DNA stain intensity. The gating strategy is shown in Fig. S9. To quantify parasite survival based on hypoxanthine incorporation into DNA, compounds were dissolved in DMSO (10 mM), diluted in hypoxanthine-free culture medium and titrated in duplicates over a 64-fold range (6 step twofold dilutions) in 96 well plates. 100 μl asexual parasite culture (prepared in hypoxanthine-free medium) were added to each well and mixed with the compound to obtain a final haematocrit of 1.25% and a final parasitemia of 0.3%. After incubation for 48 hrs, 0.25 μCi of [3H]-hypoxanthine was added per well and plates were incubated for an additional 24 hrs. Parasites were then harvested onto glass-fiber filters using a Microbeta FilterMate cell harvester (Perkin Elmer, Waltham, US) and radioactivity was counted using a MicroBeta2 liquid scintillation counter (Perkin Elmer, Waltham, US). The results were recorded and expressed as a percentage of the untreated controls. Curve fitting and IC50 calculations were performed using a non-linear, four-parameter regression model with variable slope (Graph Pad Prism, version 8.2.1).

Fluorescence microscopy

To localize PfFKBP35, NF54/iKD-FKBP35 parasites were fixed in 4% formaldehyde/0.0075% glutaraldehyde for 30 min at room temperature and washed three times in PBS. The fixed cells were permeabilized by incubation with 0.1% Triton-X100-containing PBS for 15 min. After two washing steps in PBS, the cells were incubated in a blocking/aldehyde quenching solution (3% BSA in PBS complemented with 50 mM ammonium chloride). The samples were incubated with the primary antibodies rabbit α-GFP (1:400, Abcam ab6556) and mAb mouse α-fibrillarin (1:100, Santa Cruz Biotechnology sc-166021) for 1 h in 3% BSA in PBS and washed 3 times in PBS. Subsequently, the samples were incubated with the secondary antibodies goat α-rabbit IgG (H&L) Alexa Fluor 488 (1:250, Invitrogen A11008) and goat α- mouse IgG (H&L) Alexa Fluor 594 (1:250, Invitrogen A11032), for 1 h and washed 3 times, before they were mixed with Vectashield containing DAPI (Vector laboratories, H-1200), and mounted on a microscopy slide. Slides were imaged using a Leica THUNDER 3D Assay imaging system.

Proteomics

Parasite cultures were split at 0-6 hpi in G1 and treated for 4 h with 100 nM rapamycin. Saponin lysis of paired FKBP35WT and FKBP35KO populations at 36-42 hpi in G1 and 24-30 hpi in G2 was performed as described above. Samples were collected in three independent biological replicates. The washed parasite pellet was snap frozen in liquid nitrogen. After thawing, parasites were resuspended in 25 µL of lysis buffer (5% SDS, 100 mM tetraethylammonium bromide (TEAB), 10 mM tris(2-carboxyethyl)phosphin (TCEP), pH 8.5) and sonicated (10 cycles, 30 sec on/off at 4 °C, Bioruptor, Diagnode). Lysates were subsequently reduced for 10 minutes at 95 °C. Samples were then cooled down to RT and 0.5 µL of 1M iodoacetamide was added to the samples. Cysteine residues were alkylated for 30 min at 25 °C in the dark. Digestion and peptide purification was performed using S-trap technology (Protifi) according to the manufacturer’s instructions. In brief, samples were acidified by addition of 2.5 µL of 12% phosphoric acid (1:10) and then 165 µL of S-trap buffer (90% methanol, 100 mM TEAB pH 7.1) was added to the samples (6:1). Samples were briefly vortexed and loaded onto S-trap micro spin-columns (Protifi) and centrifuged for 1 min at 4000 g. Flow-through was discarded and spin-columns were then washed 3 times with 150 µL of S-trap buffer (each time samples were centrifuged for 1 min at 4000 g and flow-through was removed). S-trap columns were then moved to the clean tubes and 20 µL of digestion buffer (50 mM TEAB pH 8.0) and trypsin (at 1:25 enzyme to protein ratio) were added to the samples. Digestion was allowed to proceed for 1h at 47 °C. Then, 40 µL of digestion buffer were added to the samples and the peptides were collected by centrifugation at 4000 g for 1 minute. To increase the recovery, S-trap columns were washed with 40 µL of 0.2% formic acid in water (400g, 1 min) and 35 µL of 0.2% formic acid in 50% acetonitrile. Eluted peptides were dried under vacuum and stored at - 20 °C until further analysis.

Peptides were resuspended in 0.1% aqueous formic acid and peptide concentration was adjusted to 0.25 µg/µL. 1 µL of each sample was subjected to LC-MS/MS analysis using an Orbitrap Elicpse Tribrid Mass Spectrometer fitted with an Ultimate 3000 nano system (both from Thermo Fisher Scientific) and a custom-made column heater set to 60 °C. Peptides were resolved using a RP-HPLC column (75 μm × 30 cm) packed in-house with C18 resin (ReproSil-Pur C18–AQ, 1.9 μm resin; Dr. Maisch GmbH) at a flow rate of 0.3 μL/min. The following gradient was used for peptide separation: from 2% buffer B to 12% B over 5 min, to 30% B over 40 min, to 50% B over 15 min, to 95% B over 2 min followed by 11 min at 95% B then back to 2% B. Buffer A was 0.1% formic acid in water and buffer B was 80% acetonitrile, 0.1% formic acid in water.

The mass spectrometer was operated in DDA mode with a cycle time of 3 seconds between master scans. Each master scan was acquired in the Orbitrap at a resolution of 240,000 FWHM (at 200 m/z) and a scan range from 375 to 1600 m/z followed by MS2 scans of the most intense precursors in the linear ion trap at “Rapid” scan rate with isolation width of the quadrupole set to 1.4 m/z. Maximum ion injection time was set to 50 ms (MS1) and 35 ms (MS2) with an AGC target set to 1e6 and 1e4, respectively. Only peptides with charge state 2-5 were included in the analysis. Monoisotopic precursor selection (MIPS) was set to Peptide, and the Intensity Threshold was set to 5e3. Peptides were fragmented by HCD (Higher-energy collisional dissociation) with collision energy set to 35%, and one microscan was acquired for each spectrum. The dynamic exclusion duration was set to 30 seconds.

The acquired raw-files were imported into the Progenesis QI software (v2.0, Nonlinear Dynamics Limited), which was used to extract peptide precursor ion intensities across all samples applying the default parameters. The generated mgf-file was searched using MASCOT against a Plasmodium falciparum (isolate 3D7) database (Uniprot, 11.2019) and 392 commonly observed contaminants using the following search criteria: full tryptic specificity was required (cleavage after lysine or arginine residues, unless followed by proline); 3 missed cleavages were allowed; carbamidomethylation (C) was set as fixed modification; oxidation (M) and acetyl (Protein N-term) were applied as variable modifications; mass tolerance of 10 ppm (precursor) and 0.6 Da (fragments). The database search results were filtered using the ion score to set the false discovery rate (FDR) to 1% on the peptide and protein level, respectively, based on the number of reverse protein sequence hits in the dataset. Quantitative analysis results from label-free quantification were processed using the SafeQuant R package v.2.3.2 (69) to obtain peptide relative abundances. This analysis included global data normalization by equalizing the total peak/reporter areas across all LC-MS runs, data imputation using the knn algorithm, summation of peak areas per protein and LC-MS/MS run, followed by calculation of peptide abundance ratios. Only isoform specific peptide ion signals were considered for quantification. To meet additional assumptions (normality and homoscedasticity) underlying the use of linear regression models and t-Tests, MS-intensity signals were transformed from the linear to the log-scale. The summarized peptide expression values were used for statistical testing of between condition differentially abundant peptides. Here, empirical Bayes moderated t-Tests were applied, as implemented in the R/Bioconductor limma package (70). The resulting per protein and condition comparison p-values were adjusted for multiple testing using the Benjamini-Hochberg method.

Gene ontology enrichment analysis was performed using the PANTHER Overrepresentation Test with the correction method “False Discovery Rate”, the test type “Fisher’s Exact”, and the annotation data set “PANTHER GO-Slim” (37).

IPP complementation

Parasite cultures were split at 0-6 hpi in G1 and treated for 4 h with 100 nM rapamycin. FKBP35KO and FKBP35WT populations were split again and supplemented either with 200 µM isopentenyl pyrophosphate (IPP) (Sigma I0503) or the corresponding volume H2O. The respective culture medium was replaced daily. Parasites treated with 156 ng/mL doxycycline (Sigma D9891) were used to confirm that IPP complementation is able to rescue apicoplast-deficient parasites. DNA content analysis was performed as described above.

Heat shock susceptibility testing

Parasites cultures were split at 0-6 hpi in G1 and treated for 4 h with 100 nM rapamycin. While control populations were always incubated at 37 °C, test populations were incubated at 40 °C for 6 h starting at 24-30 hpi. Parasitaemia was measured in G2 using flow cytometry as described above.

Surface sensing of translation (SUnSET)

Parasite cultures were split at 0-6 hpi in G1 and treated for 4 h with 100 nM rapamycin. At 24- 30 hpi in G2, the cultures were incubated with 1 µg/mL puromycin (Sigma P8833) for 1 h at 37 °C (41, 71). Saponin lysis and Western blotting were performed as described above.

Cellular thermal shift assay (CETSA)

The mass spectrometry-based isothermal dose-response cellular thermal shift assay (ITDR-MS-CETSA) was performed according to the protocol developed by Dziekan and colleagues (47). Briefly, live parasites (107 MACS-purified trophozoites per condition) were exposed to a FK506 (MedChemExpress HY-13756) concentration gradient (100 µM to 1.5 nM) and a DMSO control for 1 h. Subsequently, the parasites were washed in PBS and exposed to the denaturing temperatures 50 °C, 55 °C and 60 °C as well as to a non-denaturing temperature 37 °C for 3 min, before they were cooled down to 4 °C for 3 min. Afterwards, the cells were lysed by resuspending in a lysis buffer (50 mM HEPES pH 7.5, 5 mM β-glycerophosphate, 0.1 mM Na3VO4, 10 mM MgCl2, 2 mM TCEP, and EDTA-free protease inhibitor cocktail (Sigma)), three freeze/thaw cycles, and mechanical shearing using a syringe with a 31-gauge needle. Subsequently, insoluble proteins were pelleted by centrifugation (20 min at 4 °C and 20’000 g) and the soluble fraction was collected and subjected to protein digestion, reduction, alkylation, and TMT10 labelling followed by LC/MS analysis as described (47).

To prepare the protein lysate, saponin-lysed parasites were lysed in the afore-mentioned buffer by three freeze/thaw cycles and mechanical shearing. After centrifugation, soluble proteins were exposed to a FK506 concentration gradient (200 µM to 3 nM) and a DMSO control for 1 min before the thermal challenge was performed as described above. Afterwards, the soluble protein fraction was isolated and prepared for by LC/MS analysis as described (47).

Data analysis was performed using the Proteome Discoverer 2.1 software (Thermo Fisher Scientific) and the R package “mineCETSA” (v 1.1.1). Only proteins identified by at least three peptide spectrum matches (PSMs) were included in the analysis (47).

Transcriptomics

Sample preparation

The RNA of 500 µL iRBCs at 3-5% parasitaemia was isolated using 3 mL TRIzol reagent (Invitrogen) followed by phenol-chloroform phase separation. The aqueous phase was purified using the Direct-zol RNA MicroPrep kit (Zymo). Samples were collected from three biological replicates. Stranded RNA sequencing libraries were prepared using the Illumina TruSeq stranded mRNA library preparation kit (REF 20020594) according to the manufacturer’s protocol. The library was sequenced in 100 bp paired-end reads on an Illumina NextSeq 2000 sequencer using the Illumina NextSeq 2000 P3 reagents (REF 20040560), resulting in 5 million reads on average per sample.

RNA-Seq data analysis

Fastqc was run on all raw fastq files. Then, each file was aligned with hisat2 (version 2.0.5) (72) against the 3D7 reference genome (PlasmoDB v58) (73) in which the GFP and the recoded FBPK35 sequences were added at the end of chromosome 12. Read counts for each gene and for each sample were then counted using featureCount (gff file from PlasmoDB v58 modified to include the added GFP and FBPK35). TPM (Transcripts Per Kilobase Million) were then calculated from the raw count matrix.

Principal component analysis (PCA) was performed using the R (version 4.2.1) function prcomp using log transformed TPMs. PCA revealed that the rapamycin-treated replicate B of TP 3 (“3B.R”) was a clear outlier and is different from any other replicate at any time point (Fig. S3D). For this reason, it was excluded from further analysis. The PCA was done again without it and the other samples show the same pattern as the initial PCA (Fig. S3E).

DESeq2 (version 1.36) (74) was used for normalization and differential expression analysis. Paired differential expression tests where done for each time point using DESeq2 LRT (likelihood ratio test) using the full model ∼ Treatment + Replicate (Treatment is DMSO or Rapamycin and Replicate is the clone, added as the data is paired) and the reduced model ∼ Replicate. From this paired test and the average log fold-change calculated from the TPMs, normalized volcano plots were made using the R package EnhancedVolcano (version 1.14) (75).

Cell cycle analysis

Spearman correlation between normalized scaled read counts for each sample for all genes (as one vector) and each time point of the reference microarray time course published by Bozdech et al. (42) (3D7 smoothed, retrieved from PlasmoDB) were tested. Spearman’s coefficients rho for each time point of this RNA-Seq time course and each time point of the reference data set where plotted using ggcorrplot (version 0.1.3) (76). For each time point of this RNA-Seq dataset, the rho coefficients where then plotted against the time points of the reference data set and the FKBP35WT and FKBP35KO data were fitted separately to the linearized sinusoidal A ∗ using R lm function. The time point matching best to the FKBP35WT and FKBP35KO samples for each time point of the RNA-Seq time course was taken as the maximum of this fitted function. Both fitted functions were F-tested (using R function var.test using both fitted model as applied to the FKBP35KO data) to test whether the FKBP35WT and FKBP35KO data correlated with the reference data set (42) in a significantly different way, which would indicate a cell cycle shift if statistically significant. Only the last time point showed such a significant difference.

Data availability

The mass spectrometry proteomics data have been deposited to the ProteomeXchange Consortium repository with the dataset identifier PXD039018. The sequencing data are available in the Sequence Read Archive (SRA) via the accession number PRJNA914079.

Supporting information